### 2010-02-16 paper1/additions.tex
added draft of additions to paper

\documentclass[a4paper,12pt]{article}
\usepackage[total={6.7in,10in}, top=1.0in, left=0.9in, includefoot]{geometry}
\usepackage{color}
\usepackage{url}
\usepackage{wrapfig}
\usepackage{hyperref}

\begin{document}
\title{Katana Paper Additions}
\maketitle

\addtocounter{section}{1}

\section{Tracking Object Dependencies}
\textcolor{red}{Note that this is part is a work in progress and is
  not implemented}

\section{Automated Patching}
\subsection{Code Patching}
\subsubsection{Symbol Resolution}
\textcolor{red}{we are no longer copying symbol values into the PO
  only the names of the symbols so we can look them up on patch
  application (unless you disagree Sergey and want to go back to what
  the original paper says)}

\textcolor{red}{We \emph{are} handling relocations for code patches,
  but full support for the PLT is currently \emph{not} implemented
  (it's soon on my to-do list)}

\textcolor{red}{The addition of new symbols is not actually fully
  supported at the moment, as my first couple examples only patched
  existing variables. I could get this working in an hour or two
  though}

\subsubsection{Patch Application}
\textcolor{red}{Code patching with trampoline jumps \emph{is}
  implemented. I intend to explore relocating all calls to the
  function to avoid the overhead of the trampoline but have not
  actually made any efforts in this direction yet}

\subsection{Data Patching}
\subsubsection{Type Discovery}
\textcolor{red}{This is implemented. The only difference from the
  original paper is that I'm only insert differing types if a variable
actually using that type is encountered}

\subsubsection{Data Traversal}
\textcolor{red}{adding of the variable to the PO is implemented
  (although as mentioned earlier we do not add its address, preferring
  to rely on symbol information in the executable). Issues of scope
  and checking activation frames are not yet implemented (although
  they're next on my todo list)}
\subsection{Patch Application}
\textcolor{red}{The basics of this are implemented. Pointers are not
  yet supported. Eventually we should support custom initializers for
  new members in types}
\vspace{1in}
\textcolor{red}{the following is a new section. It should probably go
  after section 3 in the original Katana paper}
\addtocounter{section}{-1}
\subsection{Challenges}
\textcolor{red}{I would mention that there is nothing about our
  approach that inherently won't work with multithreading, just that
  dealing with it through ptrace takes considerable work. Also, I
  don't know if this is really the best place to talk about
  determining patching safety. Perhaps there should be another section
  for that?}
\section{Patch Object Format}
\subsection{Reasons and Needs}
We have developed a Patch Object format with the following properties
\begin{itemize}
 \item a valid ELF file
 \item information about types and functions requiring patches contained
   as DWARF information
 \item type transformations specified using a language defined by the
   Dwarf standard
\end{itemize}
Through the use of existing standards and well-structured ELF files
utilizing a simple expression language for data patching, we aim to
create patches which are easily examined (or modified) with existing
tools. This easy compatibility with the existing binary tools and
standards brings us to a very good point: why should patching not be a
part of the ABI and of the standard toolchain? This does not
necessarily have to be the precise format we use for Katana, any such
format that would become a standard, whether an actual standard or a
de-facto standard should be well vetted by the community, but we argue
that something like this should be included in the standard object
types, along with relocatable objects, executable objects, shared
libraries, and core dumps. Consider the situation. Relocatable objects
containing new code and data which may be inserted at runtime are
nothing new. This is the entire premise of the dynamic
library. User-written functions which may have to run upon this code
injection (in the case of patching data where the desired actions
cannot be determined automatically) already exist as the
\texttt{.init} and \texttt{.fini} sections. Due to this similarity
between some of the functionality needed by patching and the
functionality offered by dynamic
libraries, some previous systems have performed patching by creating
patches as dynamic libraries which contain not only the code and data
to be patched but the mechanism to perform the patching \cite{ginseng}
\cite{polus}. We argue that
this is an unnecessary mixing of data and logic, and that a patch
which contains merely the information necessary to fix a running
process and not the code to do so is more desirable. The code to apply
the patch should live in one place on any given system, as most other
executable content does. We do not embed emacs within our text files,
after all. Dynamic libraries and other relocatable, linkable objects
do not contain code and data intended to overwrite data in an existing
executable or process. Consider, however, that redefining certain
symbols is only a slight twist on ordinary linker behaviour. Ordinary
linker behaviour for global symbols is to fail if a symbol is defined
more than once. Ordinary behaviour for weak symbols is to use a global
definition if available or the weak symbol otherwise. When performing
dynamic linking, generally the first appropriate symbol encountered in
the chain of symbol tables is used. It is not a far difference to
define the linkage rule that the symbol definition from the most
recent patch takes precedence. Therefore, applying a patch consists of
the following steps
\begin{enumerate}
\item Injecting appropriate sections of the patch into memory. This
includes putting their contents into memory and performing relocations
on these sections (but not on the rest of the in-memory process) so
that they fit into their environment
\item Copying existing data to the appropriate regions of the newly
mapped-in patch
\item Performing relocation on the entire in-memory process such that the
symbols defined by the patch take precedence.
\end{enumerate}
These steps are all such fundamental operations that they should
become universally supported by the ABI and the toolchain.

On the other hand, note the specification of a general patch
format does not completely prescribe the patch application. From a
standard patch format, a patcher is still free to make decisions such
as when to patch safely and whether to patch functions by inserting
trampolines in the old versions of the functions or by relocating all
references to the function (we currently do the former in Katana, but
may later transition to doing the latter).

\subsection{Our Patch Object Format}
Our Patch Object (PO) format is an ELF-based
format. Figure \ref{fig:elfheaders} shows the sections contained in a
simple patch. \texttt{.text.new} and \texttt{.rodata.new} are of
course the new code and supporting
\begin{wrapfigure}{l}{0.4\textwidth}
{\footnotesize
\begin{verbatim}
  Section Headers:
    [Nr] Name              Type          
    [ 0]                   NULL          
    [ 1] .strtab           STRTAB        
    [ 2] .symtab           SYMTAB        
    [ 3] .text.new         PROGBITS      
    [ 4] .rodata.new       PROGBITS      
    [ 5] .rela.text.new    RELA          
    [ 6] .debug_info       PROGBITS      
    [ 7] .debug_abbrev     PROGBITS      
    [ 8] .debug_frame      PROGBITS      
    [ 9] .rel.debug_info   REL           
    [10] .rel.debug_frame  REL           
\end{verbatim}
}
\caption{Headers for the PO}
\label{fig:elfheaders}
\vspace{-10pt}
\end{wrapfigure}
constants to inject. \texttt{.rela.text.new} allows \texttt{.text.new}
to be properly relocated after it is adjusted. While System V based
systems use only relocation sections of type \texttt{SHT\_REL}, we
chose to use \texttt{SHT\_RELA} in our patch objects because they make
addends much easier to keep track of as we relocate from patched
binary to patch object to patched process in memory. This is all
really nothing new, storing ELF sections to be injected in-memory has
been done before in other systems. What is new in our patch object is
the inclusion of DWARF sections. The \texttt{.debug\_info} section in
an ordinary executable program contains a tree of DIEs (Debugging
Information Entities) with information about every type, variable, and
procedure in each compilation unit in the program. In a patch object,
we store information only about the procedures and variables which
have changed. This of course includes storing the type information for
changed variables. An example of the DWARF DIE information contained
in a patch can be seen in Figure \ref{fig:dies}.

Note that we store considerably less information about each entity
than is typically contained. This is because we read most of the
information from the DWARF and symbol table information of the
executing process. This allows the patch to be more flexible as it
does not require that all variables and procedures be located at
exactly the addresses they were expected to be at when the patch was
generated. This flexibility allows a single patch between versions \emph{va}
and \emph{vb} to patch both an executable which was originally compiled to \emph{va}
and an executable which was patched from earlier versions to be
equivalent to \emph{va}. Ksplice, one of the few other patchers which
operates solely at the binary level, does not have this capability \cite{ksplice}. Note
that patch versioning is currently a work in progress and not fully
implemented.
\begin{wrapfigure}{r}{0.5\linewidth}
{\footnotesize
\begin{verbatim}
  .debug_info

  COMPILE_UNIT<header overall offset = 0>:
  <0><   11>	DW_TAG_compile_unit
      DW_AT_name                  main.c

  LOCAL_SYMBOLS:
  <1><   19>	DW_TAG_subprogram
      DW_AT_name                  printThings
      DW_AT_low_pc                0x0
      DW_AT_high_pc               0x70
  <1><   40>	DW_TAG_structure_type
      DW_AT_name                  _Foo
      DW_AT_byte_size             16
      DW_AT_MIPS_fde              16
      DW_AT_sibling               <93>
  <2><   55>	DW_TAG_member
      DW_AT_name                  field1
  <2><   63>	DW_TAG_member
      DW_AT_name                  field_extra
  <2><   76>	DW_TAG_member
      DW_AT_name                  field2
  <2><   84>	DW_TAG_member
      DW_AT_name                  field3
  <1><   93>	DW_TAG_base_type
      DW_AT_name                  int
      DW_AT_byte_size             4
  <1><   99>	DW_TAG_variable
      DW_AT_name                  bar
      DW_AT_type                  <40>
\end{verbatim}
}
\caption{Dwarf DIEs in the PO}
\label{fig:dies}
\vspace{-10pt}
\end{wrapfigure}

Most of the information in the DIE tree is concerned only with names
or how to locate code within the patch object (high and low pc). Of
special interest, however, is the \texttt{fde} attribute of the
\texttt{DW\_TAG\_structure\_type}. This attribute specifies an offset
in the \texttt{.debug\_frame} section of an FDE (Frame Description
Entity). Dwarf FDEs are designed for use in transforming one call
frame into the previous call frame, and thereby walking up a call
stack for either debugging purposes or exception handling purposes
(using the \texttt{.eh\_frame} section). Transforming one call frame
to another, however, is not such a different operation from
transforming one structure to another version of the same
structure. We have aided this use with an implementation of the DWARF
virtual machine which defines several special register types
(exploiting the fact that for the purposes of generality, DWARF
registers are specified as LEB128 numbers, giving an unlimited number
of registers). The DWARF register instructions contained in the FDE
referenced in Figure \ref{fig:dies} for copying \texttt{field1},\texttt{field2}, and
\texttt{field3} from the original version of a structure \texttt{\_Foo}
to a new version of \texttt{\_Foo} which has gained the extra member
\texttt{field\_extra} in the middle of the existing fields would be
represented as in Figure \ref{fig:fdeinstrs}.
\begin{figure}[hb]
{\footnotesize
\begin{verbatim}
DW_CFA_register {CURR_TARG_NEW,0x4 bytes,0x0 off} {CURR_TARG_OLD,0x4 bytes,0x0 off} 
DW_CFA_register {CURR_TARG_NEW,0x4 bytes,0x8 off} {CURR_TARG_OLD,0x4 bytes,0x4 off} 
DW_CFA_register {CURR_TARG_NEW,0x4 bytes,0xc off} {CURR_TARG_OLD,0x4 bytes,0x8 off} 
\end{verbatim}
}
\caption{FDE instructions for data patching}
\label{fig:fdeinstrs}
\end{figure}

\texttt{CURR\_TARG\_NEW} and \texttt{CURR\_TARG\_OLD} are special
symbolic values defined by the virtual machine. If we are patching the
variable 'bar', then the \texttt{CURR\_TARG\_OLD} will be the old address of bar
(it's value in the symbol table) and \texttt{CURR\_TARG\_NEW} will be the new
address bar is being relocated to. Our registers take advantage of the
LEB128 encoding to hold a considerable amount of information in the
register identifier. In the case seen above, the first byte identifies
the class of the register (\texttt{CURR\_TARG\_NEW} or \texttt{CURR\_TARG\_OLD} in this
example), the following word specifies the size of the storage
addressed (this is included so that register assignments may copy an
arbitrary number of bytes), and the final word specifies an offset
from the address referred to by \texttt{CURR\_TARG\_}(\texttt{NEW}|\texttt{OLD}).

\textcolor{red}{the following is an addition to the related work
  section}
\section{Related Work}
   There are several hot-patching systems preceding Katana. One of the
   most well known is probably Ginseng \cite{ginseng}. Ginseng (and
   systems drawing inspiration from it such as Polus \cite{polus}) have
   successfully demonstrated patching of such important software as
   apache and sshd. These systems perform analysis of the differences
   between the original and patched versions at the source code
   level. This introduces considerable (and we argue unnecessary)
   complexity and inability to deal well with some optimizations such
   as inlining and hand-written assembly. The complexity of analyzing
   the source code ties these systems to generally a single language
   (C in the case of both Ginseng and Polus). By contrast, Katana is
   language agnostic as it works at the level of the binary ABI, and
   although we have not yet demonstrated it doing so, it should be
   able to patch binaries compiled from any language, providing the
   necessary symbol and relocation information is supplied. Ginseng
   also requires significant programmer interaction in annotating the
   code \cite{ginseng-manual} and requires
   compiling the code to use type-wrappers, allowing the patching of
   data types but at the cost of indirect access to them. The more
   programmer effort involved in generating a patch, the more likely
   the patch is to be incomplete or incorrect.

   Motivated by many of the points in the above paragraph, the
   successful Ksplice system \cite{ksplice} patches at the binary level, as we do. We
   claim the following differences from and improvements over Ksplice.
   \begin{itemize}
   \item Ksplice operates on the kernel. As their paper states, most of their
     technique is not specific to the kernel, but there is no evidence
     that it has been implemented to function on userland programs.
     Katana operates on userland.
   \item Ksplice makes no attempt to patch data, relying entirely on
     programmer-written transformation functions when data types do
     change
   \item Ksplice patches are created as kernel modules. Ksplice does not
     provide a mechanism to perform operations, such as composition,
     on these patches.
   \end{itemize}

   To the best of our knowledge, Katana is the first system to utilize
   DWARF type information in patching. 
\vspace{1in}

\textcolor{red}{References need to be merged with the original paper,
  at least the Ksplice one is already cited by the original paper}
\bibliographystyle{acm}
\bibliography{additions}
\end{document}
% LocalWords:  toolchain linkable init fini FDE FDEs DIEs Polus CURR TARG LEB
% LocalWords:  CFA DW

### 2010-02-16 paper1/bibtex.sh

@@ -1,0 +1,4 @@
+pdflatex additions
+bibtex additions
+pdflatex additions
+pdflatex additions

### 2010-02-25 paper1/additions.tex
made a few last notes on paper additions

@@
 \textcolor{red}{We \emph{are} handling relocations for code patches,
-  but full support for the PLT is currently \emph{not} implemented
-  (it's soon on my to-do list)}
-
-\textcolor{red}{The addition of new symbols is not actually fully
-  supported at the moment, as my first couple examples only patched
-  existing variables. I could get this working in an hour or two
-  though}
+  including support for the PLT}
 
@@
 addends much easier to keep track of as we relocate from patched
-binary to patch object to patched process in memory. This is all
+binary to patch objectp to patched process in memory. This is all
 really nothing new, storing ELF sections to be injected in-memory has

### 2010-02-25 katana/src/katana-prelim.c
moved a legacy function to katana prelim

@@ -7,6 +7,155 @@
                be included in the final version
 */
 
+//takes the structure stored in datap
+//of type indicated by transform->to
+//and transforms it to transform->from
+//and stores the result back in datap
+//the original contents of datap may be freed
+void fixupStructure(byte** datap,TypeTransform* transform,TransformationInfo* trans)
+{
+  byte* data=*datap;
+  TypeInfo* oldType=transform->from;
+  TypeInfo* newType=transform->to;
+  byte* newData=zmalloc(newType->length);
+  //now fix up the internals of the variable
+  int oldOffsetSoFar=0;
+  for(int i=0;i<oldType->numFields;i++)
+  {
+    if(FIELD_DELETED==transform->fieldOffsets[i])
+    {
+      continue;
+    }
+    printf("copying data from offset %i to offset %i\n",oldOffsetSoFar,transform->fieldOffsets[i]);
+    int idxOfFieldInNew=getIdxForField(newType,oldType->fields[i]);
+    TypeTransform* transformField=(TypeTransform*)dictGet(trans->typeTransformers,oldType->fieldTypes[i]->name);
+    if(!transformField)
+    {
+      int cpLength=0;
+      if(FIELD_DELETED!=idxOfFieldInNew)
+      {
+        cpLength=min(newType->fieldLengths[idxOfFieldInNew],oldType->fieldLengths[i]);
+        if(newType->fieldLengths[idxOfFieldInNew]!=oldType->fieldLengths[i])
+        {
+          fprintf(stderr,"Warning: information loss: struct field changed size between versions\n");
+        }
+      }
+      else
+      {
+        fprintf(stderr,"Warning: information loss: struct field was removed in patch");
+      }
+      memcpy(newData+transform->fieldOffsets[i],data+oldOffsetSoFar,cpLength);
+    }
+    else
+    {
+      //we can't just copy the data verbatim, its structure has changed
+      byte* fieldData=zmalloc(oldType->fieldLengths[i]);
+      memcpy(fieldData,data+oldOffsetSoFar,oldType->fieldLengths[i]);
+      fixupStructure(&fieldData,transformField,trans);
+      memcpy(newData+transform->fieldOffsets[i],fieldData,newType->fieldLengths[idxOfFieldInNew]);
+      
+    }
+    oldOffsetSoFar+=oldType->fieldLengths[i];
+  }
+  free(*datap);
+  *datap=newData;
+}
+
+//!!!legacy function. Do not use
+//returns true if it was able to transform the variable
[…]

### 2010-02-25 src/patcher/hotpatch.c
moved a legacy function to katana prelim

@@ -47,157 +47,6 @@ addr_t getFreeSpaceForTransformation(TransformationInfo* trans,uint howMuch)
   //we just discard it. This is wasteful
 }
 
-
-//takes the structure stored in datap
-//of type indicated by transform->to
-//and transforms it to transform->from
-//and stores the result back in datap
-//the original contents of datap may be freed
-void fixupStructure(byte** datap,TypeTransform* transform,TransformationInfo* trans)
-{
-  byte* data=*datap;
-  TypeInfo* oldType=transform->from;
-  TypeInfo* newType=transform->to;
-  byte* newData=zmalloc(newType->length);
-  //now fix up the internals of the variable
-  int oldOffsetSoFar=0;
-  for(int i=0;i<oldType->numFields;i++)
-  {
-    if(FIELD_DELETED==transform->fieldOffsets[i])
-    {
-      continue;
-    }
-    printf("copying data from offset %i to offset %i\n",oldOffsetSoFar,transform->fieldOffsets[i]);
-    int idxOfFieldInNew=getIdxForField(newType,oldType->fields[i]);
-    TypeTransform* transformField=(TypeTransform*)dictGet(trans->typeTransformers,oldType->fieldTypes[i]->name);
-    if(!transformField)
-    {
-      int cpLength=0;
-      if(FIELD_DELETED!=idxOfFieldInNew)
-      {
-        cpLength=min(newType->fieldLengths[idxOfFieldInNew],oldType->fieldLengths[i]);
-        if(newType->fieldLengths[idxOfFieldInNew]!=oldType->fieldLengths[i])
-        {
-          fprintf(stderr,"Warning: information loss: struct field changed size between versions\n");
-        }
-      }
-      else
-      {
-        fprintf(stderr,"Warning: information loss: struct field was removed in patch");
-      }
-      memcpy(newData+transform->fieldOffsets[i],data+oldOffsetSoFar,cpLength);
-    }
-    else
-    {
-      //we can't just copy the data verbatim, its structure has changed
-      byte* fieldData=zmalloc(oldType->fieldLengths[i]);
-      memcpy(fieldData,data+oldOffsetSoFar,oldType->fieldLengths[i]);
-      fixupStructure(&fieldData,transformField,trans);
-      memcpy(newData+transform->fieldOffsets[i],fieldData,newType->fieldLengths[idxOfFieldInNew]);
-      
-    }
-    oldOffsetSoFar+=oldType->fieldLengths[i];
-  }
-  free(*datap);
-  *datap=newData;
-}
-
[…]

### 2010-02-25 src/patcher/hotpatch.h

@@ -9,8 +9,6 @@
 #ifndef hotpatch_h
 #define hotpatch_h
 #include "types.h"
-//!!! legacy deprecated function, don't use it
-bool fixupVariable(VarInfo var,TransformationInfo* trans,int pid,ElfInfo* e);
 addr_t getFreeSpaceForTransformation(TransformationInfo* trans,uint howMuch);
 void performRelocations(ElfInfo* e,VarInfo* var);
 #endif

### 2010-02-25 src/patcher/target.c
removed legacy mmap code in target.c that used the stack, had to move away from it because of protections on executable stacks

@@ -230,19 +230,7 @@ long mmapTarget(int size,int prot)
   long code4Bytes;
   assert(sizeof(code4Bytes)==4);
   memcpy(&code4Bytes,code,4);
-  //printf("inserting code at eip 0x%x\n",newRegs.eip);
-  #ifdef OLD_MMAP_TARGET
-  modifyTarget(newRegs.esp,code4Bytes);
-  printf("inserted syscall call\n");
-  newRegs.eip=newRegs.esp;//we put our code on the stack, so direct the pc to it
-  long returnAddr=newRegs.esp+2;//the int3 instruction
-  #else
-  byte oldText[4];
-  memcpyFromTarget(oldText,newRegs.eip,4);
-  modifyTarget(newRegs.eip,code4Bytes);
-  printf("inserted syscall call\n");
-  long returnAddr=newRegs.eip+2;//the int3 instruction
-  #endif
+
   
   //the call we want to make is
   //mmap(NULL,size,prot,MAP_PRIVATE|MAP_ANONYMOUS,-1,0)

### 2010-02-25 src/patcher/target.c
error in last commit

@@ -230,6 +230,12 @@ long mmapTarget(int size,int prot)
   long code4Bytes;
   assert(sizeof(code4Bytes)==4);
   memcpy(&code4Bytes,code,4);
+  //printf("inserting code at eip 0x%x\n",newRegs.eip);
+  byte oldText[4];
+  memcpyFromTarget(oldText,newRegs.eip,4);
+  modifyTarget(newRegs.eip,code4Bytes);
+  printf("inserted syscall call\n");
+  long returnAddr=newRegs.eip+2;//the int3 instruction
 
   
   //the call we want to make is

### 2010-02-25 src/patchwrite/patchwrite.c
removed legacy code that used a dictionary of type transformers

@@ -447,10 +447,6 @@ List* getTypeTransformationInfoForCU(CompilationUnit* cuOld,CompilationUnit* cuN
 {
   List* varTransHead=NULL;
   List* varTransTail=NULL;
-  #ifdef type_transformers_dict
-  TransformationInfo* trans=zmalloc(sizeof(TransformationInfo));
-  trans->typeTransformers=dictCreate(100);//todo: get rid of magic # 100
-  #endif
   printf("Examining compilation unit %s\n",cuOld->name);
   VarInfo** vars1=(VarInfo**)dictValues(cuOld->tv->globalVars);
 
@@ -468,17 +464,11 @@ List* getTypeTransformationInfoForCU(CompilationUnit* cuOld,CompilationUnit* cuN
     TypeInfo* ti1=var->type;
     TypeInfo* ti2=patchedVar->type;
     bool needsTransform=false;
-    #ifdef type_transformers_dict
-    if(dictExists(trans->typeTransformers,ti1->name))
-    {
-      needsTransform=true;
-    }
-    #else
+
     if(ti1->transformer)
     {
       needsTransform=true;
     }
-    #endif
     else
     {
       //todo: should have some sort of caching for types we've already
@@ -494,9 +484,6 @@ List* getTypeTransformationInfoForCU(CompilationUnit* cuOld,CompilationUnit* cuN
           abort();
         }
         printf("generated type transformation for type %s\n",ti1->name);
-        #ifdef type_transformers_dict
-        dictInsert(trans->typeTransformers,ti1->name,transform);
-        #endif
         needsTransform=true;
       }
     }
@@ -505,11 +492,7 @@ List* getTypeTransformationInfoForCU(CompilationUnit* cuOld,CompilationUnit* cuN
       List* li=zmalloc(sizeof(List));
       VarTransformation* vt=zmalloc(sizeof(VarTransformation));
       vt->var=patchedVar;
-      #ifdef type_transformers_dict
-      vt->transform=(TypeTransform*)dictGet(trans->typeTransformers,var->type->name);
-      #else
       vt->transform=var->type->transformer;
-      #endif
       vt->cu=cuNew;
       if(!vt->transform)
       {

### 2010-03-02 tests/t5_pointers/v0/main.c
added test5 (not fully working yet)

@@ -1,0 +1,91 @@
+/*
+  File: main_v0.c
+  Author: James Oakley
+  Project: Katana - Preliminary Work
+  Date: January 10
+  Description: Very simple program that exists to have one of its data
+               types patched. main_v1.c is the same thing with an extra
+               field in one of the types
+*/
+#include <stdio.h>
+#include <unistd.h>
+#include <sys/mman.h>
+#include <signal.h>
+#include <string.h>
+#include <stdlib.h>
+
+typedef struct _Foo
+{
+  int* field1;
+  int* field2;
+  int* field3;
+  struct _Foo* other;
+} Foo;
+
+Foo alpha;
+Foo beta;
+Foo* gamma;
+int vals1[3]={1,2,3};
+int vals2[3]={33,22,11};
+int vals3[3]={44,55,66};
+
+
+#define MILLISECOND 1000
+
+void printFoo(char* name,Foo* foo)
+{
+  printf("name is %s\n",name);
+  fflush(stdout);
+  printf("%s lives at 0x%x\n",name,(unsigned int)foo);
+  printf("%s: %i, %i, %i\n",name,*(foo->field1),*(foo->field2),*(foo->field3));
+  if(foo->other)
+  {
+    char buf[256];
+    snprintf(buf,256,"%s.other",name);
+    printFoo(buf,foo->other);
+  }
+  else
+  {
+    printf("%s has no other member\n",name);
+  }
+}
+
+void printThings()
+{
+  printf("printThings old version starting\n");
+  char* name="alpha";
+  printf("about to call printFoo with name %s\n",name);
+  printFoo(name,&alpha);
+  printFoo("beta",&beta);
+  printFoo("gamma",gamma);
+  fflush(stdout);
+}
+
+void assignValues()
+{
+  alpha.field1=&vals1[0];
+  alpha.field2=&vals1[1];
+  alpha.field3=&vals1[2];
+  beta.field1=&vals2[0];
+  beta.field2=&vals2[1];
+  beta.field3=&vals2[2];
+  alpha.other=&beta;
+  beta.other=NULL;
+  gamma=malloc(sizeof(Foo));
+  gamma->field1=&vals3[0];
+  gamma->field2=&vals3[1];
+  gamma->field3=&vals3[2];
+  gamma->other=&alpha;
+}
+
+int main(int argc,char** argv)
+{
+  printf("has pid %i\n",getpid());
+  assignValues();
+  while(1)
+  {
[…]

### 2010-03-02 tests/t5_pointers/validate.py

@@ -1,0 +1,112 @@
+import sys,re
+def validate(logname):
+  f=open(logname)
+  if not f:
+    sys.stderr.write("Cannot open log file for test3\n");
+    return False
+
+
+  regexAlphaLives="alpha lives at 0x([a-f0-9]*)"
+  regexAlphaVal="alpha: 1, 2, 3"
+  regexAlphaOtherLives="alpha\.other lives at (0x[a-f0-9]*)"
+  regexAlphaOtherVal="alpha\.other: 33, 22, 11"
+  regexAlphaOtherOther="alpha\.other has no other member"
+  regexBetaLives="beta lives at 0x([a-f0-9]*)"
+  regexBetaVal="beta: 33, 22, 11"
+  regexBetaOther="beta has no other member"
+  regexGammaLives="gamma lives at 0x([a-f0-9]*)"
+  regexGammaVal="gamma: 44, 55, 66"
+  regexGammaOtherLives="gamma\.other lives at 0x([a-f0-9]*)"
+  regexGammaOtherVal="gamma\.other: 1, 2, 3"
+  regexGammaOtherOtherLives="gamma\.other\.other lives at 0x([a-f0-9])"
+  regexGammaOtherOtherVal="gamma\.other\.other: 33, 22, 11"
+  regexGammaOtherOtherOther="gamma\.other\.other has no other member"
+
+  
+  linecount=0
+  patchingHappened=[False,False,False] #one each for alpha, beta and gamma
+  addrs=[None,None,None] #one entry each for alpha, beta, and gamma
+
+  def validateAddr(var,match):
+    if not addrs[var]:
+      addrs[var]=match.group(1)
+      return True
+    if addrs[var]!=match.group(1):
+      if not patchingHappened[var]:
+        patchingHappened[var]=True
+        addrs[var]=match.group(1)
+      else:
+        sys.stderr.write("Address for var "+var+" changed more than once\n");
+        return False
+    
+
+  for line in f:
+    validateFunc=None
+    linecount+=1
+    patern="(no pattern chosen)"
+    def patternError():
+      sys.stderr.write("Line "+str(linecount)+" did not match  pattern\n")
+      sys.stderr.write("Offending line is:\n")
+      sys.stderr.write(line)
+      sys.stderr.write("and the pattern (regex) is:\n")
+      sys.stderr.write(pattern)
+      sys.stderr.write("\n")
+      if patchingHappened:
+        sys.stderr.write("Note, patching has been done\n")
[…]

### 2010-03-02 katana/tests/t5_pointers/v1/main.c
added test5 (not fully working yet)

@@ -1,0 +1,96 @@
+/*
+  File: main_v0.c
+  Author: James Oakley
+  Project: Katana - Preliminary Work
+  Date: January 10
+  Description: Very simple program that exists to have one of its data
+               types patched. main_v1.c is the same thing with an extra
+               field in one of the types
+*/
+#include <stdio.h>
+#include <unistd.h>
+#include <sys/mman.h>
+#include <signal.h>
+#include <string.h>
+#include <stdlib.h>
+
+typedef struct _Foo
+{
+  int* field1;
+  int field_extra;
+  int* field2;
+  int* field3;
+  struct _Foo* other;
+} Foo;
+
+Foo alpha;
+Foo beta;
+Foo* gamma;
+int vals1[3]={1,2,3};
+int vals2[3]={33,22,11};
+int vals3[3]={44,55,66};
+
+
+#define MILLISECOND 1000
+
+void printFoo(char* name,Foo* foo)
+{
+  printf("name is %s\n",name);
+  fflush(stdout);
+  printf("%s lives at 0x%x\n",name,(unsigned int)foo);
+  printf("%s: %i, %i, %i\n",name,*(foo->field1),*(foo->field2),*(foo->field3));
+  if(foo->other)
+  {
+    char buf[256];
+    snprintf(buf,256,"%s.other",name);
+    printFoo(buf,foo->other);
+  }
+  else
+  {
+    printf("%s has no other member\n",name);
+  }
+}
+
+void printThings()
+{
+  printf("printThings new version starting\n");
+  char* name="alpha";
+  printf("about to call printFoo with name %s located at 0x%x\n",name,(unsigned int)name);
+  fflush(stdout);
+  printFoo(name,&alpha);
+  name="beta";
+  printf("about to call printFoo with name %s\n",name);
+  fflush(stdout);
+  printFoo("beta",&beta);
+  printFoo("gamma",gamma);
+  fflush(stdout);
+}
+
+void assignValues()
+{
+  alpha.field1=&vals1[0];
+  alpha.field2=&vals1[1];
+  alpha.field3=&vals1[2];
+  beta.field1=&vals2[0];
+  beta.field2=&vals2[1];
+  beta.field3=&vals2[2];
+  alpha.other=&beta;
+  beta.other=NULL;
+  gamma=malloc(sizeof(Foo));
+  gamma->field1=&vals3[0];
+  gamma->field2=&vals3[1];
[…]

### 2010-03-02 run_tests.py
renamed tests

@@ -37,7 +37,7 @@ if len(sys.argv)>1:
 else:
   print "Running all tests"
   for dir in sorted(os.listdir("tests")):
-    if os.path.isdir("tests/"+dir) and dir.startswith("test"):
+    if os.path.isdir("tests/"+dir) and dir.startswith("t"):
       if True==runTestInDir("tests/"+dir):
         passTests+=1
       totalTests+=1

### 2010-03-02 references/cheating_the_elf/more_buggery.c
added reference on finding the linkmap and finding symbols at runtime

@@ -1,0 +1,265 @@
+//from http://seclists.org/bugtraq/2002/May/249
+
+/*
+ * Copyright (C) 2002 the grugq.
+ * All rights reserved.
+ *
+ * For educational purposes only.
+ *
+ * THIS SOFTWARE IS PROVIDED BY THE AUTHOR ``AS IS'' AND ANY
+ * EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
+ * IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE
+ * ARE DISCLAIMED.  IN NO EVENT SHALL THE AUTHOR BE LIABLE
+ * FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL
+ * DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS
+ * OR SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION)
+ * HOWEVER CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT
+ * LIABILITY, OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY
+ * OUT OF THE USE OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF
+ * SUCH DAMAGE.
+ */
+
+#include <elf.h>
+#define NULL    ((void *)0)
+
+#define PAGE_SIZE       4096
+// below is the real align(), but since we care about size, and we end up
+// walking down and will likely discard the rounded up value
+// we might as well use the cheat
+// #define      ALIGN(k, v)     (((k) + ((v) - 1)) & ~((v) - 1))
+
+#define PAGE_ALIGN(k)   ((k) & ~(PAGE_SIZE - 1)) // the cheat page_align()
+
+/* a POSIX structure */
+struct link_map {
+        Elf32_Addr        l_addr;
+        char            * l_name;
+        Elf32_Dyn       * l_ld;         /* .dynamic ptr */
+        struct link_map * l_next,
+                        * l_prev;
+};
+
+struct resolv {
+        Elf32_Word      * hash;
+        Elf32_Word      * chain;
+        Elf32_Sym       * symtab;
[…]

### 2010-03-02 paper1/hotpatch/abstract.tex
added paper to git repo (try 2)

\begin{abstract}
Despite advances in software modularity, security, and reliability,
offline patching remains the predominant form of updating or
protecting commodity software.  Unfortunately, the mechanics of {\it
  hot patching} (the process of upgrading a program while it executes)
remain understudied, even though such a capability offers practical
benefits for both consumer and mission-critical systems.
%
A reliable hot patching procedure would serve particularly well by
reducing the downtime necessary for critical functionality or security
upgrades.  Yet, hot patching also carries the risk -- real or
perceived -- of leaving the system in an inconsistent state, which
leads many owners to forego its benefits as too risky.  
%
%MEL:
%entity -- an
%application binary, library or even the kernel, while it is still in
%execution. 
%SB:
% It offers enormous benefits as it helps in mitigating
% process downtime and contributes to overall system reliability and
% availability.
%
%It offers great practical benefits for mission-critical systems, by
%reducing downtime for upgrades essential for maintaining their {\em
%  reliability} and {\em availability}. 
%
In this paper, we propose a novel method for hot patching ELF binaries
that supports {\em (a)} synchronized global data and code updates and
{\em (b)} reasoning about the results of applying the hot patch.  We
propose a format, which we call a {\it Patch Object}, for encoding patches as
a special type of ELF relocatable object file.  Our tool, {\em
  Katana}, automatically creates these patch objects as a by-product of the
standard source build process.  Katana also allows an end-user to
apply the Patch Objects to a running process.  In essence, our method
can be viewed as an extension of the Application Binary Interface
({\it ABI}), and we argue for its inclusion in future ABI standards.

%MEL: tried to condense to above
%We represent our hot patches as ELF-based Patch Objects, and we rely on the
%standard ELF structures in the running process' image for their
%application. 
%These Patch Objects are a special type of ELF relocatable object files,
%and are created automatically as a by-product of the standard source 
%build process by the developer-side component of our tool {\em Katana}, 
%and can then be applied by Katana's end-user binary-only component.  
%

%
%SB: This is ``how'', and is best relocated to the Intro.
% by first recognizing the importance of the stereotypical
% build process in constructing a binary through intermediate object
% files, and then delegating the overhead of maintaining object file
% dependencies to this build process. 
% In effect, what we are saying is
% that, after making the necessary source-level changes for a patch or
% an upgrade, we observe the changes to all object files and only
% consider those that have been modified or created anew by the build
% process. Correlating these with the original object files that were
% used to build the previous version of the executable, we come up with
% a mechanism to upgrade the running process with the new binary.
\end{abstract}

### 2010-03-02 paper1/hotpatch/acks.tex
added paper to git repo (try 2)

% superceded by \thanks in header
\section*{Acknowledgements}

{\small 
The first three authors' work was supported in part by the National
Science Foundation, under grant CNS-0524695.  The views and
conclusions do not necessarily represent those of the sponsors.
%
Locasto is supported in part by grant
2006-CS-001-000001 from the U.S. Department of Homeland Security under
the auspices of the I3P research program. The I3P is managed by
Dartmouth College. The opinions expressed in this paper should not be
taken as the view of the authors' institutions, the DHS, or the I3P.
}

### 2010-03-02 paper1/hotpatch/challenges.tex
added paper to git repo (try 2)

\subsection{Challenges of Patching}
\label{ssec:challenges}

% ``application'' is overloaded at this point -MEL
Requiring and encouraging the adoption of the latest security patches
is a matter of common wisdom and prudent policy.  It appears, however,
that this wisdom is routinely ignored in practice.  This disconnect
suggests that we should look for the reasons underlying users'
hesitancy to apply patches, as these reasons might be due to
fundamental technical challenges that are not yet recognized as such.
We believe that the current mechanics of applying patches prove to be
just such a stumbling block, and we contend that the underlying
challenges need to and can be addressed in a fundamental manner {\em
  by extending the core elements of the ABI and the executable file
  format}.

%The need for such fundamental changes will become apparent as we
%point out the problems facing the owners of software targeted by a
%patch.

Mission-critical systems seem hardest to patch.  They can ill afford
downtime, and the owner may be reluctant to patch due to the real or
perceived risk of the patch breaking essential functionality.  For
example, patching a component of a distributed system might lead to a
loss or corruption of state for the entire system.  An administrator
might also suspect that the patch is incompatible with some legacy
parts of the system.  Even so, the patch may target a latent
vulnerability in a software feature that is not now in active use, but
also cannot be easily made unreachable via configuration or module
unloading.  The administrator is forced to accept a particularly
thorny choice: inaction holds as much risk as a proactive
``responsible'' approach.  Since the risks of patching must
be weighed against those of staying unpatched, we seek to
%In this paper, however, we will not analyze such
%risk trade-offs, but rather concentrate on how we can 
{\bf shift the balance of this decision toward hot patching by making
  it not only possible, but also less risky in a broad range of
  circumstances}. We contend that this can only be done through good
engineering and making patching a part of the standard toolchain.

%Consider the following circumstances:
%\begin{itemize}
%  \item The patchee is part of a distributed system and cannot afford
%    loss of state, even though it can afford some very brief periods
%    of response delay ({\it e.g.,} its loss of state would require the
%    production distributed system to be restarted as whole).
%  \item The administrator suspects that the patch might not be
%    compatible with other parts of the system, and so non-applicable
%    within his legacy environment.
%  \item The patch targets an exploitable vulnerability in an
%    application feature that is not in active use but cannot be easily
%    made unreachable via the application's own configuration or module
%    support.
%\end{itemize}

Our key observation is that current binary patches, whether ``hot'' or
static, are almost entirely opaque and do not support any form of
reasoning about the impact of the patch (short of reverse engineering
both the patch and the targeted binary).  In particular, it is hard
for the software owner to find out whether and how a patch would
affect any particular subsystem or compatibility of the target with
other software in any other way than applying the patch on the test
system and trying it out, somehow finding a way to faithfully
replicate the conditions of the production environment.

Given these circumstances, our tool Katana and our Patch Object format
not only seek to make possible the mechanics of hot patching, but also
enable administrators to reduce the risk of applying a particular fix
by providing them with enough information to support examination of
the patch structure, reasoning\footnote{By which we mean manual,
  human-level reasoning, although applying automated reasoning methods
  is an interesting (and open) avenue of research.} about its
interaction with the rest of the system, and an understanding of the
tradeoffs involved in applying it.

%Thus the case for our suggested architecture is not only ``hot
%patching'' of running processes but also reducing the real or
%perceived risk though the ability to examine the structure of the
%patch and reason about its interaction with the software.

### 2010-03-02 paper1/hotpatch/conclusion.tex
added paper to git repo (try 2)

\section{Conclusion}
\label{sec:conclusion}

We introduce a method for hot patching: a technique we believe to be a
promising alternative to redundancy, ad hoc self--healing techniques,
``patch and pray,'' or other approaches to dynamic software updates.
Hot patching has the potential for aligning actual practices with
acknowledged ``best practices'' relating to critical security or 
functionality updates.  We hold that one major impediment to hot
patching is the opaque nature of most patches (be it
proprietary or open software), and our method of patching along with the
PO file format are first attempts at providing a basis for informed reasoning
about the structure and implications of a patch.

### 2010-03-02 paper1/hotpatch/design.tex
added paper to git repo (try 2)

\section{Katana Design: Tracking Object Dependencies}
\label{sec:design}

Katana leverages the typical Unix {\tt Makefile} build mechanism to track
file-level dependencies.  Normally, after applying a source patch file
and performing a top-level {\tt make}, only those object files whose
underlying sources have changed are rebuilt.  Katana thus tracks
object-level (.o) dependencies as follows.  We first replicate all
object files and the ELF executable from the existing source tree.  We
then apply the patch to the {\it original} source tree.  At this
point, only source files have been modified. Next, using the Linux
kernel's {\it inotify} mechanism\footnote{{\it inotify} allows the
  registration of filesystem triggers. We use inotify to avoid potentially
  expensive recursive directory comparisons.}, Katana sets up a notification
on the original source tree, so that it knows when an object file is
created or modified under the original tree.  Finally, we perform the
top-level {\tt make} under this source and record all created/modified
object files, along with the newly created executable. The use of
inotify is not inherent to our general patching mechanism, and indeed
our present prototype only compares two directory trees and examines
all object files.

\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.25]{making.pdf}
\end{center}
\caption{{\small {\it An Example Code Base.} From the top: each source
    file creates a corresponding object file; multiple object files
    are combined into intermediate compilation units (CU); and
    multiple CUs are merged to form the executable. All shaded blocks
    indicate modified files.}}
\label{fig:making}
\end{figure}

Figure~\ref{fig:making} illustrates how object files are modified by
{\tt make} as a result of source-level changes.  Katana only considers
objects that are closest to the source files and ignores all other
intermediate object files and compilation units (CU).  Hence, in
Figure~\ref{fig:making}, Katana only records the shaded circle-objects
along with the final executable.

To dynamically update the running application, Katana needs to patch
both the {\bf code} and the {\bf data} within the process.  It first
creates a {\it patch object} (PO): an ELF file with sections that
indicate the type of patch (code or data), the patch offsets and
lengths within the process address space, patch data, function and
data names, {\it etc.} 
%The following sections explain how Katana
%handles both code and data patching.

### 2010-03-02 paper1/hotpatch/discussion.tex
added paper to git repo (try 2)

\section{Discussion}

{\bf When to apply the patch.}
%
Dynamically updating a running application requires diligence and
patience. One cannot update the target application without any
knowledge of the program's execution state, by which we mean the
program stack, processor registers {\it etc}.  Even after possessing
this information, the application has to be in what we call a ``safe
state'' for Katana to apply the patch. We characterize a program state
as a {\it safe state} if the following two conditions hold:
\begin{itemize}
\item All activation frames in the program stack belong to functions
  that {\it do not} get updated during code patching. It is easy to
  verify this by comparing each function on the stack to the list of
  upgradeable functions contained within the PO.

\item All activation frames in the program stack belong to functions
  that do not access any global/static symbols identified during {\it
    Data Traversal}, and do not define any local variables of the
  modified types identified during {\it Type Discovery}. Again, since
  we maintain type and variable definitions within the PO, verifying
  this condition is easy.
\end{itemize}

Given the patch object, Katana uses Linux's {\tt ptrace} interface to
temporarily halt the execution of the target process, query the
current execution stack and determine if the application is currently
in a safe state. If so, then Katana applies the code patch followed by
the data patch. We note that it is not possible to apply the code and
data patches at different times since new code likely uses the new
data, and hence postponing data patching to when only the second
condition is satisfied is impractical and unsafe.

Let $A$ denote be the current activation frame on the program stack,
and the notation $(X:Y)$ define all frames from $X$ to $Y$ (on a time
scale, $X$ precedes $Y$) on the stack; so $(1:A)$ defines the current
stack. Now, in case we determine $(1:A)$ to be an unsafe state, we
could repeatedly keep querying the stack until the application reaches
some safe state. However, this is highly inefficient and cumbersome.

Instead, when we determine the program to be in an unsafe state, we
traverse up the stack from $A$, and for each preceding frame (say
$A'$), we determine if $(1:A')$ is a safe state. This takes into
consideration only $A'$ and all other frames preceding it. If $(1:A')$
is a safe state, then we insert a breakpoint on the return instruction
pointer ({\tt EIP}) pointed to by the successive frame: $(A' +
1)$. What this guarantees us is that when this breakpoint is hit, the
state of the program stack will be $(1:A')$. Since we just determined
this to be a safe state, we can reliably conduct the patching
procedure. If however, no such frames preceding $A$ satisfy a safe
state, then it means that the application cannot be patched
successfully in its current execution since there will always be a
function that violates our safety condition.

Finally we note that even after inserting a breakpoint, the problem of
determining {\it when} the breakpoint will be hit is essentially a
hard one, and so in such cases, Katana can provide no time-bounding
guarantees. Still, this is a cleaner and more efficient approach than
just na\"{i}vely retrying the full update procedure, which is both an
expensive and incomplete solution.
%JMHO:
In cases where the timely (or even eventual) application of a patch
seems unlikely, it would be possible to add support for programmers to
manually specify safe places for patching to occur. For programs built
around a central loop (the vast majority of long-running programs),
the likelihood of safe patching can be expedited by isolating the main
loop into a simple routine unlikely to ever need patching.
%JHMO end

%\subsection{Multithreading}

{\bf Address Space Randomization.}
%
{\em Load-time address randomization} has become a stable and popular
way of raising the bar for attackers, and so we must discuss how it 
interacts with our patching scheme. 
%
The gist of randomization schemes is invalidating various default
assumptions regarding the locations of code and data elements that
might facilitate exploitation.
%
%
%SB: condense
% ``Randomization''\footnote{We note that gist of this defense is not
%   randomness as such, but rather increasing the targets' diversity,
%   and denying the convenience of ``code once, run everywhere'' to the
%   attackers.}  is usually the task of the loader and the dynamic
% linker, which, at the time when a process' virtual address space is
% created or new code is loaded into it, modify its layout from the one
% prescribed by the linker that created the original executable's ELF
% structures. In particular, the virtual addresses of loadable segments
% in the file are shifted by a random\footnote{In reality, the choice of
%   offset is still limited by the platform's alignment requirements}
% offset.  This implies that the loader must perform {\em relocation},
% and indeed the code that can be so treated must be accompanied by
% relocation sections.
%
In particular, the virtual addresses of loadable segments
 are displaced by random\footnote{In reality, the choice of
   offset is still limited by the platform's alignment requirements.}
offsets by the loader, which {\em relocates} them (using their
accompanying relocation sections).
% The sections that can be so treated must be accompanied by
% relocation sections.

Patching %``randomly'' 
relocated code with our PO files requires
knowledge of the displacements introduced at loading-and-relocation
time. While there is no common ABI standard for saving this
information, conceptually it is no different from saving virtual
addresses of other files' loaded symbols in the {\it Global Offset Table}
(GOT). We note that the names of the constituent object files themselves 
are customarily included in the symbol tables, and that the symbol
table entry format can be easily adopted for storing virtual addresses
of the %``randomly'' 
relocated objects.

Thus, at the cost of small modifications to the OS loader and the 
dynamic linker, we can make the information on the layout of the
``randomly'' relocated executable and libraries available to
our patching process driven by our POs.\footnote{We note that saving 
this information 
%
about the post-relocation layout of the process 
does not weaken ``randomization'', because 
the latter does not assume the attacker's ability to arbitrarily read 
process memory (in which case the addresses of required symbols are
easily found by scanning it %). %
for their code or data patterns), 
but rather breaks hard-coding of these symbols' expected addresses.
}

{\bf Future work.} Katana is a work in progress.
%JMHO:
We have demonstrated code patching (including dynamically linked
functions), the use of our patch object format (discussed in detail in
Section \ref{sec:poformat}), and data patching including complex
structures and variable addition. The remainder of the system is still
under development. In the future we
%JMHO end
must address several 
important engineering issues, such as interaction of patched code with
dynamically loaded libraries (including the {\em dlopen} mechanism),
and assuring that accumulation of administered patches does not lead to
unacceptable performance degradation. It also poses a broader question
of describing and detecting software designs not amenable to its runtime
patching and steering programmers to avoid them if possible.

### 2010-03-02 paper1/hotpatch/intro.tex
added paper to git repo (try 2)

\section{Introduction}
\label{sec:intro}

It is somewhat ironic that users and organizatations hesistate to
apply patches whose stated purpose is to support availability or
reliability precisely because the process of doing so can lead to
downtime (both from the patching process itself as well as
unanticipated issues with the patch).  Periodic reboots in desktop
systems --- irrespective of the vendor --- are at best annoying.
Reboots in enterprise environments ({\it e.g.,} trading, e-commerce,
core network systems), even for a few minutes, imply large revenue
loss or an extensive backup and failover infrastructure with rolling
updates.  We question whether this {\em de facto} acceptence of
significant downtime and redundant infrastructure should not be
abandoned in favor of a reliable hot patching process.

%
% SB: This intro would fit in a larger version, but perhaps
%       not in this strapped-for-space one?
%
%As computing penetrates deeper into our modern infrastructure, devices
%(mobile or otherwise) and applications running on them are expected to
%be safer and more reliable compared to the previous generation. 
%SB: proposed variant:
%Even the requirement of periodic desktop reboots is nowadays
%perceived as annoying, whereas for the enterprise environments  
%taking down trading, popular e-commerce,
%or core network systems even for a few minutes implies huge
%revenue losses.
%
% Server
% uptimes are becoming ever more demanding, especially for systems
% hosting e-commerce sites, email services or social networking
% domains. Even personal computers are not spared the anathema of
% "necessary" reboots whenever an application or operating system
% service needs to be patched. Such frequent reboots interrupt both
% network-facing services and interactive applications, thereby
% frustrating users at both ends. More seriously, in the case of network
% switches and routers, disconnecting them from the network for the
% purposes of patching might render the underlying network unreachable
% in the absence of redundancy. Trading systems need to be active all
% through trading hours, and popular e-commerce sites suffer huge
% revenue losses when their servers are down for even a few minutes
% during the frenetic period of holiday shopping.
%\paragraph{}

% Despite all the foresight that goes into building what are potentially
% thought to be safe, reliable and scalable applications, such
% guarantees of security and availability cannot be proven or even
% sometimes convincingly relayed to the skeptical customer. The
% customer's trust in a system is seeded in a long-term relationship
% with the vendor whose track-record determines the product's ultimate
% success. We believe that all software is flawed, as all programmers
% are flawed. It is the magnitude of the flaw that separates the expert
% programmer from a mediocre one. As such, fixes or enhancements to
% released software is a foregone conclusion even as the original
% software is being deployed.
%
% SB: rephase and cut:

Software, the product of an inherently human process, remains a flawed
and incomplete artifact.  This reality leads to the uncomfortable
inevitability of future fixes, upgrades, and enhancements.  Given the
way such fixes are currently applied ({\it i.e.,} patch and reboot),
downtime is a foregone conclusion even as the software is released.

%Yet, all software is flawed, as all programmers are flawed. Thus
%the inevitability of fixes or enhancements to
%released software is a foregone conclusion even as the original
%software is being deployed.

%SB: small cuts
%\paragraph{}
%Such fixes or enhancements form what is routinely called a software
%patch or an upgrade.

%arcane wasn't quite the right word, i think -MEL
%barbaric, perhaps? Immature?
While patches themselves are a necessity, we believe that the process
of {\it applying} them remains rather crude.  First, the target
process is terminated, the new binary and corresponding libraries (if
any) are then written over the older versions, the system is restarted
if necessary, and finally the upgraded application begins execution.
Besides the appreciable loss in uptime, all context held by the
application is also lost, unless the application had saved its
state to persistent storage~\cite{crashonly,brown02rewind} and
later restored it (which
%seldom happens). 
is expensive to design for, implement, and execute).  In the case of
mission-critical services, even after a major flaw is unveiled and a
patch subsequently created, administrators likely wish to apply the
patch and upgrade the process without actually restarting the program
and losing state and time.  This requirement serves as our
motivation for {\it hot patching}.

### 2010-03-02 paper1/hotpatch/likelinking.tex
added paper to git repo (try 2)

\subsection{Patching In The Toolchain}
\label{ssec:patchingvlinking}
Hot patching should not be thought of as a bizarre operation, done
crudely and infrequently, something akin to reverse engineering. We
argue that it is one of the fundamental transformations in the life
cycle of any program, along with compilation, linking, dynamic
linking, and (in unfortunate cases) dumping core. We note that each of
these fundamental operations has its own type of ELF object devoted to
it (relocatable objects, executables, dynamic libraries, core dumps
respectively) and a corresponding tool in the toolchain for effecting
each transformation or working with its output. \emph{We contend that
  patching is very much like linking or dynamic linking}. Like those
operations, it combines (or replaces) parts of programs and must
generate, modify, and apply relocation information. The section types
defined for the ELF format contain nearly all of the information
necessary to describe a patch. Once we take the position that patching
is like linking, a patch needs to store the same type of symbol and
relocation information as does any relocatable ELF object.

What the base ELF specification lacks is a way to describe
types. Symbol information gives us only a location and a length, but
there is no way to describe the internal layout of a piece of
data. The DWARF format, already heavily used with ELF for debugging
and exception handling purposes, provides exactly what is needed here
as it provides a means to recursively describe types and variables, as
well as a set of instructions originally designed for restoring
register states and examining the call stack but rich in possible
applications.

Through the use of formats already used in the binary tool chain, we
hope to promote easy examination of patches, interoperability, and to
show that patching fits comfortably in to the rest of the
toolchain. We therefore propose that the standard software life cycle
should now be as shown in Figure \ref{fig:lifecycle}. Before hot
patching, only the Development and Runtime stages of the figure were
generally accepted.

\begin{figure*}[ht]
\begin{center}
\includegraphics{software_lifecycle.pdf}
\end{center}
\caption{{\small Revised software life cycle. Before hot patching, only
    the Development and Runtime portions existed}}
\label{fig:lifecycle}
\end{figure*}

% LocalWords:  toolchain

### 2010-03-02 paper1/hotpatch/main.tex
added paper to git repo (try 2)

%\documentclass[letterpaper,twocolumn,10pt]{article}
%\usepackage{usenix}
%\usepackage[margin=1in]{geometry}

\documentclass{IEEEtran}

\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{authblk}

\title{Katana: A Hot Patching Framework for ELF Executables}

\author[1]{Ashwin Ramaswamy}
\author[1]{Sergey Bratus}
\author[2]{Michael E.\ Locasto}
\author[1]{Sean W. Smith}
\affil[1]{Dartmouth College}
\affil[2]{George Mason University}

\begin{document}

\maketitle
\pagestyle{plain}

\input{abstract}

{\footnotesize
\begin{quote}
``Some reports, such as the case of the Conficker outbreak within 
Sheffield Hospital's operating ward, suggest that even security-
conscious environments may elect to forgo automated software patching, 
choosing to trade off vulnerability exposure for some perceived notion 
of platform stability...'' -- \url{http://mtc.sri.com/Conficker/}
\end{quote}
}

\input{intro}

\input{challenges}

\input{redundancy}

\input{design}

\input{patching}

%\input{discussion}

\input{poformat}

\input{related}

\input{conclusion}

{\small
\bibliographystyle{abbrv}
\bibliography{locasto,extras,oakley}
}

\end{document}

### 2010-03-02 paper1/hotpatch/patching.tex
added paper to git repo (try 2)

\section{Automated Patching}

In this section, we describe our data and code patching methods.  We
note that, compared to previous work, our PO data structures
allow reasoning about the scope, extent, and impact of the patch ({\it
 e.g.,} whether it affects particular subsystems within the process).

%\subsection{Code Patching}
{\bf Code Patching.} %\hskip 0.1in
This process involves several stages.
%\newline
\newline
(i) {\it Code Identification:} Katana first needs to identify the
section(s) of text that need to be modified within the running
process. To do this, we consider the list of all modified object files
from our tracking step, and identify all functions (both static and
global) within these files 
%SB:
from their symbol table. 
%JMHO: removed talk of copying all functions
Functions that differ between the original
and modified versions of an object file are copied into the PO and
marked as code.
%\newline
\newline
(ii) {\it Symbol Resolution:} After identifying all functions that
require a patch, we need to resolve outstanding symbol references
within each function. Typically, symbol resolution for an application
happens at both the linking stage (called {\it static linking}, when
the symbol is present within another object file or archive), and the
execution stage (or {\it dynamic linking}, when the symbol is present
within a shared library).  All code relocations are identified in the
ELF sections {\tt .rel.text} and {\tt .rela.text}, within the object
files and the final executable.  Each relocation entry contains, among
other information, the code offset that requires relocation, and the
outstanding symbol that provides this fix-up.

For each relocation entry, Katana copies the corresponding symbol into
the PO. The actual value of the symbol is not necessary in the PO
unless the target to be patched will be stripped, as the value of the
symbol can be retrieved from the running target while performing the
patching. This is key in allowing executables which have already been
patched to be patched again.
%Katana uses the replicated executable (from
%{\it before} the patch) to figure out the address of the symbol.  If
%the symbol was provided by another object file, then the symbol table
%of the old executable contains this final address, and we update the
%PO accordingly,
%SB:
%with this address as the patch target. 
If the symbol was
dynamic ({\it i.e.,} present in a shared library such as {\tt libc}),
then the fixup value is the address of a corresponding entry in the
procedure linkage table ({\it PLT}) of the executable.  The PLT is
essentially a jump table with entries for each symbol that needs to be
resolved at runtime by the dynamic linker.  When the process begins
execution, the dynamic linker maps the required shared libraries into
the address space of the process, and updates each PLT entry.

For dynamic symbols, Katana traverses the PLT entries of the
executable and compares the symbol name of each entry with
the symbol name that requires relocation. Once a match is found, the
symbol value can be determined. In our prototype,
Katana is unable to handle calls to previously
%SB:
unused functions present in any shared libraries\footnote{This would
  require creation of new PLT and GOT entries and either subsequent
  re-basing of the following segments of the executable, or creation
  of a new segment to allocate the extra entries. Although ELF
  rewriting systems like ERESI or Diablo show that such manipulations
  can be made practical, we chose not to complicate Katana with
  them yet.}
, although support for this is planned.

Finally, if the outstanding symbol's definition was not found within
the replicated executable (either within the symbol table or the PLT),
then it was newly added by the patch; it is marked as such and added
to the PO.
%\newline
\newline
(iii) {\it Patch Application:} Applying a code patch is simple enough,
and has been researched in other
systems~\cite{pannus,livepatch,ksplice,kaho}.  We map the new function
in memory, and insert a trampoline {\tt jmp} instruction at the
beginning of the old function within the process memory image. This
interposition allows the caller to execute our new function instead of
the previous one at the cost of an extra jump.  It is possible to
avoid the overhead (from branch mis-prediction) of the {\tt jmp}
instruction by adding code in the old function which traces up the
stack and modifies the caller's {\tt call} instruction operand to
point to the new address instead of the old one. Although this
optimization would ensure that all subsequent calls from the same
caller would execute the new patched function without stepping into
the old one, it does makes the process of rolling back a patch
non-trivial. A simpler method to avoid the overhead would be to
relocate all calls to the function to point to its new definition
(although the trampoline would still be desirable, to catch calls from
function pointers or anything of similar nature). The current
prototype uses only the trampoline method.

%\subsection{Data Patching}
{\bf Data Patching.} %\hskip 0.1in
Patching data within a running process is significantly harder than
patching application code.  The primary challenge here is to
synchronize the code and the data structures it acts on.\footnote{For
  example, consider adding a new member to a C struct definition and
  an additional clause to the logic that processes it.}

Tracking down previously allocated data is nontrivial (one of the
reasons why garbage collectors are interwoven with the language
implementation).  Even after identifying the allocated chunks of
memory, in the absence of some kind of a type specification, the {\it
  internal structure} of memory remains opaque.  We also need a method
for extracting only the modified data variables from the patch and a
means to discover the actual modifications that were performed. 
%JMHO: added the below because otherwise I don't think it's clear we
%solved them
Our system solves both of these problems.

We first note that any code that acts on patch-modified data is
already taken care of by Katana's code patching process. This is
because we rely on {\tt make} to build the object files that
correspond to all modified sources. We resolve the previously
identified problems towards patching data by leveraging
DWARF\footnote{\url{http://dwarfstd.org}} debugging information within the
application executable. This requires the object files to be compiled
with debugging support, but we do not see this as a limitation. Since
we only need DWARF information while building the PO, all debugging
symbols could be stripped from the executable during application
deployment, if so desired 
%JMHO:
(this would require storing symbol values in the PO, however).
%JMHO end
We recall the representation of types in
the DWARF format and then detail the various steps in Katana's data
patching process.

{\bf DWARF type information.} %\hskip 0.1in 
The DWARF structure is laid
out as a tree of DIEs ({\it Debugging Information Entries}) within the
executable file. Each DIE has an associated tag and a set of
attributes. The DIE that defines type information has the tag as one
of {\tt DW\_TAG\_base\_type}, {\tt DW\_TAG\_structure\_type} or {\tt
  DW\_TAG\_union\_type}. Typedefs and other type modifiers (such as
{\tt const}, {\tt volatile}, {\tt pointer} {\it etc.}) are referenced
by the DIE that defines the type. In case of structures or unions,
each member is contained as a separate DIE within the parent DIE that
identifies the struct/union. It is important to note that DWARF
annotates types of {\it all} visibilities from the program sources -
local, global and static.

Katana's data patching process contains a number of steps:

(i) {\it Type Discovery:} We set out to discover all newly created or
modified data types -- those that are primarily user-defined (such as
structures and unions in {\bf C}). Katana traverses the type information
(as identified by the above DWARF tags) from the newly created executable,
 and for each encountered type, it searches for
the corresponding type-name within the {\it replicated} executable
(from before the patch). If so found, the full types (i.e. the number,
type and position of all member variables contained within) are
compared to determine if they are identical. If not identical,
a transformation between the old and new versions is generated, along
with DWARF information identifying the type to insert into the PO as
soon as a variable is found making use of the altered type.
%the parent type identifying the struct/union is inserted into the PO.
 Else, if
the type name itself was not found within the replicated executable, then the
current type was created by the patch, and is added as such to the PO.
%\newline
\newline
(ii) {\it Data Traversal:} The next step is to traverse all variables
defined within the new application, and for each one encountered,
we first determine its 
%SB:
lexical scope.
If the scope is
local, then we ensure that the corresponding function (the one that
defines this variable) does not have an activation frame on  
the program stack while
applying the patch. Else, the variable has been defined as either
global or static. We first check if the replicated executable defines
the same variable. If not, then this variable has been created by the
patch and we need not worry about it and leave the symbol resolution upto
the compiler (as only {\it new} code can use this variable). Otherwise,
we verify whether the variable's type is one of the
modified types identified during {\it type discovery}. If it is, then
we add the variable along with its {\it original} address from the replicated
executable, its {\it new} address from the patch, and type information to the
PO. At the end of this stage, Katana would have identified all newly
created or modified variables from the patch. Dealing with variable
scope and safety is still a work in progress in the current prototype.
%\newline
\newline
(iii) {\it Patch Application: } Applying a data patch consists of
first tracking down the relevant symbols in program memory. Katana
reads in the PO, and for each data variable encountered, it checks if
the variable is a pointer or not. If it is, then the current validity of the
pointer is verified (by bounds-checking the pointer value to within
heap boundaries).  If the pointer is found to be invalid, no further
action is taken.  If the pointer is valid, then memory for the new
type(s) is allocated, the older structure is copied into the new one
taking into account the difference in structure definition, the old
memory is then freed, and the pointer is modified to point to the new
segment (in case of structures such as lists, trees, since we have the
type specification, we can repeat this process recursively for each
node on the list or tree). Else if the variable is not a pointer, then
Katana modifies all its references in the program text to the updated
memory location from the patch. In the current prototype, Katana
automatically zeros all {\it new} member variables within structures.
Eventually it will support default values and programmer-written
custom initializers.

{\bf Challenges.} %\hskip 0.1in 
Hot patching still faces a number of
challenges, including dealing with multithreaded programs and address
space randomization (which slight changes to the OS loader can help us
overcome).  
There is nothing inherent in our design which will not work with
multithreading, but dealing with it through ptrace takes considerable
work which we have not yet done.
Furthermore, dynamic updates require some knowledge of the
program's execution state so that the application is quiescent with
respect to the code and data being altered by Katana as it applies the
Patch Object. 
% We consider the program to be in a safe state if all
% activation records are free of functions contained in the PO and all
% activition records are free of functions that $(1)$ access any global
% or static symbols we identify during Katana's {\it Data Traversal} stage and
% $(2)$ do not define any local variables of modified types identified
% during our Type Discovery phase.  Katana uses the {\tt ptrace}
% interface to pause execution and query this state.
%
%JMHO: We repeat this again in the discussion section, so I commented
%it out here and added a brief note saying we discuss it more below
We have not
yet implemented a safety determination mechanism, but we discuss our
approach to it in the next section.

% LocalWords:  ptrace multithreading JMHO

### 2010-03-02 paper1/hotpatch/poformat.tex
added paper to git repo (try 2)

--- 2010-02-16 paper1/additions.tex
@@
-\documentclass[a4paper,12pt]{article}
-\usepackage[total={6.7in,10in}, top=1.0in, left=0.9in, includefoot]{geometry}
-\usepackage{color}
-\usepackage{url}
-\usepackage{wrapfig}
-\usepackage{hyperref}
-
-\begin{document}
-\title{Katana Paper Additions}
-\maketitle
-
-\addtocounter{section}{1}
-
-\section{Tracking Object Dependencies}
-\textcolor{red}{Note that this is part is a work in progress and is
-  not implemented}
-
-\section{Automated Patching}
-\subsection{Code Patching}
-\subsubsection{Symbol Resolution}
-\textcolor{red}{we are no longer copying symbol values into the PO
-  only the names of the symbols so we can look them up on patch
-  application (unless you disagree Sergey and want to go back to what
-  the original paper says)}
-
-\textcolor{red}{We \emph{are} handling relocations for code patches,
-  but full support for the PLT is currently \emph{not} implemented
-  (it's soon on my to-do list)}
-
-\textcolor{red}{The addition of new symbols is not actually fully
-  supported at the moment, as my first couple examples only patched
-  existing variables. I could get this working in an hour or two
-  though}
-
-\subsubsection{Patch Application}
-\textcolor{red}{Code patching with trampoline jumps \emph{is}
-  implemented. I intend to explore relocating all calls to the
-  function to avoid the overhead of the trampoline but have not
-  actually made any efforts in this direction yet}
-
-\subsection{Data Patching}
-\subsubsection{Type Discovery}
-\textcolor{red}{This is implemented. The only difference from the
-  original paper is that I'm only insert differing types if a variable
-actually using that type is encountered}
-
-\subsubsection{Data Traversal}
-\textcolor{red}{adding of the variable to the PO is implemented
-  (although as mentioned earlier we do not add its address, preferring
-  to rely on symbol information in the executable). Issues of scope
-  and checking activation frames are not yet implemented (although
-  they're next on my todo list)}
-\subsection{Patch Application}
-\textcolor{red}{The basics of this are implemented. Pointers are not
-  yet supported. Eventually we should support custom initializers for
-  new members in types}
-\vspace{1in}
-\textcolor{red}{the following is a new section. It should probably go
-  after section 3 in the original Katana paper}
-\addtocounter{section}{-1}
-\subsection{Challenges}
-\textcolor{red}{I would mention that there is nothing about our
-  approach that inherently won't work with multithreading, just that
-  dealing with it through ptrace takes considerable work. Also, I
-  don't know if this is really the best place to talk about
-  determining patching safety. Perhaps there should be another section
-  for that?}
 \section{Patch Object Format}
+\label{sec:poformat}
 \subsection{Reasons and Needs}
@@
 the patch should live in one place on any given system, as most other
-executable content does. We do not embed emacs within our text files,
+executable content does. We do not embed Emacs within our text files,
 after all. Dynamic libraries and other relocatable, linkable objects
@@
 \subsection{Our Patch Object Format}
-Our Patch Object (PO) format is an ELF-based
-format. Figure \ref{fig:elfheaders} shows the sections contained in a
-simple patch. \texttt{.text.new} and \texttt{.rodata.new} are of
-course the new code and supporting
-\begin{wrapfigure}{l}{0.4\textwidth}
-{\footnotesize
+Our Patch Object (PO) format is an ELF-based format. Figure
+\ref{fig:elfheaders} shows the sections contained in a simple
+patch. \texttt{.text.new} and \texttt{.rodata.new} are of course the
+new code and supporting
+
+%\begin{wrapfigure}{l}{0.4\textwidth}
+%{\footnotesize
+\begin{figure}[ht]
+\begin{center}
 \begin{verbatim}
@@
 \end{verbatim}
-}
+\end{center}
+%}
 \caption{Headers for the PO}
 \label{fig:elfheaders}
-\vspace{-10pt}
-\end{wrapfigure}
+%\vspace{-10pt}
+%\end{wrapfigure}
+\end{figure}
 constants to inject. \texttt{.rela.text.new} allows \texttt{.text.new}
@@
 really nothing new, storing ELF sections to be injected in-memory has
-been done before in other systems. What is new in our patch object is
+been done before in other systems \cite{eresi}. What is new in our patch object is
 the inclusion of DWARF sections. The \texttt{.debug\_info} section in
@@
 information from the DWARF and symbol table information of the
-executing process. This allows the patch to be more flexible as it
-does not require that all variables and procedures be located at
-exactly the addresses they were expected to be at when the patch was
-generated. This flexibility allows a single patch between versions \emph{va}
-and \emph{vb} to patch both an executable which was originally compiled to \emph{va}
-and an executable which was patched from earlier versions to be
-equivalent to \emph{va}. Ksplice, one of the few other patchers which
-operates solely at the binary level, does not have this capability \cite{ksplice}. Note
-that patch versioning is currently a work in progress and not fully
-implemented.
-\begin{wrapfigure}{r}{0.5\linewidth}
-{\footnotesize
+executing process (unless it will have been stripped, then more
+information must be stored in the patch). This allows the patch to be
+more flexible as it does not require that all variables and procedures
+be located at exactly the addresses they were expected to be at when
+the patch was generated. This flexibility allows a single patch
+between versions \emph{va} and \emph{vb} to patch both an executable
+which was originally compiled to \emph{va} and an executable which was
+patched from earlier versions to be equivalent to \emph{va}. Ksplice,
+one of the few other patchers which operates solely at the binary
+level, does not have this capability \cite{ksplice}. Note that patch
+versioning is currently a work in progress and not fully implemented.
+%\begin{wrapfigure}{r}{0.5\linewidth}
+\begin{figure}[ht]
+{\small
+\begin{center}
 \begin{verbatim}
@@
 \end{verbatim}
+\end{center}
 }
@@
 \vspace{-10pt}
-\end{wrapfigure}
+\end{figure}
+%\end{wrapfigure}
 
@@
 of registers). The DWARF register instructions contained in the FDE
-referenced in Figure \ref{fig:dies} for copying \texttt{field1},\texttt{field2}, and
+referenced in Figure \ref{fig:dies} for copying \texttt{field1}, \texttt{field2}, and
 \texttt{field3} from the original version of a structure \texttt{\_Foo}
@@
 represented as in Figure \ref{fig:fdeinstrs}.
-\begin{figure}[hb]
-{\footnotesize
+\begin{figure*}[ht]
+{\small
+\begin{center}
 \begin{verbatim}
@@
 \label{fig:fdeinstrs}
-\end{figure}
+\end{figure*}
 
@@
 
-\textcolor{red}{the following is an addition to the related work
-  section}
-\section{Related Work}
-   There are several hot-patching systems preceding Katana. One of the
-   most well known is probably Ginseng \cite{ginseng}. Ginseng (and
-   systems drawing inspiration from it such as Polus \cite{polus}) have
-   successfully demonstrated patching of such important software as
-   apache and sshd. These systems perform analysis of the differences
-   between the original and patched versions at the source code
-   level. This introduces considerable (and we argue unnecessary)
-   complexity and inability to deal well with some optimizations such
-   as inlining and hand-written assembly. The complexity of analyzing
-   the source code ties these systems to generally a single language
-   (C in the case of both Ginseng and Polus). By contrast, Katana is
-   language agnostic as it works at the level of the binary ABI, and
-   although we have not yet demonstrated it doing so, it should be
-   able to patch binaries compiled from any language, providing the
-   necessary symbol and relocation information is supplied. Ginseng
-   also requires significant programmer interaction in annotating the
-   code \cite{ginseng-manual} and requires
-   compiling the code to use type-wrappers, allowing the patching of
-   data types but at the cost of indirect access to them. The more
-   programmer effort involved in generating a patch, the more likely
-   the patch is to be incomplete or incorrect.
-
-   Motivated by many of the points in the above paragraph, the
-   successful Ksplice system \cite{ksplice} patches at the binary level, as we do. We
-   claim the following differences from and improvements over Ksplice.
-   \begin{itemize}
-   \item Ksplice operates on the kernel. As their paper states, most of their
-     technique is not specific to the kernel, but there is no evidence
-     that it has been implemented to function on userland programs.
-     Katana operates on userland.
-   \item Ksplice makes no attempt to patch data, relying entirely on
-     programmer-written transformation functions when data types do
-     change
-   \item Ksplice patches are created as kernel modules. Ksplice does not
-     provide a mechanism to perform operations, such as composition,
-     on these patches.
-   \end{itemize}
-
-   To the best of our knowledge, Katana is the first system to utilize
-   DWARF type information in patching. 
-\vspace{1in}
-
-\textcolor{red}{References need to be merged with the original paper,
-  at least the Ksplice one is already cited by the original paper}
-\bibliographystyle{acm}
-\bibliography{additions}
-\end{document}
 % LocalWords:  toolchain linkable init fini FDE FDEs DIEs Polus CURR TARG LEB
-% LocalWords:  CFA DW
+% LocalWords:  CFA DW Foo

### 2010-03-02 paper1/hotpatch/redundancy.tex
added paper to git repo (try 2)

\subsection{Why Not Just Employ Redundancy?}
\label{ssec:redundancy}

Redundant infrastructure, containing replicas of nodes and service
paths, often helps an organization bridge the service disruption
stemming from patches.  We believe, however, that redundancy isn't
always the best approach for ensuring availability during an upgrade
or security-critical patching process.  Rather than an established
best practice, we invite the reader to see redundancy as
an extreme measure that needlessly duplicates hardware, networking, and
software of the original system. We suggest that redundancy is:

%Such a mirrored system ensures that service remains
%uninterrupted whenever a {\it failover} is initiated. A {\it failback}
%scenario occurs when the redundant machine hands over control back to
%the original.

\renewcommand{\labelenumi}{\alph{enumi}.}

\begin{enumerate}
\item {\bf expensive -} especially in medium-sized enterprises where
  the cost of a single server, gateway, or switch is high enough to
  outweigh the benefits of redundancy.

\item {\bf wasteful -} Redundant systems are typically passive
  bystanders, lying in wait for an active machine to
  initiate a failover.

\item {{\bf requires complicated logic -}} Transferring application
  state (even across multiple homogenous systems) is non-trivial,
  especially when the state transfer occurs within hardware (such as
  for call trunks).

\item {{\bf specialized -}} The process of building system redundancy
  is not easily generalizable across heterogenous systems and requires
  full knowledge of the underlying protocol and application state in
  order to provide faithful failover and failback.
\end{enumerate}

### 2010-03-02 paper1/hotpatch/related.tex
added paper to git repo (try 2)

\section{Related Work}
\label{sec:related}

There are several hot-patching systems preceding Katana. One of the
most well known is probably Ginseng \cite{ginseng}. Ginseng (and
systems drawing inspiration from it such as Polus \cite{polus}) have
successfully demonstrated patching of such important software as
apache and sshd. These systems perform analysis of the differences
between the original and patched versions at the source code
level. This introduces considerable (and we argue unnecessary)
complexity and inability to deal well with some optimizations such as
inlining and hand-written assembly. The complexity of analyzing the
source code ties these systems to generally a single language (C in
the case of both Ginseng and Polus). By contrast, Katana is language
agnostic as it works at the level of the binary ABI, and although we
have not yet demonstrated it doing so, it should eventually be able to
patch binaries compiled from any language, providing the necessary
symbol and relocation information is supplied. Ginseng also requires
significant programmer interaction in annotating the code
\cite{ginseng-manual} and requires compiling the code to use
type-wrappers, allowing the patching of data types but at the cost of
indirect access to them. The more programmer effort involved in
generating a patch, the more likely the patch is to be incomplete or
incorrect.

Motivated by many of the points in the above paragraph, the
successful Ksplice system \cite{ksplice} patches at the binary level, as we do. We
claim the following differences from and improvements over Ksplice.
\begin{itemize}
\item Ksplice operates on the kernel. As their paper states, most of their
  technique is not specific to the kernel, but there is no evidence
  that it has been implemented to function on userland programs.
  Katana operates on userland.
\item Ksplice makes no attempt to patch data, relying entirely on
  programmer-written transformation functions when data types do
  change
\item Ksplice patches are created as kernel modules. Ksplice does not
  provide a mechanism to perform operations, such as composition,
  on these patches.
\end{itemize}

To the best of our knowledge, Katana is the first system to utilize
DWARF type information in patching.

Maintaining continuous availability, even in the absence of disruptive
events like patches, is both a challenging technical exercise and the
driving need for research on dependability, reliability, and fault
tolereance~\cite{deltaexec}. 
%The work most closely related to Katana focuses
%JMHO: reworded above because I don't think this is what's most
%closely related
Our work follows work focusing
%JMHO end
 on enabling a software
application to continue providing service or survive significant
events like errors, exploits, and patches.  This body of work includes
research on dynamic kernel updates, software survivability, and
software self-healing. However, other research areas also addressed the
challenge of enabling software to adapt at runtime, e.g., the area
of software evolution (e.g.,~\cite{software-evolution}).

The concept of crash-only software~\cite{crashonly} advocates
microrebooting: the procedure of retrofitting each component of a
system with the ability to crash and reboot safely as the default mode
of operation.  Despite its appeal as a design principle, such an
approach would be difficult to retrofit to legacy software.
Although restarting a particular service or application is disruptive
enough, rebooting the operating system itself multiplies this
disruption.  
The need to avoid that kind of downtime helped drive the
creation of frameworks like Loadable Kernel Modules for Linux, which
allow for extending the kernel during runtime without a reboot.
The ability to update the running kernel (as opposed to adding
or removing modules) without rebooting was achieved at least ten years
ago~\cite{cesare1998} and recently rediscovered, albiet mostly for
research, rather than commodity,
kernels~\cite{nonLKMkernpatch,baumann2007osupdate,soules2003k42}).
Even so,
dynamic updates of the kernel during runtime that don't require a
reboot are difficult to apply to a commodity OS, although several
efforts have been successful for the K42 experimental
system~\cite{soules2003k42,baumann2005k42}.

Software self-healing aims at ensuring continuous or increased
availability for systems subjected to exploited vulnerabilities,
either by automatically generating
patches~\cite{weimer2009genetic,stelios2005usenix} to gradually harden
the application or seeking to avoid a restart altogether by modifying
certain runtime aspects ({\it e.g.,} the memory
subsystem~\cite{rinard2004osdi}, properties of the execution
environment~\cite{rx2005sosp}), 
%
or selected control
paths~\cite{dira2005,locasto2007usenix}) of the system in response to
attacks.
One major risk of employing self-healing in production environments is
that the semantics of follow-on execution remain largely uncontrolled,
although recent work in automatically correcting memory
errors~\cite{exterminator2008cacm} seems to achieve fairly reliable
results.  Both automated responses and traditional patches can make it
difficult for an administrator to understand the implications of a
particular fix~\cite{rinard.ppp}.

### 2010-03-02 paper1/hotpatch/secse-main.tex
added paper to git repo (try 2)

\documentclass[10pt, conference, compsocconf]{IEEEtran}
\IEEEoverridecommandlockouts % for \thanks
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{url}
%\usepackage{hyperref} -- DISABLED: fails ``PDF express'' check

% \usepackage{authblk} -- must use IEEE's own
% \usepackage[margin=1in]{geometry}

\title{Katana: A Hot Patching Framework for ELF Executables}

\author{\IEEEauthorblockN{Sergey Bratus, James Oakley, Ashwin Ramaswamy, 
Sean W.\ Smith\thanks{The first four authors' work was supported in part by the National
Science Foundation, under grant CNS-0524695.  The views and
conclusions do not necessarily represent those of the sponsors.}}
\IEEEauthorblockA{Computer Science Dept.\\
  Dartmouth College\\
  Hanover, New Hampshire}
\and
\IEEEauthorblockN{Michael E.\ Locasto\thanks{Locasto is supported in part by grant
2006-CS-001-000001 from the U.S. Department of Homeland Security under
the auspices of the I3P research program. The I3P is managed by
Dartmouth College. The opinions expressed in this paper should not be
taken as the view of the authors' institutions, the DHS, or the I3P.}}
\IEEEauthorblockA{Computer Science Dept.\\
George Mason University\\
Arlington, Virginia}}

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}

\maketitle
%\pagestyle{plain}

\input{abstract}

{\footnotesize
\begin{quote}
``Some reports, such as the case of the Conficker outbreak within 
Sheffield Hospital's operating ward, suggest that even security-
conscious environments may elect to forgo automated software patching, 
choosing to trade off vulnerability exposure for some perceived notion 
of platform stability...'' -- \url{http://mtc.sri.com/Conficker/}
\end{quote}
}

\input{intro}

\input{challenges}

\input{likelinking}

\input{redundancy}

\input{design}

\input{patching}

\input{discussion}

\input{poformat}

\input{related}

\input{conclusion}

% Now in \thanks in author list.
%\input{acks}

{\small
\bibliographystyle{abbrv}
\bibliography{locasto,extras,oakley}
}

\end{document}

### 2010-03-02 katana/src/util/logging.c
new logging types

@@ -11,7 +11,7 @@
 #include <stdbool.h>
 #include <assert.h>
 
-E_LOG_LEVEL lvlEnabled;
+E_LOG_LEVEL lvlEnabled=ELL_INFO_V4;
 E_LOG_LEVEL sourceEnables[ELS_CNT];
 
 int logprintf(E_LOG_LEVEL lvl,E_LOG_SOURCE src,char* fmt,...)
@@ -69,7 +69,7 @@ void disableLogSource(E_LOG_SOURCE src)
 
 void loggingDefaults()
 {
-  lvlEnabled=ELL_INFO_V3;
+  lvlEnabled=ELL_INFO_V4;
   sourceEnables[ELS_MISC]=ELL_DISABLE;
   sourceEnables[ELS_CODEDIFF]=ELL_CNT;
   sourceEnables[ELS_TYPEDIFF]=ELL_CNT;
@@ -77,6 +77,7 @@ void loggingDefaults()
   sourceEnables[ELS_HOTPATCH]=ELL_INFO_V3;
   sourceEnables[ELS_SOURCETREE]=ELL_CNT;
   sourceEnables[ELS_SYMBOL]=ELL_INFO_V1;
-  sourceEnables[ELS_RELOCATION]=ELL_CNT;
+  sourceEnables[ELS_RELOCATION]=ELL_INFO_V3;
   sourceEnables[ELS_PATCHAPPLY]=ELL_CNT;
+  sourceEnables[ELS_LINKMAP]=ELL_INFO_V3;
 }

### 2010-03-02 katana/src/util/logging.h

@@ -32,6 +32,7 @@ typedef enum
   ELS_SYMBOL,
   ELS_RELOCATION,
   ELS_PATCHAPPLY,
+  ELS_LINKMAP,
   ELS_CNT
 } E_LOG_SOURCE;//describes what subsystem the message came from
 

### 2010-03-02 src/symbol.c
getSymtabIdx supports a subset of the flags supported by findSymbol

@@ -33,6 +33,20 @@ addr_t getSymAddress(ElfInfo* e,int symIdx)
   return sym.st_value;
 }
 
+char* unmangleSymbolName(char* name)
+{
+  char* symbolNameUnmangled=zmalloc(strlen(name)+1);
+  char* atSignPtr=strchr(name,'@');
+  if(atSignPtr)
+  {
+    int atSignIdx=atSignPtr-name;
+    symbolNameUnmangled=zmalloc(strlen(name)+1);
+    strcpy(symbolNameUnmangled,name);
+    symbolNameUnmangled[atSignIdx]='\0';
+  }
+  return symbolNameUnmangled;
+}
+
 //find the symbol matching the given symbol
 int findSymbol(ElfInfo* e,GElf_Sym* sym,ElfInfo* ref,int flags)
 {
@@ -52,14 +66,7 @@ int findSymbol(ElfInfo* e,GElf_Sym* sym,ElfInfo* ref,int flags)
   char* symbolNameUnmangled=symbolName;
   if(flags & ESFF_MANGLED_OK)
   {
-    char* atSignPtr=strchr(symbolName,'@');
-    if(atSignPtr)
-    {
-      int atSignIdx=atSignPtr-symbolName;
-      symbolNameUnmangled=zmalloc(strlen(symbolName)+1);
-      strcpy(symbolNameUnmangled,symbolName);
-      symbolNameUnmangled[atSignIdx]='\0';
-    }
+    symbolNameUnmangled=unmangleSymbolName(symbolName);
   }
   char* symbolNameDot=zmalloc(strlen(symbolName)+2);//for --fdata-sections and --ffunction-sections
   strcpy(symbolNameDot,".");
@@ -171,7 +178,7 @@ int findSymbol(ElfInfo* e,GElf_Sym* sym,ElfInfo* ref,int flags)
            (scnNameRef && scnNameNew && strcmp(scnNameRef,scnNameNew)))
         {
           //we might still be saved by considering data and bss to be the same section
-          if((type!= STT_SECTION) && 
+          if(type == STT_SECTION ||
              (!(flags & ESFF_BSS_MATCH_DATA_OK) ||
               !((!strncmp(scnNameRef,".data",strlen(".data")) &&
                  !strncmp(scnNameNew,".bss",strlen(".bss"))) ||
[…]

### 2010-03-02 src/symbol.h

@@ -20,6 +20,9 @@ typedef enum
 {
   ESFF_MANGLED_OK=1,//permit matching mangled names
   ESFF_NEW_DYNAMIC=2,//look in the dynamic symbol table instead for the new symbol
+  ESFF_DYNAMIC=ESFF_NEW_DYNAMIC,//will be OR'd NEW_DYNAMIC and
+                                //OLD_DYNAMIC when OLD_DYNAMIC IS
+                                //implemented
   ESFF_VERSIONED_SECTIONS_OK=4,//permit fuzzy matching names in versioned sections
   ESFF_BSS_MATCH_DATA_OK=8,
   ESFF_FUZZY_MATCHING_OK=ESFF_MANGLED_OK | ESFF_VERSIONED_SECTIONS_OK
@@ -35,7 +38,9 @@ Elf32_Sym gelfSymToNativeSym(GElf_Sym);
 //find it's index in the new ELF structure. Return -1 if it cannot be found
 int reindexSymbol(ElfInfo* old,ElfInfo* new,int oldIdx,int flags);
 
-int getSymtabIdx(ElfInfo* e,char* symbolName);
+//flags is OR'd E_SYMBOL_FIND_FLAGS
+//only ESFF_MANGLED_OK and ESFF_DYNAMIC are relevant
+int getSymtabIdx(ElfInfo* e,char* symbolName,int flags);
 
 idx_t findSymbolContainingAddress(ElfInfo* e,addr_t addr,byte type);
 #endif

### 2010-03-02 src/elfparse.c
elfparse find the .dynamic section

@@ -333,6 +333,10 @@ void findELFSections(ElfInfo* e)
     {
       e->sectionIndices[ERS_DYNSTR]=elf_ndxscn(scn);
     }
+    else if(!strcmp(".dynamic",name))
+    {
+      e->sectionIndices[ERS_DYNAMIC]=elf_ndxscn(scn);
+    }
   }
 }
 

### 2010-03-02 src/elfparse.h

@@ -33,6 +33,7 @@ typedef enum
   ERS_REL_PLT,
   ERS_DYNSYM,
   ERS_DYNSTR,
+  ERS_DYNAMIC,
   ERS_CNT,
   ERS_INVALID
 } E_RECOGNIZED_SECTION;

### 2010-03-02 src/patcher/target.h
added memcpyFromTargetNoDeath function and mallocTarget function

@@ -30,11 +30,26 @@ void memcpyToTarget(long addr,byte* data,int numBytes);
 //copies numBytes to data from addr in target
 //todo: does addr have to be aligned
 void memcpyFromTarget(byte* data,long addr,int numBytes);
+
+//like memcpyFromTarget except doesn't kill katana
+//if ptrace fails
+//returns true if it succeseds
+bool memcpyFromTargetNoDeath(byte* data,long addr,int numBytes);
+
 void getTargetRegs(struct user_regs_struct* regs);
 void setTargetRegs(struct user_regs_struct* regs);
 //allocate a region of memory in the target
 //return the address (in the target) of the region
 //or NULL if the operation failed
-long mmapTarget(int size,int prot);
+addr_t mmapTarget(int size,int prot);
 
+//must be called before any calls to mallocTarget
+void setMallocPLTAddress(addr_t addr);
+addr_t mallocTarget(word_t len);
+
+//compare a string to a string located
+//at a certain address in the target
+//return true if the strings match
+//up to strlen(str) characters
+bool strnmatchTarget(char* str,addr_t strInTarget);
 #endif

### 2010-03-02 katana/src/patcher/target.c
added memcpyFromTargetNoDeath function and mallocTarget function

@@ -24,6 +24,12 @@
 
 //todo: bad programming
 extern int pid;
+addr_t mallocPLTAddress=0;
+
+void setMallocPLTAddress(addr_t addr)
+{
+  mallocPLTAddress=addr;
+}
 
 void startPtrace()
 {
@@ -157,9 +163,10 @@ void memcpyToTarget(long addr,byte* data,int numBytes)
   }
 }
 
-
-//copies numBytes to data from addr in target
-void memcpyFromTarget(byte* data,long addr,int numBytes)
+//like memcpyFromTarget except doesn't kill katana
+//if ptrace fails
+//returns true if it succeseds
+bool memcpyFromTargetNoDeath(byte* data,long addr,int numBytes)
 {
   logprintf(ELL_INFO_V4,ELS_HOTPATCH,"memcpyFromTarget: getting %i bytes from 0x%x\n",numBytes,(uint)addr);
   //todo: force address to be aligned
@@ -168,8 +175,7 @@ void memcpyFromTarget(byte* data,long addr,int numBytes)
     uint val=ptrace(PTRACE_PEEKDATA,pid,addr+i);
     if(errno)
     {
-      perror("ptrace PEEKDATA failed\n");
-      death(NULL);
+      return false;
     }
     if(i+PTRACE_WORD_SIZE<=numBytes)
     {
@@ -182,6 +188,17 @@ void memcpyFromTarget(byte* data,long addr,int numBytes)
       memcpy(data+i,&val,numBytes-i);
     }
   }
+  return true;
+}
+
+//copies numBytes to data from addr in target
+void memcpyFromTarget(byte* data,long addr,int numBytes)
+{
+  if(!memcpyFromTargetNoDeath(data,addr,numBytes))
+  {
+    perror("ptrace PEEKDATA failed\n");
+    death(NULL);
+  }      
 }
 
 void getTargetRegs(struct user_regs_struct* regs)
@@ -201,12 +218,69 @@ void setTargetRegs(struct user_regs_struct* regs)
     death(NULL);
   }
 }
+
+//allocate a region of memory in the target using malloc
+//should be used for when creating objects to be used in the program,
+//as opposed to mmapTarget which should be used when mapping in new sections
+//inspiration for some of this code came from
+//http://www.hick.org/code/skape/papers/needle.txt
+addr_t mallocTarget(word_t len)
+{
+  struct user_regs_struct oldRegs,newRegs;
[…]

### 2010-03-02 katana/src/patcher/patchapply.c
use of mallocTarget for creating new memory in the target (experimental, not fully working) Also committed a bunch of uses of getSymtabIdx that last commit broke (oops)

@@ -20,7 +20,8 @@
 #include <assert.h>
 #include "symbol.h"
 #include "util/logging.h"
- 
+#include "linkmap.h"
+
 ElfInfo* patchedBin=NULL;
 ElfInfo* targetBin=NULL;
 addr_t patchTextAddr=0;
@@ -56,7 +57,7 @@ void transformVarData(VarInfo* var,Map* fdeMap,ElfInfo* patch)
 void relocateVar(VarInfo* var,ElfInfo* targetBin)
 {
   //record in the patched binary that we're putting the variable here
-  int symIdx=getSymtabIdx(targetBin,var->name);
+  int symIdx=getSymtabIdx(targetBin,var->name,0);
   GElf_Sym sym;
   getSymbol(targetBin,symIdx,&sym);
   sym.st_value=var->newLocation;
@@ -111,7 +112,7 @@ void applyFunctionPatch(SubprogramInfo* func,int pid,ElfInfo* targetBin,ElfInfo*
   //now look up the symbol in the old binary to discover where to insert
   //the trampoline jump
   //todo: return value checking
-  int idx=getSymtabIdx(targetBin,func->name);
+  int idx=getSymtabIdx(targetBin,func->name,0);
   addr_t oldAddr=getSymAddress(targetBin,idx);
   GElf_Sym sym;
   getSymbol(targetBin,idx,&sym);
@@ -454,6 +455,28 @@ void readAndApplyPatch(int pid,ElfInfo* targetBin_,ElfInfo* patch)
   
   writeOutPatchedBin(false);
 
+  #ifdef legacy
+  //todo: this will not work in programs which do not use malloc directly.
+  //what to do about this?
+    idx_t symIdxDynamic=getSymtabIdx(targetBin,"malloc",ESFF_DYNAMIC|ESFF_MANGLED_OK);
+  if(STN_UNDEF==symIdxDynamic)
+  {
+    death("could not locate symbol for malloc in the target program\n");
+  }
+  setMallocPLTAddress(getPLTEntryForSym(targetBin,symIdxDynamic));
+  #endif
+
+  addr_t mallocAddr=locateRuntimeSymbolInTarget(targetBin,"malloc");
+  if(mallocAddr)
+  {
+    setMallocPLTAddress(mallocAddr);
+  }
+  else
+  {
+    death("Cannot find malloc in the target program\n");
+  }
+  
+
[…]

### 2010-03-02 src/dwarfvm.c
use of mallocTarget for creating new memory in the target (experimental, not fully working) Also committed a bunch of uses of getSymtabIdx that last commit broke (oops)

@@ -248,7 +248,15 @@ List* makePatchData(PoRegRule* rule,SpecialRegsState* state,ElfInfo* patch,ElfIn
         //the problem now is, we have to know how much space to allocate
         //for this purpose we can use the address_range field
         //of the fde we're targeting
-        pointedObjectNewLocation=getFreeSpaceInTarget(patch->fdes[rule->index-1].memSize);
+
+        //todo: deal with freeing the original memory
+        //      and issues if the original var is
+        //      part of a larger block, not on its own
+        //      (this is very hard to get right because we're
+        //      lacking important information)
+        
+        //pointedObjectNewLocation=getFreeSpaceInTarget(patch->fdes[rule->index-1].memSize);
+        pointedObjectNewLocation=mallocTarget(patch->fdes[rule->index-1].memSize);
       }
 
       addr_t* value=zmalloc(sizeof(addr_t));

### 2010-03-02 src/patcher/hotpatch.c

@@ -55,7 +55,7 @@ addr_t getFreeSpaceInTarget(uint howMuch)
 
 void performRelocations(ElfInfo* e,VarInfo* var)
 {
-  int symIdx=getSymtabIdx(e,var->name);
+  int symIdx=getSymtabIdx(e,var->name,0);
   addr_t addr=getSymAddress(e,symIdx);
   GElf_Sym sym;
   getSymbol(e,symIdx,&sym);

### 2010-03-17 paper1/hotpatch/patching.tex
updated paper with more info about safety, limitations of type patching, thoughts on multithreading

@@
 the symbol name that requires relocation. Once a match is found, the
-symbol value can be determined. In our prototype,
-Katana is unable to handle calls to previously
-%SB:
-unused functions present in any shared libraries\footnote{This would
-  require creation of new PLT and GOT entries and either subsequent
-  re-basing of the following segments of the executable, or creation
-  of a new segment to allocate the extra entries. Although ELF
-  rewriting systems like ERESI or Diablo show that such manipulations
-  can be made practical, we chose not to complicate Katana with
-  them yet.}
-, although support for this is planned.
+symbol value can be determined. Our current prototype cannot add calls
+to previously unused functions in shared libraries, but support for
+this will be added using the ``ALTPLT'' technique described in \cite{elf-debugging}.
+
+%JMHO: this got too many negative comments from reviewers, removing
+%and clarifying above
+% Katana is unable to handle calls to previously
+% %SB:
+% unused functions present in any shared libraries\footnote{This would
+%   require creation of new PLT and GOT entries and either subsequent
+%   re-basing of the following segments of the executable, or creation
+%   of a new segment to allocate the extra entries. Although ELF
+%   rewriting systems like ERESI or Diablo show that such manipulations
+%   can be made practical, we chose not to complicate Katana with
+%   them yet.}
+% , although support for this is planned.
 
@@
 PO. At the end of this stage, Katana would have identified all newly
-created or modified variables from the patch. Dealing with variable
-scope and safety is still a work in progress in the current prototype.
+created or modified variables from the patch. 
 %\newline
@@
 Katana modifies all its references in the program text to the updated
-memory location from the patch. In the current prototype, Katana
-automatically zeros all {\it new} member variables within structures.
-Eventually it will support default values and programmer-written
+memory location from the patch. Katana supports default values in the
+sense that if the variable has an initializer in the new version of
+the program text, that initializer will be used for any member
+variables within structures which or array elements which did not
+previously exist in the executing target. Initializers from the new
+version are used for all \texttt{const} global variables.
+Eventually Katana it will support default values and programmer-written
 custom initializers.
@@
 {\bf Challenges.} %\hskip 0.1in 
-Hot patching still faces a number of
+Data patching, as described above, is not an easy task, even with
+DWARF type information. C, as well as many other directly compiled
+languages, does not have a strong type system, and is not designed to
+allow reflection. Structures are relatively straightforward to patch,
+because they contain a detailed specification. Unions and arrays, on
+the other hand, are generally quite opaque. If substantial changes
+have been made to one of the types in a union, Katana cannot do
+anything automatically, as there is no automatic way to determine
+which unioned type to act on. If an array changes size, Katana will
+assume that it is growing or shrinking at the end, and copy old data
+accordingly, but will issue a warning that this may not be the desired
+behavior. Pointers are of course handled, but with some limitations
+currently. Unfortunately for our purposes, memory management is not
+part of the ABI. Further, there is no standard way to determine what
+block a given pointer is part of. Our current implementation correctly
+handles only memory management with \texttt{malloc} and only pointers to the
+beginning of blocks. Improvement of this is a major area of future
+work. Multiple pointers to the same address are handled by keeping track
+of which addresses have been relocated. \texttt{void}* also poses a
+problem, as there is generally no way to determine the ``real'' type
+of the pointer. The general solution to all of these type problems is
+to ask the programmer for routines which perform the
+application-specific work. Minimizing programmer work is of foremost
+importance, because greater human interaction adds greater possibility
+for human error, thus possibly decreasing the reliability of the
+patch, but it is not avoidable in all situations.
+
+Hot patching still faces a number of other
 challenges, including dealing with multithreaded programs and address
@@
 multithreading, but dealing with it through ptrace takes considerable
-work which we have not yet done.
-Furthermore, dynamic updates require some knowledge of the
-program's execution state so that the application is quiescent with
-respect to the code and data being altered by Katana as it applies the
-Patch Object. 
+work which we have not yet done. More importantly, the deadlock
+avoidance work required to ensure that we do at some point pause all
+threads, but that we never pause a thread that another thread may be
+waiting on in order to reach a safe state is nontrivial. We plan to
+address this in the future.
 % We consider the program to be in a safe state if all
@@
 %it out here and added a brief note saying we discuss it more below
-We have not
-yet implemented a safety determination mechanism, but we discuss our
-approach to it in the next section.
 

### 2010-03-18 paper1/hotpatch/conclusion.tex
fleshed out conclusion a little bit

@@
 about the structure and implications of a patch.
+
+We present a reasoned approach to making patching a part of the
+standard tool chain. We demonstrate a working binary userland patcher
+operating completely at the object level. Our system is, to our
+knowledge, the first to utilize DWARF type information to automate the
+transformation between old and new versions of a type. There yet
+remains much work to be done, and our future work involves support for
+patching multi-threaded targets, better support for handling opaque
+types such as \texttt{void}*, and further development of patch
+versioning and the ability to perform operations on patch objects.

### 2010-03-18 paper1/hotpatch/design.tex
played down the role of inotify

@@
-\section{Katana Design: Tracking Object Dependencies}
+\section{Katana Design}
 \label{sec:design}
 
-Katana leverages the typical Unix {\tt Makefile} build mechanism to track
-file-level dependencies.  Normally, after applying a source patch file
-and performing a top-level {\tt make}, only those object files whose
-underlying sources have changed are rebuilt.  Katana thus tracks
-object-level (.o) dependencies as follows.  We first replicate all
-object files and the ELF executable from the existing source tree.  We
-then apply the patch to the {\it original} source tree.  At this
-point, only source files have been modified. Next, using the Linux
-kernel's {\it inotify} mechanism\footnote{{\it inotify} allows the
-  registration of filesystem triggers. We use inotify to avoid potentially
-  expensive recursive directory comparisons.}, Katana sets up a notification
-on the original source tree, so that it knows when an object file is
-created or modified under the original tree.  Finally, we perform the
-top-level {\tt make} under this source and record all created/modified
-object files, along with the newly created executable. The use of
-inotify is not inherent to our general patching mechanism, and indeed
-our present prototype only compares two directory trees and examines
-all object files.
+In our prototype, a patch is may be generated as follows. The source
+directory corresponding to the target to be patched is replicated and
+the source code is patched to the version desired. The modified source
+tree is then built, and compared with the original source tree at the
+object (.o) level. Those object files which have changed between the
+modified and original source trees are added to the list of objects
+which must be examined for type and code transformations. Future work
+will include use of the {\it inotify} mechanism to avoid potentially
+expensive recursive directory comparison and provide more precise
+notification of changed files.
+
+% Katana leverages the typical Unix {\tt Makefile} build mechanism to track
+% file-level dependencies.  Normally, after applying a source patch file
+% and performing a top-level {\tt make}, only those object files whose
+% underlying sources have changed are rebuilt.  Katana thus tracks
+% object-level (.o) dependencies as follows.  We first replicate all
+% object files and the ELF executable from the existing source tree.  We
+% then apply the patch to the {\it original} source tree.  At this
+% point, only source files have been modified. Next, using the Linux
+% kernel's {\it inotify} mechanism\footnote{{\it inotify} allows the
+%   registration of filesystem triggers. We use inotify to avoid potentially
+%   expensive recursive directory comparisons.}, Katana sets up a notification
+% on the original source tree, so that it knows when an object file is
+% created or modified under the original tree.  Finally, we perform the
+% top-level {\tt make} under this source and record all created/modified
+% object files, along with the newly created executable. The use of
+% inotify is not inherent to our general patching mechanism, and indeed
+% our present prototype only compares two directory trees and examines
+% all object files.
 
@@
 
-Figure~\ref{fig:making} illustrates how object files are modified by
-{\tt make} as a result of source-level changes.  Katana only considers
-objects that are closest to the source files and ignores all other
-intermediate object files and compilation units (CU).  Hence, in
-Figure~\ref{fig:making}, Katana only records the shaded circle-objects
-along with the final executable.
+%reworded to how we're actually doing it now
+% Figure~\ref{fig:making} illustrates how object files are modified by
+% {\tt make} as a result of source-level changes.  Katana only considers
+% objects that are closest to the source files and ignores all other
+% intermediate object files and compilation units (CU).  Hence, in
+% Figure~\ref{fig:making}, Katana only records the shaded circle-objects
+% along with the final executable.
 
@@
 lengths within the process address space, patch data, function and
-data names, {\it etc.} 
+data names, {\it etc.}. The patch object may then be applied to the
+target at any time.
 %The following sections explain how Katana

### 2010-03-18 paper1/hotpatch/intro.tex
changed the title and moved around the conficker quote

@@
 loss or an extensive backup and failover infrastructure with rolling
-updates.  We question whether this {\em de facto} acceptence of
+updates.  
+{\footnotesize
+\begin{quote}
+``Some reports, such as the case of the Conficker outbreak within 
+Sheffield Hospital's operating ward, suggest that even security-
+conscious environments may elect to forgo automated software patching, 
+choosing to trade off vulnerability exposure for some perceived notion 
+of platform stability...'' -- \url{http://mtc.sri.com/Conficker/}
+\end{quote}
+}
+We question whether this {\em de facto} acceptence of
 significant downtime and redundant infrastructure should not be

### 2010-03-18 paper1/hotpatch/secse-main.tex
use APA bibliographic style
back to using abbrv bib style until I figure out how it's actually supposed to be
changed the title and moved around the conficker quote

@@
 \usepackage{url}
+%\usepackage{natbib}
 %\usepackage{hyperref} -- DISABLED: fails ``PDF express'' check
@@
 
-\title{Katana: A Hot Patching Framework for ELF Executables}
+\title{Katana: Towards Patching as a Part of the ELF Toolchain}
 
@@
 
-{\footnotesize
-\begin{quote}
-``Some reports, such as the case of the Conficker outbreak within 
-Sheffield Hospital's operating ward, suggest that even security-
-conscious environments may elect to forgo automated software patching, 
-choosing to trade off vulnerability exposure for some perceived notion 
-of platform stability...'' -- \url{http://mtc.sri.com/Conficker/}
-\end{quote}
-}
-
 \input{intro}
@@
 {\small
+%\bibliographystyle{apa-good}
 \bibliographystyle{abbrv}

### 2010-03-18 paper1/hotpatch/redundancy.tex
slight expansion to redundancy argument

@@
 \end{enumerate}
+
+It should be noted that the last two items apply even for virtualized
+redundant systems, which often do not have the traditional overhead of
+redundant hardware.  We do not claim that redundancy does not have its
+place, but redundancy does not provide the easy, ubiquitous solution to
+high-availability stateful applications which we hope to provide
+through patching
+
+% LocalWords:  stateful virtualized

### 2010-03-18 paper1/hotpatch/poformat.tex
slight clarification on versioning issue

@@
     [ 3] .text.new         PROGBITS      
-    [ 4] .rodata.new       PROGBITS      
-    [ 5] .rela.text.new    RELA          
-    [ 6] .debug_info       PROGBITS      
-    [ 7] .debug_abbrev     PROGBITS      
-    [ 8] .debug_frame      PROGBITS      
-    [ 9] .rel.debug_info   REL           
-    [10] .rel.debug_frame  REL           
+    [ 4] .unsafe_functions LOUSER+1
+    [ 5] .rodata.new       PROGBITS      
+    [ 6] .rela.text.new    RELA          
+    [ 7] .debug_info       PROGBITS      
+    [ 8] .debug_abbrev     PROGBITS      
+    [ 9] .debug_frame      PROGBITS      
+    [10] .rel.debug_info   REL           
+    [11] .rel.debug_frame  REL           
 \end{verbatim}
@@
 one of the few other patchers which operates solely at the binary
-level, does not have this capability \cite{ksplice}. Note that patch
-versioning is currently a work in progress and not fully implemented.
+level, does not have this capability \cite{ksplice}. We will provide a
+mechanism for composing patches, such that a patch from version
+\emph{va} to \emph{vb} may be composed with a patch from \emph{vb} to
+\emph{vc} to produce a patch from \emph{va} to \emph{vb}. Note that
+patch versioning is currently a work in progress and not fully
+implemented.
 %\begin{wrapfigure}{r}{0.5\linewidth}

### 2010-03-18 paper1/hotpatch/conclusion.tex
changed hyphenation of multithreading to be consistent

@@
 remains much work to be done, and our future work involves support for
-patching multi-threaded targets, better support for handling opaque
+patching multithreaded targets, better support for handling opaque
 types such as \texttt{void}*, and further development of patch

### 2010-03-18 paper1/hotpatch/response_to_reviews.txt
added response to reviewers

The following document details how we handled each comment for each of
the reviewers for IJSSE paper #2493. Comments from the reviewers are
quoted using the characters >> at the start of each line.

REVIEW 1
>> I urge the authors to complete the missing work, especially in
>> the area of safety and multithreading. Both are prerequisites for
>> this work being acceptable in the real world.
We have actually made significant work on implementing safety analysis
since the submission for review, so we updated the paper slightly with
that information. We added a slightly greater nod to the importance of
multithreading (and our intent to address it) in the Challenges
subsection of section III.

>> Section I should mention virtualization as a path to upgrading
>> existing systems with patches, basically rolling upgrades. While
>> this won't work for monolithic systems, it does work extremly well
>> for horizontally scaled software architectures and eliminates the
>> need for hotpatching. If traditional patching is - as you claim -
>> often not done due to the preception of instability, why would
>> hotpatching be perceived as a better choice?
Virtualization is a type of redundancy. The hardware expense and waste
does not apply to it, but our arguments about state loss and
complexity do apply. We have clarified our argument about redundancy
slightly to show that this case is included.

>> I would redraw Fig.1 as follows: - remove second patcher - add arrow
>> from patcher back into runtime. Basically, the future/patched runtime
>> should not appear any different that the current unpatched
>> runtime. Think about it, in the next phase the previously patched
>> runtime is the basis you are starting from. - Rename "Patched Runtime"
>> to something like "Hotpatch application phase", etc.
We disagree. We are showing the lifecycle of software and how it is
extended, not that it is running differently after it is patched. The
point about the name of that stage is well taken, however, and it has
been renamed.

>> Your description includes only sequential patch application (n-1 ->
>> n). What about wanting to upgrade from n-5 to n?  
This is entirely possible, and we have added a slight clarification in
section V to make this harder to miss

REVIEW 2 
>> The quote after the abstract is not necessary. If the authors
>> strongly feel about it, they can include that in an appropriate place
>> of the main text of the paper.
The quote has been moved to a more appropriate location.

>> The extension from the workshop paper is not clear, it may be
>> mentioned briefly by referring to the workshop paper in the
>> Introduction or the related works.

>> The title of both workshop and journal papers same. Based on the new
>> addition in the journal version, the title of the paper needs to be
>> changed.
The title has been changed

>> Section VII could be expanded with the details of major contributions,
>> lessons learned, limitations and future work.
Section VII has been expanded slightly. Much of this is covered
elsewhere in the paper, so to repeat it would be to risk excessive
redundancy.

REVIEW 3
>> The quote concerning conficker at the start of the paper is an
>> interesting anecdote but should perhaps be placed where you make the
>> argument that some trade security for perceived stability. 
The quote has been moved to a more appropriate location.

>> Your argument concerning redundancy is a bit of a strawman. Redundancy
>> is primarily used to ensure high availability, and in that context the
>> cost and resource is can be motivated. The ability to upgrade systems
>> with minimal disruption is just a nice byproduct. For patching
>> purposes alone, one could operate with a minimal amount of extra
>> resources (negating the "expensive" and "wasteful" arguments). Complex
>> and specialized do still apply in many (but far from all) cases.
We disagree. We are not arguing that redundancy is useless, merely
that it is not as attractive as a ubiquitous patching
mechanism. Building a system with redundancy advanced enough that
state can be transferred from one instance of a task to another
without requiring per-application coding of the state transferral is
not known to us to be a solved problem, and the reviewer points us to
no evidence that it is.

>> Figure one is far too large. 
This has been remedied

>> First you say Katana uses inotify, then that your current prototype
>> doesn't use it. Which is it? 
The original plan was to use inotify. We have never yet implemented its
use, as it is not central to patching and other features were more
important. We have therefore modified the section mentioning it to
give it only a very brief mention as "future work".

>> Data patching is the most challenging part of the problem. You discuss
>> this, but it isn't clear exacly what the limits of your approach are.
>> Can you handle pointers to members of structures (when the structure
>> changes)? Can you handle arrays? Pointers to array elements? Changed
>> meaning of data (e.g. if an element is added to an enum)?
>> Variable-size structures (i.e. those with a variable-size buffer at
>> the end)? Will pointer casts cause problems (e.g. if a pointer to a
>> changed data object is stored in a void* or caddr_t variable)? Do you
>> handle multiple pointers to the same heap object (or the situation
>> where you have one pointer to the start of a heap object, and one to
>> an element inside it)? Do unions cause problems (e.g. union between an
>> int and a pointer)?  
These are very valid questions. Section III
has been expanded slightly in an effort to address these concerns and
the current limitations of our system.

>> The limitation to code that does not call previously unused functions
>> in dynamically loaded libraries seems quite serious. Since you have
>> chosen not to support this, presumably you do not consider it a severe
>> limitation of Katana. Can you explain why? 
This is very true. When we were designing the system, it was deemed
difficult. Given its importance, however, we have made plans to
support it. We have fixed this in the paper, and cite the technique
that will be used.

>> Katana will only update when the program is in a safe state. In real
>> programs, how often will the program never enter a safe state? What
>> practical experience do you have with this? What practical experiences
>> do you have so far?
Unfortunately, we have little. We will certainly be publishing
real-world results when we have optained them.

### 2010-03-18 paper1/hotpatch/intro.tex
added note about the power grid

@@
 \label{sec:intro}
-
 It is somewhat ironic that users and organizatations hesistate to
@@
 updates.
+
 {\footnotesize
@@
 }
+
 We question whether this {\em de facto} acceptence of
@@
 patch and upgrade the process without actually restarting the program
-and losing state and time.  This requirement serves as our
-motivation for {\it hot patching}.
+and losing state and time.  This requirement serves as our motivation
+for {\it hot patching}. We focus on systems which are both
+high-availability and store significant state (which would be lost on
+a restart) such as those found in the power grid.

### 2010-03-18 paper1/hotpatch/poformat.tex
removed numbered subsection headings, as required by style for this publication

@@
 \label{sec:poformat}
-\subsection{Reasons and Needs}
+%\subsection{Reasons and Needs}
+{\bf Reasons and Needs.}
 We have developed a Patch Object format with the following properties
@@
 
-\subsection{Our Patch Object Format}
+%\subsection{Our Patch Object Format}
+{\bf Our Patch Object Format.}
 Our Patch Object (PO) format is an ELF-based format. Figure

### 2010-03-18 src/dwarftypes.c
fixed a bug where isPrototype flag not set for subroutine types

@@ -1018,7 +1018,7 @@ void* addSubroutineType(Dwarf_Debug dbg,Dwarf_Die die,CompilationUnit* cu)
   TypeInfo* type=zmalloc(sizeof(TypeInfo));
   type->type=TT_SUBROUTINE_TYPE;
   type->cu=cu;
-
+  type->declaration=isPrototype;
   type->name=getNameForDie(dbg,die,cu);
   logprintf(ELL_INFO_V4,ELS_DWARFTYPES,"reading subroutine type of name %s\n",type->name);
   type->length=0;//subroutine type in a way isn't a real type. It certainly has no length

### 2010-03-18 katana/src/info/dwinfo_dump.c
display the FDE offset that a type is transformed by as an intermediate measure. Really need to figure out the index

@@ -48,8 +48,9 @@ void printPatchDwarfInfo(ElfInfo* patch)
     for(int i=0;vars[i];i++)
     {
       printf("    %s %s\n",vars[i]->type->name,vars[i]->name);
+      //todo: should display an index
+      printf("      transformed by FDE at offset %i\n",vars[i]->type->fde);
     }
     free(vars);
   }
-
 }

### 2010-03-18 katana/src/patchwrite/sourcetree.c
fixed bug where sourcetree analyzer did not correctly handle recursing on a subdirectory that existed only in one version but not the other

@@ -126,32 +126,56 @@ List* getChangedObjectFilesInSourceTree(char* origSourceTree,char* modSourceTree
       }
       else
       {
-        logprintf(ELL_INFO_V1,ELS_SOURCETREE,"Object file %s does not differ between versions\n",fullPathOrig);
-        free(fullPathOrig);
-        free(fullPathMod);
-        free(dirEntriesOrig[i]);
-        free(dirEntriesMod[j]);
-        i++,j++;
+          logprintf(ELL_INFO_V1,ELS_SOURCETREE,"Object file %s does not differ between versions\n",fullPathOrig);
+          free(fullPathOrig);
+          free(fullPathMod);
+          free(dirEntriesOrig[i]);
+          free(dirEntriesMod[j]);
+          i++,j++;
       }
       free(buf);
     }
     else if(cmp<0)
     {
-      logprintf(ELL_WARN,ELS_SOURCETREE,"An object file '%s' has been removed from the new version of the source\n",dirEntriesOrig[i]->d_name);
+      if(DT_DIR==dirEntriesOrig[i]->d_type)
+      {
+        //recurse on the directories, only
+        //the old version is present though
+        head=concatLists(head,tail,
+                         getChangedObjectFilesInSourceTree(fullPathOrig,NULL),
+                         NULL,&tail);
+      }
+      else
+      {
+        logprintf(ELL_WARN,ELS_SOURCETREE,"An object file '%s' has been removed from the new version of the source\n",dirEntriesOrig[i]->d_name);
+      }
       free(fullPathOrig);
       free(fullPathMod);
       free(dirEntriesOrig[i]);
       i++;
+
     }
     else if (cmp>0)
     {
-      //file only exists in new version
-      List* li=zmalloc(sizeof(List));
-      ObjFileInfo* obj=zmalloc(sizeof(ObjFileInfo));
-      obj->state=EOS_NEW;
-      obj->pathToModified=fullPathMod;
-      li->value=obj;
-      listAppend(&head,&tail,li);
+      if(DT_DIR==dirEntriesOrig[i]->d_type)
+      {
+        //recurse on the directories, only
+        //the old version is present though
+        head=concatLists(head,tail,
+                         getChangedObjectFilesInSourceTree(NULL,fullPathMod),
+                         NULL,&tail);
+      }
+      else
+      {
+        
+        //file only exists in new version
+        List* li=zmalloc(sizeof(List));
[…]

### 2010-03-18 src/patchwrite/write_to_dwarf.c
write out the DW_AT_prototyped flag when appropriate, when writing to dwarf. Allows writing subrouting types to dwarf

@@ -106,6 +106,10 @@ void writeTypeToDwarf(Dwarf_P_Debug dbg,TypeInfo* type)
   Dwarf_P_Die die=dwarf_new_die(dbg,tag,parent,NULL,type->cu->lastDie,NULL,&err);
   type->cu->lastDie=die;
   type->die=die;
+  if(type->declaration)
+  {
+    dwarf_add_AT_flag(dbg,die,DW_AT_prototyped,true,&err);
+  }
   if(TT_CONST!=type->type && TT_SUBROUTINE_TYPE!=type->type)
   {
     dwarf_add_AT_name(die,type->name,&err);

### 2010-03-18 real_tests/apache/validate.py
added some nod to the apache test

@@ -1,0 +1,4 @@
+
+def validate(logname):
+  #todo: write this function
+  return False

### 2010-03-27 paper1/hotpatch/abstract.tex
many grammar fixes on advice from a retired english teacher

@@
 perceived -- of leaving the system in an inconsistent state, which
-leads many owners to forego its benefits as too risky.  
+leads many owners to forgo its benefits as too risky.  
 %

### 2010-03-27 paper1/hotpatch/conclusion.tex
many grammar fixes on advice from a retired english teacher

@@
 patching is the opaque nature of most patches (be it
-proprietary or open software), and our method of patching along with the
-PO file format are first attempts at providing a basis for informed reasoning
+proprietary or open software), and our method of patching, along with the
+PO file format, is a first attempt at providing a basis for informed reasoning
 about the structure and implications of a patch.

### 2010-03-27 paper1/hotpatch/design.tex
many grammar fixes on advice from a retired english teacher

@@
 In our prototype, a patch is may be generated as follows. The source
-directory corresponding to the target to be patched is replicated and
+directory corresponding to the target to be patched is replicated, and
 the source code is patched to the version desired. The modified source
-tree is then built, and compared with the original source tree at the
-object (.o) level. Those object files which have changed between the
+tree is then built and compared with the original source tree at the
+object (.o) level. Those object files that have changed between the
 modified and original source trees are added to the list of objects
-which must be examined for type and code transformations. Future work
+that must be examined for type and code transformations. Future work
 will include use of the {\it inotify} mechanism to avoid potentially
@@
 lengths within the process address space, patch data, function and
-data names, {\it etc.}. The patch object may then be applied to the
+data names, {\it etc}. The patch object may then be applied to the
 target at any time.

### 2010-03-27 paper1/hotpatch/discussion.tex
updated paper with more info about safety, limitations of type patching, thoughts on multithreading
many grammar fixes on advice from a retired english teacher

@@
 knowledge of the program's execution state, by which we mean the
-program stack, processor registers {\it etc}.  Even after possessing
+program stack, processor registers, {\it etc}.  Even after possessing
 this information, the application has to be in what we call a ``safe
@@
   that {\it do not} get updated during code patching. It is easy to
-  verify this by comparing each function on the stack to the list of
+  verify this by comparing each function on the stack with the list of
   upgradeable functions contained within the PO.
@@
   that do not access any global/static symbols identified during {\it
-    Data Traversal}, and do not define any local variables of the
+    Data Traversal} and that do not define any local variables of the
   modified types identified during {\it Type Discovery}. Again, since
@@
 temporarily halt the execution of the target process, query the
-current execution stack and determine if the application is currently
+current execution stack, and determine whether the application is currently
 in a safe state. If so, then Katana applies the code patch followed by
@@
 
-Let $A$ denote be the current activation frame on the program stack,
+Let $A$ denote the current activation frame on the program stack,
 and the notation $(X:Y)$ define all frames from $X$ to $Y$ (on a time
@@
 just na\"{i}vely retrying the full update procedure, which is both an
-expensive and incomplete solution.
+expensive and an incomplete solution.
 %JMHO:
@@
 the likelihood of safe patching can be expedited by isolating the main
-loop into a simple routine unlikely to ever need patching.
+loop into a simple routine unlikely ever to need patching. While
+safety determination is implemented in our prototype, we do not yet
+have enough real data about the likelihood of applications' failing to
+reach a safe state.
 %JHMO end
@@
 (GOT). We note that the names of the constituent object files themselves 
-are customarily included in the symbol tables, and that the symbol
+are customarily included in the symbol tables and that the symbol
 table entry format can be easily adopted for storing virtual addresses
@@
 about the post-relocation layout of the process 
-does not weaken ``randomization'', because 
+does not weaken ``randomization,'' for 
 the latter does not assume the attacker's ability to arbitrarily read 
@@
 must address several 
-important engineering issues, such as interaction of patched code with
-dynamically loaded libraries (including the {\em dlopen} mechanism),
+important engineering issues, such as the interaction of patched code with
+dynamically loaded libraries (including the {\em dlopen} mechanism)
 and assuring that accumulation of administered patches does not lead to
-unacceptable performance degradation. It also poses a broader question
-of describing and detecting software designs not amenable to its runtime
-patching and steering programmers to avoid them if possible.
+unacceptable performance degradation. 
+%JMHO: reword
+%% It also poses a broader question
+%% of describing and detecting software designs not amenable to its runtime
+%% patching and steering programmers to avoid them if possible. 
+We must also address the broader issue of describing and detecting
+software designs not amenable to runtime patching, and we must steer
+programmers to avoid them if possible.

### 2010-03-27 paper1/hotpatch/likelinking.tex
made figure one smaller
removed numbered subsection headings, as required by style for this publication
many grammar fixes on advice from a retired english teacher

@@
-\subsection{Patching In The Toolchain}
+%\subsection{Patching In The Toolchain}
+{\bf Patching In The Toolchain.}
 \label{ssec:patchingvlinking}
@@
 necessary to describe a patch. Once we take the position that patching
-is like linking, a patch needs to store the same type of symbol and
-relocation information as does any relocatable ELF object.
+is like linking, it follows that a patch needs to store the same type
+of symbol and relocation information as does any relocatable ELF
+object.
 
@@
 data. The DWARF format, already heavily used with ELF for debugging
-and exception handling purposes, provides exactly what is needed here
+and exception-handling purposes, provides exactly what is needed here
 as it provides a means to recursively describe types and variables, as
@@
 
-Through the use of formats already used in the binary tool chain, we
+Through the use of formats already employed in the binary tool chain, we
 hope to promote easy examination of patches, interoperability, and to
-show that patching fits comfortably in to the rest of the
+show that patching fits comfortably into the rest of the
 toolchain. We therefore propose that the standard software life cycle
-should now be as shown in Figure \ref{fig:lifecycle}. Before hot
+now be as shown in Figure \ref{fig:lifecycle}. Before hot
 patching, only the Development and Runtime stages of the figure were
@@
 
-\begin{figure*}[ht]
+\begin{figure}[ht]
 \begin{center}
-\includegraphics{software_lifecycle.pdf}
+\includegraphics[scale=0.5]{software_lifecycle.pdf}
 \end{center}
@@
 \label{fig:lifecycle}
-\end{figure*}
+\end{figure}
 

### 2010-03-27 paper1/hotpatch/patching.tex
many grammar fixes on advice from a retired english teacher

@@
 {\bf Code Patching.} %\hskip 0.1in
-This process involves several stages.
+This process involves several stages:
 %\newline
@@
 process. To do this, we consider the list of all modified object files
-from our tracking step, and identify all functions (both static and
+from our tracking step and identify all functions (both static and
 global) within these files 
@@
 within each function. Typically, symbol resolution for an application
-happens at both the linking stage (called {\it static linking}, when
+happens at both the linking stage (called {\it static linking} when
 the symbol is present within another object file or archive), and the
@@
 files and the final executable.  Each relocation entry contains, among
-other information, the code offset that requires relocation, and the
+other information, the code offset that requires relocation and the
 outstanding symbol that provides this fix-up.
@@
 symbol can be retrieved from the running target while performing the
-patching. This is key in allowing executables which have already been
+patching. This is key in allowing executables that have already been
 patched to be patched again.
@@
 execution, the dynamic linker maps the required shared libraries into
-the address space of the process, and updates each PLT entry.
+the address space of the process and updates each PLT entry.
 
@@
 to previously unused functions in shared libraries, but support for
-this will be added using the ``ALTPLT'' technique described in \cite{elf-debugging}.
+this will be added using the ``ALTPLT'' technique described by \cite{elf-debugging}.
 
@@
 Finally, if the outstanding symbol's definition was not found within
-the replicated executable (either within the symbol table or the PLT),
+the replicated executable (either within the symbol table or within the PLT),
 then it was newly added by the patch; it is marked as such and added
@@
 \newline
-(iii) {\it Patch Application:} Applying a code patch is simple enough,
+(iii) {\it Patch Application:} Applying a code patch is simple enough
 and has been researched in other
 systems~\cite{pannus,livepatch,ksplice,kaho}.  We map the new function
-in memory, and insert a trampoline {\tt jmp} instruction at the
+in memory and insert a trampoline {\tt jmp} instruction at the
 beginning of the old function within the process memory image. This
@@
 avoid the overhead (from branch mis-prediction) of the {\tt jmp}
-instruction by adding code in the old function which traces up the
+instruction by adding into the old function code that traces up the
 stack and modifies the caller's {\tt call} instruction operand to
@@
 implementation).  Even after identifying the allocated chunks of
-memory, in the absence of some kind of a type specification, the {\it
+memory, in the absence of some kind of type specification, the {\it
   internal structure} of memory remains opaque.  We also need a method
@@
 We first note that any code that acts on patch-modified data is
-already taken care of by Katana's code patching process. This is
+already taken care of by Katana's code patching process
 because we rely on {\tt make} to build the object files that
@@
 with debugging support, but we do not see this as a limitation. Since
-we only need DWARF information while building the PO, all debugging
+we need DWARF information only while building the PO, all debugging
 symbols could be stripped from the executable during application
-deployment, if so desired 
+deployment, if desired 
 %JMHO:
@@
 applying the patch. Else, the variable has been defined as either
-global or static. We first check if the replicated executable defines
+global or static. We first check whether the replicated executable defines
 the same variable. If not, then this variable has been created by the
-patch and we need not worry about it and leave the symbol resolution upto
+patch and we need not worry about it and may leave the symbol
+resolution up to
 the compiler (as only {\it new} code can use this variable). Otherwise,
@@
 we add the variable along with its {\it original} address from the replicated
-executable, its {\it new} address from the patch, and type information to the
+executable, its {\it new} address from the patch, and its type information to the
 PO. At the end of this stage, Katana would have identified all newly
@@
 first tracking down the relevant symbols in program memory. Katana
-reads in the PO, and for each data variable encountered, it checks if
+reads in the PO, and for each data variable encountered, it checks whether
 the variable is a pointer or not. If it is, then the current validity of the
@@
 the program text, that initializer will be used for any member
-variables within structures which or array elements which did not
+variables within structures or array elements that did not
 previously exist in the executing target. Initializers from the new
@@
 which unioned type to act on. If an array changes size, Katana will
-assume that it is growing or shrinking at the end, and copy old data
-accordingly, but will issue a warning that this may not be the desired
+assume that it is growing or shrinking at the end and will copy old data
+accordingly, but it will also issue a warning that this may not be the desired
 behavior. Pointers are of course handled, but with some limitations
@@
 avoidance work required to ensure that we do at some point pause all
-threads, but that we never pause a thread that another thread may be
+threads but that we never pause a thread that another thread may be
 waiting on in order to reach a safe state is nontrivial. We plan to

### 2010-03-27 paper1/hotpatch/poformat.tex
many grammar fixes on advice from a retired english teacher

@@
  \item type transformations specified using a language defined by the
-   Dwarf standard
+   DWARF standard
 \end{itemize}
@@
 utilizing a simple expression language for data patching, we aim to
-create patches which are easily examined (or modified) with existing
+create patches that are easily examined (or modified) with existing
 tools. This easy compatibility with the existing binary tools and
@@
 part of the ABI and of the standard toolchain? This does not
-necessarily have to be the precise format we use for Katana, any such
+necessarily have to be the precise format we use for Katana. Any such
 format that would become a standard, whether an actual standard or a
-de-facto standard should be well vetted by the community, but we argue
+de-facto standard, should be well vetted by the community, but we argue
 that something like this should be included in the standard object
@@
 cannot be determined automatically) already exist as the
-\texttt{.init} and \texttt{.fini} sections. Due to this similarity
+\texttt{.init} and \texttt{.fini} sections. Because of this similarity
 between some of the functionality needed by patching and the
@@
 libraries, some previous systems have performed patching by creating
-patches as dynamic libraries which contain not only the code and data
-to be patched but the mechanism to perform the patching \cite{ginseng}
+patches as dynamic libraries that contain not only the code and data
+to be patched but also the mechanism to perform the patching \cite{ginseng}
 \cite{polus}. We argue that
-this is an unnecessary mixing of data and logic, and that a patch
-which contains merely the information necessary to fix a running
+this is an unnecessary mixing of data and logic and, further, that a patch
+that contains merely the information necessary to fix a running
 process and not the code to do so is more desirable. The code to apply
@@
 
-On the other hand, note the specification of a general patch
+On the other hand, note that the specification of a general patch
 format does not completely prescribe the patch application. From a
@@
 binary to patch object to patched process in memory. This is all
-really nothing new, storing ELF sections to be injected in-memory has
+really nothing new; storing ELF sections to be injected in-memory has
 been done before in other systems \cite{eresi}. What is new in our patch object is
@@
 Note that we store considerably less information about each entity
-than is typically contained. This is because we read most of the
+than is typically contained. This is so because we read most of the
 information from the DWARF and symbol table information of the
-executing process (unless it will have been stripped, then more
+executing process (unless it will have been stripped; then more
 information must be stored in the patch). This allows the patch to be
@@
 between versions \emph{va} and \emph{vb} to patch both an executable
-which was originally compiled to \emph{va} and an executable which was
+that was originally compiled to \emph{va} and an executable that was
 patched from earlier versions to be equivalent to \emph{va}. Ksplice,
-one of the few other patchers which operates solely at the binary
+one of the few other patchers that operate solely at the binary
 level, does not have this capability \cite{ksplice}. We will provide a
-mechanism for composing patches, such that a patch from version
+mechanism for composing patches such that a patch from version
 \emph{va} to \emph{vb} may be composed with a patch from \emph{vb} to
@@
 }
-\caption{Dwarf DIEs in the PO}
+\caption{DWARF DIEs in the PO}
 \label{fig:dies}
@@
 in the \texttt{.debug\_frame} section of an FDE (Frame Description
-Entity). Dwarf FDEs are designed for use in transforming one call
+Entity). DWARF FDEs are designed for use in transforming one call
 frame into the previous call frame, and thereby walking up a call
-stack for either debugging purposes or exception handling purposes
+stack for either debugging purposes or exception-handling purposes
 (using the \texttt{.eh\_frame} section). Transforming one call frame
@@
 structure. We have aided this use with an implementation of the DWARF
-virtual machine which defines several special register types
+virtual machine that defines several special register types
 (exploiting the fact that for the purposes of generality, DWARF
@@
 \texttt{field3} from the original version of a structure \texttt{\_Foo}
-to a new version of \texttt{\_Foo} which has gained the extra member
+to a new version of \texttt{\_Foo} that has gained the extra member
 \texttt{field\_extra} in the middle of the existing fields would be
@@
 symbolic values defined by the virtual machine. If we are patching the
-variable 'bar', then the \texttt{CURR\_TARG\_OLD} will be the old address of bar
-(it's value in the symbol table) and \texttt{CURR\_TARG\_NEW} will be the new
+variable \texttt{bar}, then the \texttt{CURR\_TARG\_OLD} will be the old address of \texttt{bar}
+(its value in the symbol table), and \texttt{CURR\_TARG\_NEW} will be the new
 address bar is being relocated to. Our registers take advantage of the

### 2010-03-27 paper1/hotpatch/redundancy.tex
removed numbered subsection headings, as required by style for this publication
many grammar fixes on advice from a retired english teacher

--- 2010-03-02 paper1/hotpatch/redundancy.tex
@@
-\subsection{Why Not Just Employ Redundancy?}
+%\subsection{Why Not Just Employ Redundancy?}
+{\bf Why Not Just Employ Redundancy?}
 \label{ssec:redundancy}
@@
 
-\item {{\bf requires complicated logic -}} Transferring application
+\item {{\bf dependent upon complicated logic -}} Transferring application
   state (even across multiple homogenous systems) is non-trivial,
@@
 \end{enumerate}
+
+It should be noted that the last two items apply even for virtualized
+redundant systems, which often do not have the traditional overhead of
+redundant hardware.  We do not claim that redundancy does not have its
+place, but redundancy does not provide the easy, ubiquitous solution to
+high-availability stateful applications that we hope to provide
+through patching
+
+% LocalWords:  stateful virtualized

### 2010-03-27 paper1/hotpatch/related.tex
many grammar fixes on advice from a retired english teacher

@@
 There are several hot-patching systems preceding Katana. One of the
-most well known is probably Ginseng \cite{ginseng}. Ginseng (and
+most well-known is probably Ginseng \cite{ginseng}. Ginseng (and
 systems drawing inspiration from it such as Polus \cite{polus}) have
@@
 apache and sshd. These systems perform analysis of the differences
-between the original and patched versions at the source code
+between the original and the patched versions at the source code
 level. This introduces considerable (and we argue unnecessary)
@@
 agnostic as it works at the level of the binary ABI, and although we
-have not yet demonstrated it doing so, it should eventually be able to
-patch binaries compiled from any language, providing the necessary
+have not yet demonstrated its doing so, it should eventually be able to
+patch binaries compiled from any language, providing that the necessary
 symbol and relocation information is supplied. Ginseng also requires
@@
 or removing modules) without rebooting was achieved at least ten years
-ago~\cite{cesare1998} and recently rediscovered, albiet mostly for
+ago~\cite{cesare1998} and recently rediscovered, albeit mostly for
 research, rather than commodity,
@@
 patches~\cite{weimer2009genetic,stelios2005usenix} to gradually harden
-the application or seeking to avoid a restart altogether by modifying
+the application or by seeking to avoid a restart altogether by modifying
 certain runtime aspects ({\it e.g.,} the memory
@@
 One major risk of employing self-healing in production environments is
-that the semantics of follow-on execution remain largely uncontrolled,
+that the semantics of follow-on execution remains largely uncontrolled,
 although recent work in automatically correcting memory

### 2010-03-31 paper1/hotpatch/intro.tex
removed conficker quote

@@
 updates.
-
-{\footnotesize
-\begin{quote}
-``Some reports, such as the case of the Conficker outbreak within 
-Sheffield Hospital's operating ward, suggest that even security-
-conscious environments may elect to forgo automated software patching, 
-choosing to trade off vulnerability exposure for some perceived notion 
-of platform stability...'' -- \url{http://mtc.sri.com/Conficker/}
-\end{quote}
-}
 

### 2010-04-02 tests/t1_basic/v1/main.c
changes to test1 so that main doesn't need patching

@@ -12,6 +12,20 @@
 #include <string.h>
 #include <stdlib.h>
 
+#define MILLISECOND 1000
+
+void printThings();
+int main(int argc,char** argv)
+{
+  printf("has pid %i\n",getpid());
+  while(1)
+  {
+    printThings();
+    usleep(100*MILLISECOND);
+  }
+  return 0;
+}
+
 typedef struct _Foo
 {
   int field1;
@@ -29,15 +43,4 @@ void printThings()
   printf("Foo: %i,%i\n",bar.field1,bar.field2);
   printf("field 1 at addr: %x, field 2 at addr: %x\n",(unsigned int)&bar.field1,(unsigned int)&bar.field2);
   fflush(stdout);
-}
-
-int main(int argc,char** argv)
-{
-  printf("has pid %i\n",getpid());
-  while(1)
-  {
-    printThings();
-    usleep(100000);
-  }
-  return 0;
 }

### 2010-04-02 katana/tests/t1_basic/v0/main.c
changes to test1 so that main doesn't need patching

@@ -14,24 +14,9 @@
 #include <string.h>
 #include <stdlib.h>
 
-typedef struct _Foo
-{
-  int field1;
-  int field2;
-  int field3;
-} Foo;
-
-Foo bar={42,66,111};
-
 #define MILLISECOND 1000
 
-void printThings()
-{
-  printf("Foo: %i,%i\n",bar.field1,bar.field2);
-  printf("field 1 at addr: %lx, field 2 at addr: %lx\n",(unsigned long) &bar.field1,(unsigned long)&bar.field2);
-  fflush(stdout);
-}
-
+void printThings();
 int main(int argc,char** argv)
 {
   printf("has pid %i\n",getpid());
@@ -42,3 +27,22 @@ int main(int argc,char** argv)
   }
   return 0;
 }
+
+
+typedef struct _Foo
+{
+  int field1;
+  int field2;
+  int field3;
+} Foo;
+
+Foo bar={42,66,111};
+
+
+void printThings()
+{
+  printf("Foo: %i,%i\n",bar.field1,bar.field2);
+  printf("field 1 at addr: %lx, field 2 at addr: %lx\n",(unsigned long) &bar.field1,(unsigned long)&bar.field2);
+  fflush(stdout);
+}
+

### 2010-04-02 katana/src/relocation.c
added interface to get the section containing the relocations for a given function

@@ -457,3 +457,34 @@ addr_t getAddendForReloc(RelocInfo* reloc)
   }
   return computeAddend(reloc->e,reloc->relocType,reloc->symIdx,reloc->r_offset,reloc->scnIdx);
 }
+
+//get the section containing relocations for the given function
+//if want only the general relocation section, pass null for function name
+Elf_Scn* getRelocationSection(ElfInfo* e,char* fnname)
+{
+ //also include any relocations which exist for anything inside this function
+  //if -ffunction-sections is specified,
+  //each function will have its own relocation section,
+  //check this out first
+  char buf[1024];
+  snprintf(buf,1024,".rel.text.%s",fnname);
+  Elf_Scn* relocScn=getSectionByName(e,buf);
+  if(!relocScn)
+  {
+    snprintf(buf,1024,".rela.text.%s",fnname);
+    relocScn=getSectionByName(e,buf);
+  }
+  if(!relocScn)
+  {
+    relocScn=getSectionByName(e,".rel.text");
+  }
+  if(!relocScn)
+  {
+    relocScn=getSectionByName(e,".rela.text");
+  }
+  if(!relocScn)
+  {
+    death("function %s in %s does not have a .rel.text section\n",fnname,e->fname);
+  }
+  return relocScn;
+}

### 2010-04-02 src/relocation.h
added interface to get the section containing the relocations for a given function

@@ -88,4 +88,8 @@ addr_t getAddendForReloc(RelocInfo* reloc);
 
 //todo: does this belong in this module?
 addr_t getPLTEntryForSym(ElfInfo* e,int symIdx);
+
+//get the section containing relocations for the given function
+//if want only the general relocation section, pass null for function name
+Elf_Scn* getRelocationSection(ElfInfo* e,char* fnname);
 #endif

### 2010-04-02 src/patchwrite/codediff.c
use the new interface for getting the relocations section for a function to make other code simpler

@@ -43,35 +43,10 @@ bool areSubprogramsIdentical(SubprogramInfo* patcheeFunc,SubprogramInfo* patched
      variable of a changed type.
   2. If there is no relocation, return false if the bytes differ
   */
-  Elf_Scn* relocScn=NULL;
-  //if -ffunction-sections is specified,
-  //each function will have its own relocation section,
-  //check this out first
-  char buf[1024];
-  snprintf(buf,1024,".rel.text.%s",patcheeFunc->name);
-  relocScn=getSectionByName(oldBinary,buf);
-  if(!relocScn)
-  {
-    relocScn=getSectionByName(oldBinary,".rel.text");
-  }
-  if(!relocScn)
-  {
-    death("%s does not have a .rel.text section\n");
-  }
+  Elf_Scn* relocScn=getRelocationSection(oldBinary,patcheeFunc->name);
   List* oldRelocations=getRelocationItemsInRange(oldBinary,relocScn,patcheeFunc->lowpc,patcheeFunc->highpc);
-  //if -ffunction-sections is specified,
-  //each function will have its own relocation section,
-  //check this out first
-  snprintf(buf,1024,".rel.text.%s",patchedFunc->name);
-  relocScn=getSectionByName(newBinary,buf);
-  if(!relocScn)
-  {
-    relocScn=getSectionByName(newBinary,".rel.text");
-  }
-  if(!relocScn)
-  {
-    death("%s does not have a .rel.text section\n");
-  }
+
+  relocScn=getRelocationSection(newBinary,patchedFunc->name);
   List* newRelocations=getRelocationItemsInRange(newBinary,relocScn,patchedFunc->lowpc,patchedFunc->highpc);
 
   if(listLength(oldRelocations) != listLength(newRelocations))
@@ -87,6 +62,7 @@ bool areSubprogramsIdentical(SubprogramInfo* patcheeFunc,SubprogramInfo* patched
   sortList(newRelocations,&cmpRelocs);
   //if -ffunction-sections is used, the function might have its own text section
   Elf_Scn* textScn=NULL;
+  char buf[1024];
   snprintf(buf,1024,".text.%s",patcheeFunc->name);
   textScn=getSectionByName(oldBinary,buf);
   if(!textScn)

### 2010-04-02 src/patchwrite/patchwrite.c

@@ -477,20 +477,7 @@ addr_t writeFuncToPatchText(SubprogramInfo* func,CompilationUnit* cu,idx_t* outS
   *outSymIdx=addSymtabEntry(getDataByERS(patch,ERS_SYMTAB),&sym);
   
 
-  //also include any relocations which exist for anything inside this function
-  //if -ffunction-sections is specified,
-  //each function will have its own relocation section,
-  //check this out first
-  snprintf(buf,1024,".rel.text.%s",func->name);
-  Elf_Scn* relocScn=getSectionByName(cu->elf,buf);
-  if(!relocScn)
-  {
-    relocScn=getSectionByName(cu->elf,".rel.text");
-  }
-  if(!relocScn)
-  {
-    death("%s does not have a .rel.text section\n");
-  }
+  Elf_Scn* relocScn=getRelocationSection(cu->elf,func->name);
   writeRelocationsInRange(func->lowpc,func->highpc,relocScn,funcSegmentBase,cu->elf);
   return offset;
 }

### 2010-04-02 src/elfparse.c
elfparse looks for .rel.text sections

@@ -346,6 +346,12 @@ void findELFSections(ElfInfo* e)
     {
       e->sectionIndices[ERS_RELA_TEXT]=elf_ndxscn(scn);
     }
+    else if(!strncmp(".rel.text",name,strlen(".rel.text"))) //allow versioned
+                         //sections in patches. //todo: problem with this. Also allows section-specific ones, probably don't want this
+    {
+      e->sectionIndices[ERS_REL_TEXT]=elf_ndxscn(scn);
+    }
+
     else if(!strcmp(".plt",name))
     {
       e->sectionIndices[ERS_PLT]=elf_ndxscn(scn);

### 2010-04-02 src/elfparse.h

@@ -29,6 +29,7 @@ typedef enum
   ERS_STRTAB,
   ERS_HASHTABLE,
   ERS_RELA_TEXT,
+  ERS_REL_TEXT,
   ERS_PLT,
   ERS_REL_PLT,
   ERS_DYNSYM,

### 2010-04-02 src/patcher/linkmap.c
added a comment about a warning that seems to always exist in one of 32-bit or 64-bit no matter how I do the cast

@@ -171,6 +171,8 @@ addr_t locateRuntimeSymbolInTarget(ElfInfo* e,char* name)
   for(;;memcpyFromTarget((byte*)&lm,(addr_t)lm.l_next,sizeof(lm)))
   {
     addr_t addr;//store the result address
+    //can't seem to get rid of a sign cast warning in below line, it seems
+    //different library versions of libelf have different signdness for the param there
     if(locateSymbolInLinkMap(&lm,&addr,name,elf_hash(name)))
     {
       return addr;

### 2010-04-09 katana/src/dwarf_instr.c
cleanup: removed an unneeded print statement

@@ -273,6 +273,7 @@ uint leb128ToUInt(byte* bytes,usint* outLEBBytesRead)
   //valgrind gives this as a mem leak, but I can't figure out why,
   //as I free the result below. . .
   byte* result=decodeLEB128(bytes,false,&resultBytes,outLEBBytesRead);
+  printf("result bytes is %i\n",resultBytes);
   assert(resultBytes <= sizeof(uint));
   uint val=0;
   memcpy(&val,result,resultBytes);

### 2010-04-09 src/elfparse.c
added recognitionof .rela.plt section

@@ -360,6 +360,11 @@ void findELFSections(ElfInfo* e)
     {
       e->sectionIndices[ERS_REL_PLT]=elf_ndxscn(scn);
     }
+    else if(!strcmp(".rela.plt",name))
+    {
+      e->sectionIndices[ERS_RELA_PLT]=elf_ndxscn(scn);
+    }
+
     else if(!strcmp(".dynsym",name))
     {
       e->sectionIndices[ERS_DYNSYM]=elf_ndxscn(scn);

### 2010-04-09 src/elfparse.h

@@ -32,6 +32,7 @@ typedef enum
   ERS_REL_TEXT,
   ERS_PLT,
   ERS_REL_PLT,
+  ERS_RELA_PLT,
   ERS_DYNSYM,
   ERS_DYNSTR,
   ERS_DYNAMIC,

### 2010-04-09 katana/src/register.c
size field in a register is always in int, not word_t

@@ -14,6 +14,7 @@
 #include "symbol.h"
 #include <stdio.h>
 
+
 PoReg readRegFromLEB128(byte* leb,usint* bytesRead)
 {
   PoReg result;
@@ -36,10 +37,10 @@ PoReg readRegFromLEB128(byte* leb,usint* bytesRead)
   {
   case ERT_CURR_TARG_NEW:
   case ERT_CURR_TARG_OLD:
-    assert(numBytes>=1+sizeof(word_t) && numBytes<=1+2*sizeof(word_t));
-    memcpy(&result.size,bytes+1,sizeof(word_t));
+    assert(numBytes>=1+sizeof(int) && numBytes<=1+2*sizeof(int));
+    memcpy(&result.size,bytes+1,sizeof(int));
     result.u.offset=0;
-    memcpy(&result.u.offset,bytes+1+sizeof(word_t),min(numBytes-1-sizeof(word_t),sizeof(word_t)));
+    memcpy(&result.u.offset,bytes+1+sizeof(int),min(numBytes-1-sizeof(int),sizeof(int)));
     break;
   case ERT_EXPR:
     //todo: I forget what I was doing here
@@ -68,10 +69,10 @@ char* strForReg(PoReg reg)
   switch(reg.type)
   {
   case ERT_CURR_TARG_OLD:
-    snprintf(buf,128,"{CURR_TARG_OLD,0x%lx bytes,0x%x off}",(unsigned long)reg.size,reg.u.offset);
+    snprintf(buf,128,"{CURR_TARG_OLD,0x%x bytes,0x%x off}",(unsigned int)reg.size,reg.u.offset);
     break;
   case ERT_CURR_TARG_NEW:
-    snprintf(buf,128,"{CURR_TARG_NEW,0x%lx bytes,0x%x off}",(unsigned long)reg.size,reg.u.offset);
+    snprintf(buf,128,"{CURR_TARG_NEW,0x%x bytes,0x%x off}",(unsigned int)reg.size,reg.u.offset);
     break;
   case ERT_EXPR:
     //todo: print out expression

### 2010-04-09 katana/src/register.h

@@ -37,7 +37,7 @@ static inline bool isPoRegType(byte b)
 typedef struct
 {
   E_REG_TYPE type;
-  word_t size;//used for ERT_CURR_TARG_*
+  int size;//used for ERT_CURR_TARG_*
   union
   {
     int offset;//for ERT_CURR_TARG_*, ERT_EXPR

### 2010-04-09 src/patchwrite/write_to_dwarf.c
offset and size fields for register both int, not word_t (this should have been part of the last commit, I fail)

@@ -231,17 +231,17 @@ void writeTransformationToDwarf(Dwarf_P_Debug dbg,TypeTransform* trans)
   {
     DwarfInstruction inst;
     inst.opcode=DW_CFA_register;
-    byte bytes[1+sizeof(word_t)];
+    byte bytes[1+sizeof(int)];
     bytes[0]=ERT_CURR_TARG_NEW;
     assert(trans->from->length<=trans->to->length);
-    memcpy(bytes+1,&trans->to->length,sizeof(word_t));
+    memcpy(bytes+1,&trans->to->length,sizeof(int));
     //offset on the register is always 0, doing a complete copy of everything
-    inst.arg1Bytes=encodeAsLEB128(bytes,1+sizeof(word_t),false,&inst.arg1NumBytes);
+    inst.arg1Bytes=encodeAsLEB128(bytes,1+sizeof(int),false,&inst.arg1NumBytes);
     bytes[0]=ERT_CURR_TARG_OLD;
     //todo: should we just use the from length for everything, since it's the
     //shorter of the two
-    memcpy(bytes+1,&trans->from->length,sizeof(word_t));
-    inst.arg2Bytes=encodeAsLEB128(bytes,1+sizeof(word_t),false,&inst.arg2NumBytes);
+    memcpy(bytes+1,&trans->from->length,sizeof(int));
+    inst.arg2Bytes=encodeAsLEB128(bytes,1+sizeof(int),false,&inst.arg2NumBytes);
     addInstruction(&instrs,&inst);
     free(inst.arg1Bytes);
     free(inst.arg2Bytes);
@@ -260,14 +260,14 @@ void writeTransformationToDwarf(Dwarf_P_Debug dbg,TypeTransform* trans)
       writeTransformationToDwarf(dbg,transformer);
     }
     idx_t fdeIdx=transformer->fdeIdx;
-    byte bytes[1+sizeof(word_t)];
+    byte bytes[1+sizeof(int)];
     bytes[0]=ERT_CURR_TARG_NEW;
-    word_t size=sizeof(addr_t);
-    memcpy(bytes+1,&size,sizeof(word_t));
+    int size=sizeof(addr_t);
+    memcpy(bytes+1,&size,sizeof(int));
     //offset on the register is always 0
-    inst.arg1Bytes=encodeAsLEB128(bytes,1+sizeof(word_t),false,&inst.arg1NumBytes);
[…]

### 2010-04-09 katana/src/patcher/patchapply.c
trampoline jump and relocations fixup support x86_64

@@ -82,18 +82,30 @@ void relocateVar(VarInfo* var,ElfInfo* targetBin)
 
 void insertTrampolineJump(addr_t insertAt,addr_t jumpTo)
 {
-  printf("inserting at 0x%x, address to jump to is 0x%x\n",(uint)insertAt,(uint)jumpTo);
+  printf("inserting at 0x%zx, address to jump to is 0x%zx\n",insertAt,jumpTo);
   
   //remember that the JMP absolute is indirect, have to specify
   //memory location which hold the memory location to jump to
+#ifdef KATANA_X86_ARCH
   int len=2+sizeof(addr_t)*2;
   byte *code=zmalloc(len);
   code[0]=0xFF;//jmp instruction for a near absolute jump
   code[1]=0x25;//specify the addressing mode
-
   addr_t addrAddr=insertAt+2+sizeof(addr_t);//address of mem location holding jmp target
   memcpy(code+2,&addrAddr,sizeof(addr_t));
   memcpy(code+2+sizeof(addr_t),&jumpTo,sizeof(addr_t));
+  #elif defined(KATANA_X86_64_ARCH)
+  int len=2+4+sizeof(addr_t);
+  byte *code=zmalloc(len);
+  code[0]=0xFF;//jmp instruction for a near absolute jump
+  code[1]=0x25;//specify the addressing mode
+  addr_t addrAddr=0;//insertAt+2+4;//address of mem location holding jmp target
+  memcpy(code+2,&addrAddr,sizeof(addr_t));
+  memcpy(code+2+4,&jumpTo,sizeof(addr_t));
+  #else
+  #error Unknown architecture
+  #endif
+  
   memcpyToTarget(insertAt,code,len);
   //todo: probably don't need verify as memcpyToTarget can
   //be made to verify it's writings
@@ -498,8 +510,14 @@ void fixupPatchRelocations(ElfInfo* patch)
         //for something relative to the patch
         addr_t diff=patchTextAddr-oldTextNewStart;
         logprintf(ELL_INFO_V2,ELS_RELOCATION,"for PC32 relocation, modifying access at 0x%x to access 0x%x by adding 0x%x\n",(uint)newOffset,(uint)(addrAccessed+diff),(uint)diff);
-        modifyTarget(newOffset,addrAccessed+diff);
-        //todo: I'm not sure this is necessary here, since we do relocations later
+        addr_t newAddr=addrAccessed+diff;
[…]

### 2010-04-09 katana/src/patchwrite/patchwrite.c
always init dwarf producer objects as 32-bit. Note that even when working with x86_64 executables, it's usually possible to have 32-bit dwarf objects. The dwarf standard recommends against generating 64-bit dwarf objects by default. What we lack, and should have, is a method to detect if a 64-bit dwarf object is necessary and generate one if it is

@@ -698,7 +698,11 @@ void writePatch(char* oldSourceTree,char* newSourceTree,char* oldBinName,char* n
   findELFSections(oldBinary);
   patch=startPatchElf(patchOutName);
   Dwarf_Error err;
-  dbg=dwarf_producer_init(DW_DLC_WRITE,dwarfWriteSectionCallback,dwarfErrorHandler,NULL,&err);
+  //todo: always creating 32-bit Dwarf sections. DWARF standard
+  //recommends never having 64-bit Dwarf as the default, even on
+  //x86_64. We should have some way of recognizing when we want 64-bit
+  //though, if things are really huge.
+  dbg=dwarf_producer_init(DW_DLC_WRITE|DW_DLC_SIZE_32,dwarfWriteSectionCallback,dwarfErrorHandler,NULL,&err);
   Dwarf_P_Die tmpRootDie=dwarf_new_die(dbg,DW_TAG_compile_unit,NULL,NULL,NULL,NULL,&err);
   dwarf_add_die_to_debug(dbg,tmpRootDie,&err);
   cie=dwarf_add_frame_cie(dbg,"",1,1,0,NULL,0,&err);

### 2010-04-09 katana/src/relocation.c
towards fully supporting relocations in x86_64

@@ -77,20 +77,55 @@ addr_t getPLTEntryForSym(ElfInfo* e,int symIdx)
   GElf_Shdr shdr;
   getShdrByERS(e,ERS_PLT,&shdr);
   addr_t pltStart=shdr.sh_addr;
+
+  Elf_Scn* pltRelScn=NULL;
+  if(e->sectionIndices[ERS_REL_PLT])
+  {
+    pltRelScn=getSectionByERS(e,ERS_REL_PLT);
+  }
+  else if(e->sectionIndices[ERS_RELA_PLT])
+  {
+    pltRelScn=getSectionByERS(e,ERS_RELA_PLT);
+  }
+  else
+  {
+    //todo: maybe don't die here, just return 0
+    death("Executable has neither rel.plt nor rela.plt\n");
+  }
+  
   //plt entries are aligned on 16-byte boundaries
   //we start with plt1, since plt0 has special function
+  //todo: support x86_64 large code PLT model (which is much more
+  //complicated and things are not separated nicely by 16 bytes)
   int i=0x10;
   for(;i<shdr.sh_size;i+=0x10)
   {
-    word_t relocOffset;
+    word_t relocOffset=0;
+    addr_t startWithOffset;
     //the first jmp instruction takes 6 bytes, then the
     //opcode of the push instruction takes 1 byte, so the reloc_offset
     //is after 7 bytes
+    startWithOffset=pltStart+7;
+    int offsetLen=4;
+    #if defined(KATANA_X86_64_ARCH)
+    //todo: support large-mode plt
+    //in which  offsetLen will be 8
+    //not supported yet because I'm not sure how to detect it best
+    #endif
+    
     //todo: does this need to be aligned, because it isn't
-    memcpyFromTarget((byte*)&relocOffset,pltStart+i+7,sizeof(word_t));
+    memcpyFromTarget((byte*)&relocOffset,startWithOffset+i,offsetLen);
     logprintf(ELL_INFO_V4,ELS_RELOCATION,"relocOffset is %i\n",relocOffset);
     //now look up the appropriate relocation entry
-    RelocInfo* reloc=getRelocationEntryAtOffset(e,getSectionByERS(e,ERS_REL_PLT),relocOffset);
+#if defined(KATANA_X86_64_ARCH)
+    //relocationOffset is actually an index, not an offset. We have to make it into
+    //an offset therefore
+    GElf_Shdr pltRelShdr;
+    getShdr(pltRelScn,&pltRelShdr);
[…]

### 2010-04-09 src/util/util.h
cleaned up MALLOC_CHECK macro a little

@@ -30,9 +30,8 @@ void death(char* reason,...);
 
 //! \brief Check whether \a s is NULL or not on a memory allocation. Quit this program if it is NULL.
 #define MALLOC_CHECK(s)  if ((s) == NULL)   {                     \
-    fprintf(stderr,"No enough memory at %s:line%d ", __FILE__, __LINE__); \
-    perror(":");                                                  \
-    exit(-1); \
+      perror("system error:");                                                  \
+      death("No enough memory at %s:line %d ", __FILE__, __LINE__); \
   }
 
 //! \brief Set memory space starts at pointer \a n of size \a m to zero. 

### 2010-04-09 katana/src/symbol.c
logprintf(ELL_WARN rather than fprintf(stderr

@@ -275,7 +275,7 @@ int reindexSymbol(ElfInfo* old,ElfInfo* new,int oldIdx,int flags)
   if(STN_UNDEF==idx)
   {
     char* symbolName=getString(old,sym.st_name);
-    fprintf(stderr,"Symbol '%s' could not be reindexed\n",symbolName);
+    logprintf(ELL_WARN,ELS_SYMBOL,"Symbol '%s' could not be reindexed\n",symbolName);
   }
   return idx;
 }

### 2010-04-09 paper1/hotpatch/poformat.tex
final last changes to the paper before publication

@@
 {\bf Reasons and Needs.}
-We have developed a Patch Object format with the following properties
+We have developed a Patch Object format for which the following holds:
 \begin{itemize}
- \item a valid ELF file
- \item information about types and functions requiring patches contained
-   as DWARF information
- \item type transformations specified using a language defined by the
-   DWARF standard
+ \item A PO is a valid ELF file.
+ \item A PO utilizes DWARF information to describe types, variables, and functions requiring patching.
+ \item A PO allows type transformations to be specified using a language defined by the DWARF standard.
 \end{itemize}

### 2010-05-24 doc/katana.tex
documentation work

% Created 2010-05-24 Mon 14:34
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{amssymb}
\usepackage{hyperref}

\title{Katana: A Userland Toolchain-Oriented Hotpatching System}
\author{James Oakley}
\date{24 May 2010}

\begin{document}

\maketitle

\setcounter{tocdepth}{3}
\tableofcontents
\vspace*{1cm}

\section{Introduction}
\label{sec-1}

\section{What Katana Does}
\label{sec-2}

\section{What Katan Does Not Do (Yet)}
\label{sec-3}

\section{How to Use Katana}
\label{sec-4}

  Katana is intended to be used in two stages. The first stage
  generates a patch object from two different versions of an
  treee. By an object tree, we mean the set of object files (.o files)
  and the executable binary they comprise. Katana works completely at
  the object level, so the source code itself is not strictly
  required, although all objects must be compiled with debugging
  information. This step may be done by the software vendor. In the
  second stage, the patch is applied to a running process. The
  original source trees are not necessary during patch application, as
  the patch object contains all information necessary to patch the
  in-memory process at the object level. It is also possible to view
  the contents of a patch object in a human-readable way for the
  purposes of sanity-checking, determining what changes the patch
  makes, etc.
\subsection{Preparing a Package for Patching Support}
\label{sec-4.1}

   Katana aims to be much less invasive than other hot-patching system
   and require minimal work to be used with any project. It does,
   however, have some requirements.\\
   Required CFLAGS:
\begin{itemize}
\item -g
\end{itemize}
   Recommended CFLAGS:
\begin{itemize}
\item -ffunction-sections
\item -fdata-sections
\end{itemize}
   Recommended LDFLAGS:
\begin{itemize}
\item --emit-relocs
\end{itemize}
\subsection{To Generate a Patch}
\label{sec-4.2}

\subsection{To Apply a Patch}
\label{sec-4.3}

\subsection{To View a Patch}
\label{sec-4.4}

\subsection{See Also}
\label{sec-4.5}

   the katana manpage
\section{Patch Object Format}
\label{sec-5}

\section{Patch Generation Process}
\label{sec-6}

\section{Patch Application Process}
\label{sec-7}

\section{Roadmap}
\label{sec-8}

\end{document}

### 2010-06-16 katana/src/katana.c
whitespace and comment fixes

@@ -112,15 +112,15 @@ int main(int argc,char** argv)
     case 's':
       if(EKM_NONE==mode)
       {
-	death("Must specify a mode (-g,-p, or -l) before any other options");
+        death("Must specify a mode (-g,-p, or -l) before any other options");
       }
       if(EKM_PATCH_INFO==mode)
       {
-	flags|=PF_STOP_TARGET;
+        flags|=PF_STOP_TARGET;
       }
       else
       {
-	death("-s flag has no meaning for this mode\n");
+        death("-s flag has no meaning for this mode\n");
       }
       
     }

### 2010-06-16 doc/katana.tex
updated documentation

@@
-% Created 2010-05-24 Mon 14:34
+% Created 2010-06-15 Tue 22:01
 \documentclass[11pt]{article}
@@
 \usepackage{hyperref}
+\usepackage{graphicx}
 
@@
 \author{James Oakley}
-\date{24 May 2010}
+\date{15 June 2010}
 
@@
 
-\section{What Katana Does}
+  Katana aims to provide a hot-patching system for userland. Further
+  it aims to work with existing toolchains and formats so as to be
+  easy to use and to hopefully pave the way for incorporating patching
+  as a standard part of the toolchain. Because of this aim, Katana
+  operates at the object level rather than requiring any access to the
+  source code itself. This has the added bonus of making it, in
+  theory, language agnostic (although no work has been done to test it
+  with anything besides programs written in C). A diagram of software
+  lifecycle with hotpatching is shown below
+
+\begin{figure}[h!]
+\includegraphics[width=3in]{./softwarelifecycle.pdf}
+\end{figure}
+
+  This document is intended to provide a users guide to Katana,
+  insight into its inner workings, and discussion of its flaws and
+  plans for the future. As the software is not complete, making use of
+  Katana without understanding the inner workings and technical
+  shortcomings is not recommended. Nevertheless, the only sections of
+  this document necessary for ``Users' Guide'' purposes are 
+  \hyperref[sec-3]{``What Katana Does''}, \hyperref[sec-4]{``What Katana Does Not Do (Yet)''}, and most importantly 
+  \hyperref[sec-6]{``How to Use Katana''}.
+
+  This document is a work in progress. It is not a polished guide yet.
+\section{Other Systems}
 \label{sec-2}
 
-\section{What Katan Does Not Do (Yet)}
+  There are other hotpatching systems in existence. The curious are
+  invited to explore Ginseng and Polus. Both of these systems parse
+  the source code, which adds significant complexity to them and
+  results in significant programmer annotation of the code to give
+  hints to the systems. Ginseng uses complicated type-wrappers
+  when patching variables which does not fit cleanly with existing
+  executables and has some impact on the performance of the
+  software. Ginseng is considerably more mature than Katana,
+  however. Neither system is production ready, but Ginseng is probably
+  closer than Katana at the moment.
+
+  The system most like Katana in many ways is KSplice, and the curious
+  reader is definitely invited to investigate. KSplice patches the
+  kernel and not userland, does not attempt to patch variables, and
+  creates patches as kernel modules rather than working towards a
+  general ELF-based patch format.
+\section{What Katana Does}
 \label{sec-3}
 
+\begin{itemize}
+\item Runs on x86 and x86-64
+\item Generates patches for simple programs
+\item Applies simple patches
+\end{itemize}
+\section{What Katana Does Not Do (Yet)}
+\label{sec-4}
+
+\begin{itemize}
+\item Patch any major programs: it has not yet been demonstrated on
+    anything more than toy examples
+\item Provide any method to handle opaque data it cannot patch (void*,
+    situations where which action a user would prefer is unclear, etc)
+\item Patch previously patched processes
+\item Provide robust operation
+\item Run on any architectures other than x86 and x86-64
+\item Tested on any operating system besides Linux
+\item Allow for calls in patched code to previously unused functions
+\item Work for programs which actually make use of some of the large
+    code model features of the x86-64 ABI.
+\item And much more
+\end{itemize}
+  See \hyperref[sec-10]{Roadmap} for more things which are not complete
+
+\section{What Katana May Never Do}
+\label{sec-5}
+
+\begin{itemize}
+\item Work on any binary formats besides ELF
+\end{itemize}
 \section{How to Use Katana}
-\label{sec-4}
+\label{sec-6}
 
@@
 \subsection{Preparing a Package for Patching Support}
-\label{sec-4.1}
+\label{sec-6.1}
 
@@
 \subsection{To Generate a Patch}
-\label{sec-4.2}
+\label{sec-6.2}
 
+   Let the location of your project be /project. You must have two
+   versions of your software available: the version identical to the
+   running software which must be hotpatched, call it v0, and the
+   version to which you wish to hotpatch the running software, call it
+   v1. Let foo be the name of your program. Then /project/v0/foo must
+   exist and /project/v0 must also contain (possibly in
+   subdirectories) all of the object files which contributed to
+   /project/v0/foo. The source code itself is immaterial, as Katana
+   does not parse it. Similarly, /project/v1/foo must exist and
+   /project/v1 contain all of the object files contributing to
+   /project/v1/foo. Katana is then invoked as
+
+   \texttt{katana -g [-o OUTPUT\_FILE] /project/v0 /project/v1 foo}
+
+   or more formally
+
+   \texttt{katana -g [-o OUTUT\_FILE] OLD\_OBJECTS\_DIR NEW\_OBJECTS\_DIR EXECUTABLE\_NAME}
+
+   If \texttt{-o OUTPUT\_FILE} is not specified, the output file will be \texttt{OLD\_OBJECTS\_DIR/EXECUTABLE\_NAME.po}
 \subsection{To Apply a Patch}
-\label{sec-4.3}
+\label{sec-6.3}
 
+   The process to be patched is running with a pid of PID. It can be
+   patched from its current version to a more recent version by the
+   Patch Object (PO) file PATCH. Katana is then invoked as
+
+   \texttt{katana -p [-s] PATCH PID}
+
+   If all goes well, the patcher will run, print out some status
+   messages, and leave your program in better state than it found
+   it. The optional -s flag tells Katana to stop the target program
+   after patching it and detaching from it. This is mostly of use for
+   debugging Katana.
 \subsection{To View a Patch}
-\label{sec-4.4}
+\label{sec-6.4}
 
+   One of the goals of Katana and its Patch Object (PO) format is to
+   increase the transparency of patches: a user about to apply a patch
+   should know what it will do. This goal is not yet fully realized,
+   but it is possible to view some information about a patch with
+
+   \texttt{katana -l PATCH}
 \subsection{See Also}
-\label{sec-4.5}
+\label{sec-6.5}
 
-   the katana manpage
+   the katana manpage (once it's written, which it is not yet)
 \section{Patch Object Format}
-\label{sec-5}
-
-\section{Patch Generation Process}
-\label{sec-6}
-
-\section{Patch Application Process}
 \label{sec-7}
 
-\section{Roadmap}
+  This section of the document is not yet written. It will provide a description and specification of the PO format used by Katana
+\section{Patch Generation Process}
 \label{sec-8}
 
+  This section of the document is not yet written. It will provide a
+  description of the internal process that Katana uses to generate a
+  patch. Understanding it is not necessary for using Katana.
+\section{Patch Application Process}
+\label{sec-9}
+
+  This section of the document is not yet written. It will provide a
+  description of the internal process that Katana uses to apply a
+  patch. Understanding it is not necessary for using Katana.
+\section{Roadmap}
+\label{sec-10}
+
+  This section is highly incomplete. Future goals include
+\begin{itemize}
+\item Better interaction with the heap and dynamically allocated variables
+\item Better interaction with void*
+\item More efficient use of .rodata
+\item Patching already patched processes
+\item Patch composition
+\item Patch safety checking: make sure a patch actually corresponds to
+    the process it's being applied to
+\item Storing warnings from generation inside a patch
+\end{itemize}
+
 \end{document}

### 2010-06-21 doc/katana.tex
some fixes for licensing. Documentation is under GNU FDL

@@
-% Created 2010-06-15 Tue 22:01
+% Created 2010-06-20 Sun 22:23
 \documentclass[11pt]{article}
@@
 \author{James Oakley}
-\date{15 June 2010}
+\date{20 June 2010}
 
@@
 \item Run on any architectures other than x86 and x86-64
-\item Tested on any operating system besides Linux
+\item Tested on any operating system besides GNU/Linux
 \item Allow for calls in patched code to previously unused functions
@@
 \end{itemize}
+\section{Credits and Licensing}
+\label{sec-11}
+
+  Katana is under development at Dartmouth College and Copyright 2010
+  Dartmouth College. It may be distributed under the terms of the GNU
+  General Public License with attribution to Dartmouth College as
+  specified in the file COPYING distributed with Katana. This document
+  is Copyright 2010 Dartmouth College and may be distributed under the
+  terms of the GNU Free Documentation License
+
+  Katana is being written by James Oakley and was designed
+  by Sergey Bratus, Ashwin Ramaswamy, James Oakley, Michael Locasto,
+  and Sean Smith.
 

### 2010-06-21 run_tests.py

@@ -1,6 +1,8 @@
 #!/usr/bin/env python
 #File: run_tests.py
 #Author: James Oakley
+#Copyright (C): 2010 Dartmouth College
+#License: GNU General Public License
 #Date: January, 2010
 #Description: runs unit tests and reports their success or failure
 

### 2010-06-21 src/patcher/pmap.c

@@ -8,7 +8,7 @@
   Date: April 2010
   Description: Find out information about the target processes memory map.
                Requires a linux-like /proc filesystem. It should be useable
-               on BSD and Solaris as well as Linux, but these have not been tested.
+               on BSD and Solaris as well as GNU/Linux, but these have not been tested.
                FreeBSD seems to be getting rid of /proc
 
 */

### 2010-06-21 src/patcher/pmap.h

@@ -7,7 +7,7 @@
   Date: April 2010
   Description: Find out information about the target processes memory map.
                Requires a linux-like /proc filesystem. It should be useable
-               on BSD and Solaris as well as Linux, but these have not been tested.
+               on BSD and Solaris as well as GNU/Linux, but these have not been tested.
                FreeBSD seems to be getting rid of /proc
 
 */

### 2010-06-21 src/patcher/versioning.c

@@ -40,7 +40,7 @@ ElfInfo* getElfRepresentingProc(int pid)
     struct stat s;
     if(0!=stat(execPath,&s))
     {
-      fprintf(stderr,"%s does not exist. Is this a linux system or other unix system with a Plan-9 style /proc filesystem? If it is, then the process may have exited",execPath);
+      fprintf(stderr,"%s does not exist. Is this a GNU/Linux system or other unix system with a Plan-9 style /proc filesystem? If it is, then the process may have exited",execPath);
       death(NULL);
     }
     result=openELFFile(execPath);

### 2010-06-21 tests/t1_basic/v0/main.c

@@ -1,9 +1,9 @@
 /*
   File: main_v0.c
   Author: James Oakley
-  Written For: Dartmouth College
+  Copyright (C): 2010 Dartmouth College
   License: GNU General Public License
-  Project: Katana - Preliminary Work
+  Project: Katana
   Date: January 10
   Description: Very simple program that exists to have one of its data
                types patched. main_v1.c is the same thing with an extra

### 2010-06-21 tests/t1_basic/v1/main.c

@@ -1,9 +1,9 @@
 /*
   File: main_v1.c
   Author: James Oakley
-  Written For: Dartmouth College
+  Copyright (C): 2010 Dartmouth College
   License: GNU General Public License
-  Project: Katana - Preliminary Work
+  Project: Katana
   Date: January 10
   Description: Very simple program that exists to patch a data type in main_v0.
                main_v0.c is the same thing with one less field in the Foo type

### 2010-06-21 tests/t1_basic/validate.py

@@ -1,6 +1,6 @@
 #File: validate.py
 #Author: James Oakley
-#Written For: Dartmouth College
+#Copyright (C): 2010 Dartmouth College
 #License: GNU General Public License
 #Date, January, 2010
 #Description: validate the output of a single unit test for katana

### 2010-06-21 tests/t2_nested_types/v0/main.c

@@ -1,9 +1,9 @@
 /*
   File: main.c
   Author: James Oakley
-  Written For: Dartmouth College
+  Copyright (C): 2010 Dartmouth College
   License: GNU General Public License
-  Project: Katana (testing)
+  Project: Katana
   Date: February 2010
   Description: Test program to demonstrate patching a modified type
                which itself has a member which is a modified type

### 2010-06-21 tests/t3_multiple_cu/v0/main.c

@@ -1,9 +1,9 @@
 /*
   File: main.c v0
   Author: James Oakley
-  Written For: Dartmouth College
+  Copyright (C): 2010 Dartmouth College
   License: GNU General Public License
-  Project: Katana - Testing
+  Project: Katana
   Date: January 2010
   Description: Very simple program that exists to have one of its data
                types patched. main_v1.c is the same thing with an extra

### 2010-06-21 tests/t3_multiple_cu/v1/main.c

@@ -1,9 +1,9 @@
 /*
   File: main.c v1
   Author: James Oakley
-  Written For: Dartmouth College
+  Copyright (C): 2010 Dartmouth College
   License: GNU General Public License
-  Project: Katana - Test Harness
+  Project: Katana
   Date: January 10
   Description: Very simple program that exists to have one of its data
                types patched. main_v1.c is the same thing with an extra

### 2010-06-21 tests/t3_multiple_cu/v1/other.c

@@ -1,9 +1,9 @@
 /*
   File: other2.c
   Author: James Oakley
-  Written For: Dartmouth College
+  Copyright (C): 2010 Dartmouth College
   License: GNU General Public License
-  Project: Katana - Testing
+  Project: Katana
   Date: January 2010
   Description: different definition of struct Foo than in main.c to make sure
                patching system treats the different versions properly

### 2010-06-21 tests/t4_new_var/v0/main.c

@@ -1,7 +1,7 @@
 /*
   File: main.c
   Author: James Oakley
-  Written For: Dartmouth College
+  Copyright (C): 2010 Dartmouth College
   License: GNU General Public License
   Project: Katana (testing)
   Date: February 2010

### 2010-06-21 tests/t4_new_var/v1/main.c

@@ -1,9 +1,10 @@
 /*
   File: main.c
   Author: James Oakley
-  Written For: Dartmouth College
+  Copyright (C): 2010k
+  Dartmouth College
   License: GNU General Public License
-  Project: Katana (testing)
+  Project: Katana
   Date: February 2010
   Description: Test program to demonstrate adding a variable to a program
 */

### 2010-06-21 tests/t4_new_var/v1/other.c

@@ -1,7 +1,7 @@
 /*
   File: other.c
   Author: James Oakley
-  Written For: Dartmouth College
+  Copyright (C): 2010 Dartmouth College
   License: GNU General Public License
   Project: Katana
   Date: February 2010

### 2010-06-21 tests/t4_new_var/validate.py

@@ -1,6 +1,6 @@
 #File: validate.py
 #Author: James Oakley
-#Written For: Dartmouth College
+#Copyright (C): 2010 Dartmouth College
 #License: GNU General Public License
 #Date, February, 2010
 #Description: validate the output of a single unit test for katana

### 2010-06-21 tests/t5_pointers/v0/main.c

@@ -1,7 +1,7 @@
 /*
   File: v0/main.c
   Author: James Oakley
-  Written For: Dartmouth College
+  Copyright (C): 2010 Dartmouth College
   License: GNU General Public License
   Project: Katana
   Date: January 10

### 2010-06-21 tests/t5_pointers/v1/main.c

@@ -1,7 +1,7 @@
 /*
   File: v1/main.c
   Author: James Oakley
-  Written For: Dartmouth College
+  Copyright (C): 2010 Dartmouth College
   License: GNU General Public License
   Project: Katana - Preliminary Work
   Date: January 10

### 2010-06-21 tests/t6_unions_enums/v1/main.c

@@ -1,7 +1,7 @@
 /*
   File: v1/main.c
   Author: James Oakley
-  Written For: Dartmouth College
+  Copyright (C): 2010 Dartmouth College
   License: GNU General Public License
   Project: Katana
   Date: January 10

### 2010-06-21 tests/t6_unions_enums/validate.py

@@ -1,6 +1,6 @@
 #File: validate.py
 #Author: James Oakley
-#Written For: Dartmouth College
+#Copyright (C): 2010 Dartmouth College
 #License: GNU General Public License
 #Date, March, 2010
 #Description: validate the output of a single unit test for katana

### 2010-06-21 tests/t7_const/v0/main.c

@@ -3,7 +3,7 @@
   Author: James Oakley
   Written For: Dartmouth College
   License: GNU General Public License
-  Project: Katana - Preliminary Work
+  Project: Katana
   Date: January 10
   Description: Very simple program that exists to have one of its data
                types patched. v1/main.c is the same thing with slight

### 2010-06-21 tests/t7_const/v1/main.c

@@ -1,9 +1,9 @@
 /*
   File: v0/main.c
   Author: James Oakley
-  Written For: Dartmouth College
+  Copyright (C): 2010 Dartmouth College
   License: GNU General Public License
-  Project: Katana - Preliminary Work
+  Project: Katana
   Date: January 10
   Description: Very simple program that exists to have one of its data
                types patched. v1/main.c is the same thing with slight

### 2010-06-21 validator.py

@@ -1,6 +1,6 @@
 #File: validator.py
 #Author: James Oakley
-#Written For: Dartmouth College
+#Copyright (C): 2010 Dartmouth College
 #License: GNU General Public License
 #Date, January, 2010
 #Description: validate the output of a single unit test for katana

### 2010-06-22 src/katana.c
removed pid as a global variable

@@ -2,7 +2,48 @@
   File: typepatch.c
   Author: James Oakley
   Copyright (C): 2010 Dartmouth College
-  License: GNU General Public License
+  License: Katana is free software: you may redistribute it and/or
+    modify it under the terms of the GNU General Public License as
+    published by the Free Software Foundation, either version 3 of the
+    License, or (at your option) any later version. Regardless of
+    which version is chose, the following stipulation also applies:
+    
+    Any redistribution must include copyright notice attribution to
+    Dartmouth College as well as the Warranty Disclaimer below, as well as
+    this list of conditions in any related documentation and, if feasible,
+    on the redistributed software; Any redistribution must include the
+    acknowledgment, “This product includes software developed by Dartmouth
+    College,” in any related documentation and, if feasible, in the
+    redistributed software; and The names “Dartmouth” and “Dartmouth
+    College” may not be used to endorse or promote products derived from
+    this software.  
+
+                             WARRANTY DISCLAIMER
+
+    PLEASE BE ADVISED THAT THERE IS NO WARRANTY PROVIDED WITH THIS
+    SOFTWARE, TO THE EXTENT PERMITTED BY APPLICABLE LAW. EXCEPT WHEN
+    OTHERWISE STATED IN WRITING, DARTMOUTH COLLEGE, ANY OTHER COPYRIGHT
+    HOLDERS, AND/OR OTHER PARTIES PROVIDING OR DISTRIBUTING THE SOFTWARE,
+    DO SO ON AN "AS IS" BASIS, WITHOUT WARRANTY OF ANY KIND, EITHER
+    EXPRESSED OR IMPLIED, INCLUDING, BUT NOT LIMITED TO, THE IMPLIED
+    WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
+    PURPOSE. THE ENTIRE RISK AS TO THE QUALITY AND PERFORMANCE OF THE
+    SOFTWARE FALLS UPON THE USER OF THE SOFTWARE. SHOULD THE SOFTWARE
+    PROVE DEFECTIVE, YOU (AS THE USER OR REDISTRIBUTOR) ASSUME ALL COSTS
[…]

### 2010-06-22 src/patcher/patchapply.c

@@ -723,7 +723,7 @@ void katanaPLT()
 
 void readAndApplyPatch(int pid,ElfInfo* targetBin_,ElfInfo* patch,int flags)
 {
-  startPtrace();
+  startPtrace(pid);
   targetBin=targetBin_;
   
   

### 2010-06-22 src/patcher/target.c

@@ -23,8 +23,8 @@
 #include "util/logging.h"
 #include "util/map.h"
 
-//todo: bad programming to have globals like this
-extern int pid;
+
+int pid;
 addr_t mallocAddress=0;
 addr_t targetTextStart=0;
 
@@ -45,8 +45,9 @@ void setTargetTextStart(addr_t addr)
   targetTextStart=addr;
 }
 
-void startPtrace()
+void startPtrace(int pid_)
 {
+  pid=pid_;
   if(ptrace(PTRACE_ATTACH,pid,NULL,NULL)<0)
   {
     fprintf(stderr,"ptrace failed to attach to process\n");
@@ -199,7 +200,6 @@ void memcpyToTarget(addr_t addr,byte* data,int numBytes)
 bool memcpyFromTargetNoDeath(byte* data,long addr,int numBytes)
 {
   logprintf(ELL_INFO_V4,ELS_HOTPATCH,"memcpyFromTarget: getting %i bytes from 0x%x\n",numBytes,(uint)addr);
-  //todo: force address to be aligned
   for(int i=0;i<numBytes;i+=4)
   {
     uint val=ptrace(PTRACE_PEEKDATA,pid,addr+i);

### 2010-06-22 src/patcher/target.h

@@ -24,7 +24,9 @@
 #include "arch.h"
 //#include <sys/user.h>
 
-void startPtrace();
+//this must be called before any other functions in this file
+void startPtrace(int pid);
+
 void continuePtrace();
 void endPtrace(bool stopProcess);
 void modifyTarget(addr_t addr,word_t value);

### 2010-06-22 src/dwarftypes.c
fixed some logging and error messages

@@ -655,7 +655,7 @@ void* addStructureFromDie(Dwarf_Debug dbg,Dwarf_Die die,CompilationUnit* cu)
       if(DW_DLV_OK==res)
       {
         type->fieldOffsets[idx]=readAttributeAsAddr(attr,dbg,0);
-        printf("from location attribute, set offset for %s.%s to %i\n",name,type->fields[idx],type->fieldOffsets[idx]);
+        logprintf(ELL_INFO_V4,ELS_DWARFTYPES,"from location attribute, set offset for %s.%s to %i\n",name,type->fields[idx],type->fieldOffsets[idx]);
       }
       else
       {

### 2010-06-22 src/elfparse.c

@@ -32,7 +32,7 @@ ElfInfo* openELFFile(char* fname)
   e->e=elf_begin(e->fd,ELF_C_READ,NULL);
   if(!e->e)
   {
-    fprintf(stderr,"Failed to open as an ELF file %s\n",elf_errmsg(-1));
+    fprintf(stderr,"Failed to open file %s as an ELF file %s\n",fname,elf_errmsg(-1));
     free(e);
     death(NULL);
   }

### 2010-06-22 doc/katana.tex
updated documentation

@@
-% Created 2010-06-20 Sun 22:23
+% Created 2010-06-20 Sun 23:06
 \documentclass[11pt]{article}
@@
     however, have some requirements.\\
+\subsubsection{Source Code Practices}
+\label{sec-6.1.1}
+
+    Katana does not look at the source code, therefore unlike several
+    other hotpatching systems, it does not require any annotation in
+    the source code. There are, however, some best practices to
+    follow.
+\begin{itemize}
+\item Avoid the use of \texttt{void*} at least for global variables (since
+      Katana does not currently patch local variables, preferring to
+      wait until any functions using changed variables are no longer
+      on the stack). Since it is typeless and opaque, it is very hard
+      to analyze and patch.
+\item Avoid unnamed types. i.e., instead of \texttt{typedef struct \{...\} Foo;}
+      use \texttt{typedef struct Foo\_ \{...\} Foo;}.
+\item Avoid accessing structure members by offsets instead of by the
+      member names. As long as you keep all the code where you do this
+      up to date, it should not be a problem, but katana cannot detect
+      when you do this.
+\end{itemize}
+\subsubsection{Compilation/Linking}
+\label{sec-6.1.2}
+
     Required CFLAGS:

### 2010-06-23 doc/katana.tex
more slight license massaging

@@
-% Created 2010-06-20 Sun 23:06
+% Created 2010-06-22 Tue 19:27
 \documentclass[11pt]{article}
@@
 \author{James Oakley}
-\date{20 June 2010}
+\date{22 June 2010}
 
@@
   is Copyright 2010 Dartmouth College and may be distributed under the
-  terms of the GNU Free Documentation License
+  terms of the GNU Free Documentation License as found in the file FDL
+  which should have been distributed with this documentation. If it
+  was not, it may be found at \href{http://www.gnu.org/licenses/fdl.txt}{http://www.gnu.org/licenses/fdl.txt}.
 

### 2010-06-23 src/katana-prelim.h

@@ -4,7 +4,7 @@
   Copyright (C): 2010 Dartmouth College
   License: Katana is free software: you may redistribute it and/or
     modify it under the terms of the GNU General Public License as
-    published by the Free Software Foundation, either version 3 of the
+    published by the Free Software Foundation, either version 2 of the
     License, or (at your option) any later version. Regardless of
     which version is chose, the following stipulation also applies:
     

### 2010-06-23 src/util/dictionary.h

@@ -5,7 +5,7 @@
   Copyright (C): 2010 Dartmouth College
   License: Katana is free software: you may redistribute it and/or
     modify it under the terms of the GNU General Public License as
-    published by the Free Software Foundation, either version 3 of the
+    published by the Free Software Foundation, either version 2 of the
     License, or (at your option) any later version. Regardless of
     which version is chose, the following stipulation also applies:
     

### 2010-06-23 src/util/map.h

@@ -6,7 +6,7 @@
                  Portions 2009 James Oakley
   License: Katana is free software: you may redistribute it and/or
     modify it under the terms of the GNU General Public License as
-    published by the Free Software Foundation, either version 3 of the
+    published by the Free Software Foundation, either version 2 of the
     License, or (at your option) any later version. Regardless of
     which version is chose, the following stipulation also applies:
     

### 2010-06-23 tests/t4_new_var/v1/main.c

@@ -5,7 +5,7 @@
   Dartmouth College
   License: Katana is free software: you may redistribute it and/or
     modify it under the terms of the GNU General Public License as
-    published by the Free Software Foundation, either version 3 of the
+    published by the Free Software Foundation, either version 2 of the
     License, or (at your option) any later version. Regardless of
     which version is chose, the following stipulation also applies:
     

### 2010-06-23 tests/t7_const/v0/main.c

@@ -4,7 +4,7 @@
   Written For: Dartmouth College
   License: Katana is free software: you may redistribute it and/or
     modify it under the terms of the GNU General Public License as
-    published by the Free Software Foundation, either version 3 of the
+    published by the Free Software Foundation, either version 2 of the
     License, or (at your option) any later version. Regardless of
     which version is chose, the following stipulation also applies:
     

### 2010-11-23 doc/katana.tex
updated documentation

@@
-% Created 2010-06-22 Tue 19:27
+% Created 2010-11-14 Sun 15:47
 \documentclass[11pt]{article}
@@
 \author{James Oakley}
-\date{22 June 2010}
+\date{14 November 2010}
 
@@
 
-   \texttt{katana -g [-o OUTPUT\_FILE] /project/v0 /project/v1 foo}
+   \texttt{katana [OPTIONS] -g [-o OUTPUT\_FILE] /project/v0 /project/v1 foo}
 
@@
 
-   \texttt{katana -g [-o OUTUT\_FILE] OLD\_OBJECTS\_DIR NEW\_OBJECTS\_DIR EXECUTABLE\_NAME}
+   \texttt{katana [OPTIONS] -g [-o OUTUT\_FILE] OLD\_OBJECTS\_DIR NEW\_OBJECTS\_DIR EXECUTABLE\_NAME}
 
@@
 
-   \texttt{katana -p [-s] PATCH PID}
+   \texttt{katana [OPTIONS] -p [-s] PATCH PID}
 
@@
 
-   \texttt{katana -l PATCH}
-\subsection{See Also}
+   \texttt{katana [OPTIONS] -l PATCH}
+\subsection{Options}
 \label{sec-6.5}
 
-   the katana manpage (once it's written, which it is not yet)
+   The following options may be passed to katana regardless of whether
+   one is generating, applying, or viewing a patch:
+\begin{itemize}
+\item -c CONFIG
+     where CONFIG is the name of a configuration file to load
+\end{itemize}
+\subsection{Configuration Files}
+\label{sec-6.6}
+
+   Katana loads configuration files as follows. Configuration files
+   loaded later in the sequence may overwrite settings from files
+   earlier in the sequence.
+\begin{itemize}
+\item \emph{etc/katana    + \~{}}.katana
+\item \~{}/.config/katana
+\item ./katana
+\item any file specified with -c
+\end{itemize}
+   Configuration files are written in JSON. The JSON requirement that
+   strings be quoted is relaxed (i.e. anything is assumed to be a
+   string unless it can be interpreted otherwise). The following
+   properties are recognized:
+\begin{itemize}
+\item maxWaitForPatching <INTEGER>
+     This value specifies the maximum number of seconds to wait for
+     the target to enter a safe state.
+\item flags <OBJECT>
+     The value of flags should be an object which may contain the
+     following properties, all of which should be bool-valued:
+
+\begin{itemize}
+\item checkPtraceWrites
+       Whenever something is written into the target memory, read the
+       value back out and verify that it was written correctly. This
+       has a performance penalty, but does provide some more robust
+       error checking, although it should not be necessary.
+\end{itemize}
+
+\end{itemize}
+\subsection{See Also}
+\label{sec-6.7}
+
+   the katana manpage (although the information in this document is
+   considerably more extensive than in the manpage)
 \section{Patch Object Format}
@@
 
-  This section of the document is not yet written. It will provide a
-  description of the internal process that Katana uses to generate a
-  patch. Understanding it is not necessary for using Katana.
+  This section of the document is still under construction. When
+  complete, it will provide a description of the internal process that
+  Katana uses to generate a patch. Understanding it is not necessary
+  for using Katana.
+
+\subsection{Configuration}
+\label{sec-8.1}
+
+   Katana reads configuration files from (in order, with later
+   configuration files overriding options found in earlier ones) from
+   \texttt{/etc/katana}, \texttt{\textasciitilde{}/.katana}, \texttt{\textasciitilde{}/.config/katana}, and \texttt{./.katana}.
+
+\subsection{Initializing the patch object}
+\label{sec-8.2}
+
+   Katana sets up a patch object ELF file with the necessary sections,
+   see \hyperref[sec-7]{Patch Object Format}
+\subsection{Comparing source trees}
+\label{sec-8.3}
+
+\begin{itemize}
+\item Katana compare the old and new source trees, looking at the object (.o)
+     files.
+\item For object files which exist only in the new tree, their contents
+     are added to the patch object being created.
+\item For object files which exist only in the old tree, a warning
+     about their removal is issued and nothing further is done.
+\item For object files which exist in both trees, type diffing and
+     function diffing are performed and the differences are written
+     tot he patch object being created.
+\end{itemize}
 \section{Patch Application Process}

### 2010-11-23 src/util/stack.c
wrote stack data type

@@ -1,0 +1,63 @@
+/*
+  File: stack.c
+  Author: James Oakley
+  Copyright (C): 2010 James Oakley
+  License: Katana is free software: you may redistribute it and/or
+     modify it under the terms of the GNU General Public License as
+     published by the Free Software Foundation, either version 2 of the
+     License, or (at your option) any later version.
+
+     This file was not written while under employment by Dartmouth College
+  Project:  katana
+  Date: October 2010
+  Description: abstract stack data type
+*/
+
+#include "stack.h"
+
+typedef struct SNode
+{
+  struct SNode* next;
+  void* data;
+} SNode;
+
+
+Stack* stackCreate()
+{
+  return zmalloc(sizeof(Stack));
+
+}
+
+void stackDelete(Stack* st,void (*deleteData)(void*))
+{
+  if(deleteData)
+  {
+    for(SNode* node=st->top;node;node=node->next)
+    {
+      (*deleteData)(node->data);
+    }
+  }
+  free(st);
+}
+
+void stackPush(Stack* st,void* item)
+{
+  SNode* node=malloc(sizeof(SNode));
+  MALLOC_CHECK(node);
+  node->next=st->top;
+  node->data=item;
+  st->top=node;
+}
+
+//return NULL if there is no item at the top of the stack
+void* stackPop(Stack* st)
+{
+  SNode* result=st->top;
+  if(result)
+  {
+    st->top=result->next;
+  }
+  return result->data;;
+}
+
+

### 2010-11-23 src/util/stack.h

@@ -1,0 +1,36 @@
+/*
+  File: stack.h
+  Author: James Oakley
+  Copyright (C): 2010 James Oakley
+  License: Katana is free software: you may redistribute it and/or
+     modify it under the terms of the GNU General Public License as
+     published by the Free Software Foundation, either version 2 of the
+     License, or (at your option) any later version.
+
+     This file was not written while under employment by Dartmouth College
+  Project:  katana
+  Date: October 2010
+  Description: abstract stack data type
+*/
+
+#ifndef stack_h
+#define stack_h
+
+#include "util.h"
+
+typedef struct _Stack
+{
+  struct SNode* top;
+} Stack;
+
+
+Stack* stackCreate();
+
+void stackDelete(Stack* st,void (*deleteData)(void*));
+
+void stackPush(Stack* st,void* item);
+
+//return NULL if there is no item at the top of the stack
+void* stackPop(Stack* st);
+
+#endif

### 2010-11-23 src/arch.h
-better support for printing fde/call frame info -support for printing out DWARF information other than that used for patching (this has no application for patching, only in using Katana as a general DWARF tool) -support for some more DWARF instructions

@@ -70,6 +70,12 @@
 #define REG_8(regs_struct) (regs_struct).r8
 #define REG_9(regs_struct) (regs_struct).r9
 #define REG_10(regs_struct) (regs_struct).r10
+#define REG_11(regs_struct) (regs_struct).r10
+#define REG_12(regs_struct) (regs_struct).r10
+#define REG_13(regs_struct) (regs_struct).r10
+#define REG_14(regs_struct) (regs_struct).r10
+#define REG_15(regs_struct) (regs_struct).r10
+#define NUM_REGS 15
 #define ElfXX_Sym Elf64_Sym
 #define ElfXX_Rel Elf64_Rel
 #define ElfXX_Rela Elf64_Rela
@@ -98,6 +104,7 @@
 #define REG_BP(regs_struct) (regs_struct).ebp
 #define REG_SI(regs_struct) (regs_struct).esi
 #define REG_DI(regs_struct) (regs_struct).edi
+#define NUM_REGS 8
 #define ElfXX_Sym Elf32_Sym
 #define ElfXX_Rel Elf32_Rel
 #define ElfXX_Rela Elf32_Rela

### 2010-11-23 src/dwarf_instr.c

@@ -178,7 +178,7 @@ void printInstruction(RegInstruction inst)
     {
       signed long int tmp;
       memcpy(&tmp,&inst.arg2,sizeof(inst.arg2));
-      printf("DW_CFA_offset_ r%i %zi\n",inst.arg1,tmp);
+      printf("DW_CFA_offset r%i %zi\n",inst.arg1,tmp);
     }
     break;
   case DW_CFA_register:
@@ -259,10 +259,16 @@ void printInstruction(RegInstruction inst)
   case DW_CFA_def_cfa_offset:
     printf("DW_CFA_def_cfa_offset %i\n",inst.arg1);
     break;
+  case DW_CFA_remember_state:
+    printf("DW_CFA_remember_state\n");
+    break;
+  case DW_CFA_restore_state:
+    printf("DW_CFA_restore_state\n");
+    break;
   case DW_CFA_nop:
     printf("DW_CFA_nop\n");
     break;
   default:
-    death("unsupported instruction");
+    death("unsupported DWARF instruction 0x%x",inst.type);
   }
 }

### 2010-11-23 src/dwarfvm.c

@@ -64,12 +64,19 @@
 #include "symbol.h"
 #include "util/map.h"
 #include "patcher/hotpatch.h"
+#include "util/stack.h"
 
 //returns a list of PatchData objects
 List* generatePatchesFromFDEAndState(FDE* fde,SpecialRegsState* state,ElfInfo* patch,ElfInfo* patchedBin);
 
 //keys are old addresses of data objects (variables). Values are the new addresses
 Map* dataMoved=NULL;
+
+//this is the stack of saved register states used by the
+//DW_CFA_remember_state and DW_CFA_restore_state instructions
+static Stack* stateStack;
+//unfortunately because the way we currently evaluate
+//static Map* rememberedStatesByLocation;
 
 //evaluates the given instructions and stores them in the output rules
 //dictionary.The initial condition of regarray IS taken into account.
@@ -79,39 +86,92 @@ Map* dataMoved=NULL;
 //considered to start at 0) if stopLocation is negative, it is ignored
 //if stopLocation is negative, it is ignored (evaluation continues to
 //the end of the instructions) returns the location stopped at (will
-//be the lowest location that a change was actually made)
-int evaluateInstructionsToRules(RegInstruction* instrs,int numInstrs,Dictionary* rules,int startLocation, int stopLocation)
+//be the lowest location that a change was actually made).
+//outInstrsCnt, if non-NULL, is used to store the number of instructions read
+int evaluateInstructionsToRules(RegInstruction* instrs,int numInstrs,Dictionary* rules,int startLocation, int stopLocation,int* outInstrsCnt)
 {
   PoRegRule* rule=NULL;
   int loc=startLocation;
   for(int i=0;i<numInstrs;i++)
   {
     RegInstruction inst=instrs[i];
-    if(inst.type==DW_CFA_set_loc)
+
+    //deal first with the location-advancing instructions and other
+    //instructions (remember/restore state) that don't do the ordinary
+    //register manipulation
+    switch(inst.type)
     {
+    case DW_CFA_set_loc:
       if(stopLocation >= 0 && inst.arg1>stopLocation)
       {
         return loc;
       }
       loc=inst.arg1;
       continue;
-    }
-    else if(inst.type==DW_CFA_advance_loc ||
-       inst.type==DW_CFA_advance_loc1 ||
-       inst.type==DW_CFA_advance_loc2)
-    {
+    case DW_CFA_advance_loc:
+    case DW_CFA_advance_loc1:
+    case DW_CFA_advance_loc2:
[…]

### 2010-11-23 src/dwarfvm.h

@@ -63,7 +63,7 @@ void patchDataWithFDE(VarInfo* var,FDE* transformerFDE,ElfInfo* targetBin,ElfInf
 //execution continues until the end of the instructions or until the location is advanced
 //past stopLocation. stopLocation should be relative to the start of the instructions (i.e. the instructions are considered to start at 0)
 //if stopLocation is negative, it is ignored
-int evaluateInstructionsToRules(RegInstruction* instrs,int numInstrs,Dictionary* rules,int startLocation, int stopLocation);
+int evaluateInstructionsToRules(RegInstruction* instrs,int numInstrs,Dictionary* rules,int startLocation, int stopLocation,int* outInstrsCnt);
 
 //stack length given in words
 word_t evaluateDwarfExpression(byte* bytes,int len,word_t* startingStack,int stackLen);

### 2010-11-23 src/fderead.c

@@ -193,6 +193,10 @@ RegInstruction* parseFDEInstructions(Dwarf_Debug dbg,unsigned char* bytes,int le
         bytes+=uleblen;
         len-=uleblen;
         break;
+      case DW_CFA_remember_state:
+      case DW_CFA_restore_state:
+        //don't need to store any arguments, this instruction doesn't take any
+        break;
       case DW_CFA_nop:
         (*numInstrs)--;//since not actually using up an instruction here
         break;
@@ -288,8 +292,9 @@ Map* readDebugFrame(ElfInfo* elf,bool ehInsteadOfDebug)
                            &elf->cies[i].numInitialInstructions);
     elf->cies[i].initialRules=dictCreate(100);//todo: get rid of
     //arbitrary constant 100
-    evaluateInstructionsToRules(elf->cies[i].initialInstructions,elf->cies[i].numInitialInstructions,
-                                elf->cies[i].initialRules,0,-1);
+    evaluateInstructionsToRules(elf->cies[i].initialInstructions,
+                                elf->cies[i].numInitialInstructions,
+                                elf->cies[i].initialRules,0,-1,NULL);
   
     //todo: bizarre bug, it keeps coming out as -1, which is wrong
     elf->cies[i].codeAlign=1;
@@ -314,7 +319,7 @@ Map* readDebugFrame(ElfInfo* elf,bool ehInsteadOfDebug)
     dwarf_get_cie_index(dcie,&cieIndex,&err);
     elf->fdes[i].cie=&elf->cies[cieIndex];
     CIE* cie=elf->fdes[i].cie;
-    
+    logprintf(ELL_INFO_V2,ELS_DWARF_FRAME,"Reading instructions in FDE #%i\n",i);
     elf->fdes[i].instructions=parseFDEInstructions(dbg,instrs,ilen,cie->dataAlign,cie->codeAlign,&elf->fdes[i].numInstructions);
     Dwarf_Addr lowPC = 0;
     Dwarf_Unsigned addrRange = 0;

### 2010-11-23 src/fderead.h

@@ -70,7 +70,10 @@ typedef struct CIE
 {
   RegInstruction* initialInstructions;
   int numInitialInstructions;
-  Dictionary* initialRules;
+  Dictionary* initialRules; //dictionary mapping the stringified
+                            //version of a register to the rule (of
+                            //type PoRegRule) for setting that
+                            //register
   Dwarf_Signed dataAlign;
   Dwarf_Unsigned codeAlign;
   Dwarf_Half returnAddrRuleNum;

### 2010-11-23 src/info/fdedump.c

@@ -82,7 +82,7 @@ void printCIEInfo(CIE* cie)
 
 void printFDEInfo(CIE* cie,FDE* fde,int num)
 {
-  printf("--------FDE #%i-----\n",num);
+  printf("--------FDE #%i (at offset 0x%x----\n",num,fde->offset);
   //printf("\tfunction guess:%s\n",get_fde_proc_name(dbg,fde->lowpc));
   printf("\tlowpc: 0x%x\n",fde->lowpc);
   printf("\thighpc:0x%x\n",fde->highpc);
@@ -93,20 +93,26 @@ void printFDEInfo(CIE* cie,FDE* fde,int num)
     printInstruction(fde->instructions[i]);
   }
   printf("    The table would be as follows\n");
+  Dictionary* rulesDict=dictDuplicate(cie->initialRules,(DictDataCopy)duplicatePoRegRule);
+  //use this to keep track of which instructions we've read so far so
+  //we don't read the same ones over and over
+  int numInstrsReadSoFar=0;
+
   //todo: figure out what we're actually using highpc for.
   //at one point I had highpc replaced by 1 in the below check but I'm
   //not sure what the purpose of that was
   for(int i=fde->lowpc;i<fde->highpc || 0==i;i++)
   {
-    Dictionary* rulesDict=dictDuplicate(cie->initialRules,NULL);
-    int stopLocation=evaluateInstructionsToRules(fde->instructions,fde->numInstructions,rulesDict,fde->lowpc,i);
-    if(stopLocation != i)
-    {
+    int instrsRead=0;
+    int stopLocation=evaluateInstructionsToRules(fde->instructions+numInstrsReadSoFar,fde->numInstructions,rulesDict,fde->lowpc,i,&instrsRead);
+    numInstrsReadSoFar+=instrsRead;
+    if(stopLocation != i) 
+    { 
+      //dictDelete(rulesDict,NULL);
       continue;//Don't need to print this because will be dup
     }
-    printf("    ----Register Rules at text address 0x%x(%x)------\n",i,stopLocation);
+    printf("    ----Register Rules at text address 0x%x------\n",i);
     printRules(rulesDict,"      ");
-    dictDelete(rulesDict,NULL);
   }
 }
 

### 2010-11-23 src/register.c

@@ -108,6 +108,73 @@ PoReg readRegFromLEB128(byte* leb,usint* bytesRead)
   return result;
 }
 
+
+char* getX86RegNameFromDwarfRegNum(int num)
+{
+  switch(num)
+  {
+  case 0:
+    return "eax";
+  case 1:
+    return "ecx";
+  case 2:
+    return "edx";
+  case 3:
+    return "ebx";
+  case 4:
+    return "esp";
+  case 5:
+    return "ebp";
+  case 6:
+    return "esi";
+  case 7:
+    return "edi";
+  case 8:
+    return "eip";
+  default:
+    fprintf(stderr,"unknown register number %i, cannot get reg name\n",num);
+    death(NULL);
+  }
+  return "error";
+}
+
+char* getX86_64RegNameFromDwarfRegNum(int num)
+{
+  switch(num)
+  {
+  case 0:
+    return "rax";
+  case 1:
+    return "rdx";
+  case 2:
+    return "rcx";
+  case 3:
+    return "rbx";
+  case 4:
+    return "rsi";
+  case 5:
+    return "rdi";
+  case 6:
+    return "rbp";
+  case 7:
+    return "rsp";
+  default:
+    return NULL;
+  }
+}
+
+char* getArchRegNameFromDwarfRegNum(int num)
+{
+  #ifdef KATANA_X86_ARCH
+  return getX86RegNameFromDwarfRegNum(num);
+  #elif defined(KATANA_X86_64_ARCH)
+  return getX86_64RegNameFromDwarfRegNum(num);
+  #else
+  #error "Unsupported architecture"
+  #endif
+}
+
+
 //the returned string should be freed
 char* strForReg(PoReg reg)
 {
@@ -137,7 +204,17 @@ char* strForReg(PoReg reg)
     snprintf(buf,128,"{CFA}");
     break;
   case ERT_BASIC:
-    snprintf(buf,128,"r%i",reg.u.index);
+    {
+      char* regName=getArchRegNameFromDwarfRegNum(reg.u.index);
+      if(regName)
+      {
+        snprintf(buf,128,"r%i/%s",reg.u.index,regName);
+      }
+      else
+      {
+        snprintf(buf,128,"r%i",reg.u.index);
+      }
+    }
     break;
   default:
     death("unsupported register type\n");
@@ -235,13 +312,27 @@ void printRule(PoRegRule rule,int regnum)
     printf("%s = Undefined\n",regStr);
[…]

### 2010-11-23 src/register.h

@@ -92,6 +92,9 @@ typedef struct
   } u;
 } PoReg;
 
+PoReg* copyPoReg(PoReg* old);
+
+
 PoReg readRegFromLEB128(byte* leb,usint* bytesRead);
 
 //the returned string should be freed
@@ -148,4 +151,9 @@ typedef enum
 //flags determines this behaviour (should be OR'd E_REG_RESOLVE_FLAGS
 int resolveRegisterValue(PoReg* reg,SpecialRegsState* state,byte** result,int flags);
 
+char* getArchRegNameFromDwarfRegNum(int num);
+
+PoRegRule* duplicatePoRegRule(PoRegRule* rule);
+
+PoReg* duplicatePoReg(PoReg* old);
 #endif

### 2010-11-23 src/util/dictionary.c

@@ -108,6 +108,7 @@ DNODE* dictNodeDuplicate(DNODE* n,void* (*dataCP)(void*))
   return new;
 }
 
+//If dataCP is NULL then values will simply be copied over as pointers
 Dictionary* dictDuplicate(Dictionary* dict,void* (*dataCP)(void*))
 {
   Dictionary* new=dictCreate(dict->numBuckets);

### 2010-11-23 src/util/dictionary.h

@@ -87,10 +87,12 @@ typedef struct _DICTIONARY {
 
 typedef struct _DICTIONARY DICTIONARY;
 typedef struct _DICTIONARY Dictionary;
+typedef void* (*DictDataCopy)(void*);
 
 //! create an empty dictionary
 Dictionary* dictCreate(int nBuckets);
 
+//If dataCP is NULL then values will simply be copied over as pointers
 Dictionary* dictDuplicate(Dictionary* dict,void* (*dataCP)(void*));
 
 //! clean up a dictionary that's no longer needed

### 2010-11-23 src/util/logging.c

@@ -1,5 +1,5 @@
 /*
-  File: logging.h
+  File: logging.c
   Author: James Oakley
   Copyright (C): 2010 Dartmouth College
   License: Katana is free software: you may redistribute it and/or
@@ -120,7 +120,7 @@ void loggingDefaults()
   sourceEnables[ELS_MISC]=ELL_WARN;
   sourceEnables[ELS_CODEDIFF]=ELL_WARN;
   sourceEnables[ELS_TYPEDIFF]=ELL_WARN;
-  sourceEnables[ELS_DWARF_FRAME]=ELL_WARN;
+  sourceEnables[ELS_DWARF_FRAME]=ELL_INFO_V2;
   sourceEnables[ELS_HOTPATCH]=ELL_INFO_V2;
   sourceEnables[ELS_SOURCETREE]=ELL_WARN;
   sourceEnables[ELS_SYMBOL]=ELL_WARN;

### 2010-11-23 src/patcher/patchapply.c
fixed up some comments and extended support for converting between dwarf registers and machine registers

@@ -841,8 +841,6 @@ void readAndApplyPatch(int pid,ElfInfo* targetBin_,ElfInfo* patch,int flags)
   getShdrByERS(targetBin,ERS_GOTPLT,&shdr);
   amount+=shdr.sh_size;
 
-
-
   #ifdef KATANA_X86_64_ARCH
   addr_t desiredAddress=0;
 
@@ -859,9 +857,6 @@ void readAndApplyPatch(int pid,ElfInfo* targetBin_,ElfInfo* patch,int flags)
     desiredAddress-=misalignment;
   }
   #endif
-
-  
-
 
   #ifdef KATANA_X86_64_ARCH
   addr_t receivedAddres=reserveFreeSpaceInTarget(amount,desiredAddress);

### 2010-11-23 src/patcher/safety.c

@@ -110,9 +110,13 @@ char* getFunctionNameAtPC(ElfInfo* elf,addr_t pc)
   return getString(elf,sym.st_name);
 }
 
+//this function and the one below it are coded badly. This can be done
+//more efficiently
+
 //conversion between dwarf register numbers and x86 registers
 void setRegValueFromDwarfRegNum(struct user_regs_struct* regs,int num,long int newValue)
 {
+  #ifdef KATANA_X86_ARCH
   switch(num)
   {
   case 0:
@@ -143,14 +147,75 @@ void setRegValueFromDwarfRegNum(struct user_regs_struct* regs,int num,long int n
     REG_IP(*regs)=newValue;
     break;
   default:
+    //If this is hit, probably more need to be implemented
     fprintf(stderr,"unknown register number %i, cannot set reg val\n",num);
     death(NULL);
   }
+  #elif defined(KATANA_X86_64_ARCH)
+  switch(num)
+  {
+  case 0:
+    REG_AX(*regs)=newValue;
+    break;
+  case 1:
+    REG_DX(*regs)=newValue;
+    break;
+  case 2:
+    REG_CX(*regs)=newValue;
+    break;
+  case 3:
+    REG_BX(*regs)=newValue;
+    break;
+  case 4:
+    REG_SI(*regs)=newValue;
+    break;
+  case 5:
+    REG_DI(*regs)=newValue;
+    break;
+  case 6:
+    REG_BP(*regs)=newValue;
+    break;
+  case 7:
+    REG_SP(*regs)=newValue;
+    break;
+  case 8:
+    REG_8(*regs)=newValue;
+    break;
+  case 9:
+    REG_9(*regs)=newValue;
+    break;
+  case 10:
+    REG_10(*regs)=newValue;
+    break;
+  case 11:
+    REG_10(*regs)=newValue;
+    break;
+  case 12:
+    REG_10(*regs)=newValue;
+    break;
+  case 13:
+    REG_10(*regs)=newValue;
+    break;
+  case 14:
+    REG_10(*regs)=newValue;
+    break;
+  case 15:
+    REG_10(*regs)=newValue;
+    break;;
+  default:
+    //If this is hit, probably more need to be implemented
+    fprintf(stderr,"unknown register number %i, cannot set reg val\n",num);
+    death(NULL);
+  }
+  #else
+  #error "Unsupported architecture"
+  #endif
 }
 
 //conversion between dwarf register numbers and x86 registers
[…]

### 2010-11-23 src/patchwrite/codediff.c

@@ -69,6 +69,8 @@ int cmpRelocs(void* a,void* b)
   return relocA->r_offset-relocB->r_offset;
 }
 
+//compare program text modulo relocations which refer to the same
+//symbol, symbol of changed type, or changed offset on symbol
 bool areSubprogramsIdentical(SubprogramInfo* patcheeFunc,SubprogramInfo* patchedFunc,
                              ElfInfo* oldBinary,ElfInfo* newBinary)
 {

### 2011-04-13 doc/katana.tex
updated documentation with info about the shell

@@
-% Created 2010-11-14 Sun 15:47
+% Created 2011-04-13 Wed 00:01
 \documentclass[11pt]{article}
@@
 
-\title{Katana: A Userland Toolchain-Oriented Hotpatching System}
+\title{Katana: An ELF/DWARF Manipulation Tool with Hotpatching Capabilities}
 \author{James Oakley}
-\date{14 November 2010}
+\date{13 April 2011}
 
@@
 
+  Katana is a research system for ELF/DWARF manipulation. It was
+  originally developed for research into hotpatching. It was later
+  revised for research into security implication of gcc/C++ exception
+  handling, which is implemented primarily using DWARF call frame
+  information. Therefore, if you are interested in vulnerabilities
+  related to exceptiong handling/DWARF you may probably ignore the
+  parts of this manual which discuss hotpatching. If you are instead
+  interested in hotpatching, you may probably ignore the parts of this
+  manual that deal with manipulating exception handling structures.
+
   Katana aims to provide a hot-patching system for userland. Further
@@
   with anything besides programs written in C). A diagram of software
-  lifecycle with hotpatching is shown below
+  lifecycle with hotpatching is shown below (unless you are reading this in plain text)
 
@@
   this document necessary for ``Users' Guide'' purposes are 
-  \hyperref[sec-3]{``What Katana Does''}, \hyperref[sec-4]{``What Katana Does Not Do (Yet)''}, and most importantly 
-  \hyperref[sec-6]{``How to Use Katana''}.
+  \hyperref[sec-3.2]{``What Katana Does''}, \hyperref[sec-3.3]{``What Katana Does Not Do (Yet)''}, and most importantly 
+  \hyperref[sec-3.5]{``How to Use Katana For Hotpatching''}.
 
   This document is a work in progress. It is not a polished guide yet.
-\section{Other Systems}
+
+\section{General Usage Information}
 \label{sec-2}
+
+\subsection{Shell}
+\label{sec-2.1}
+
+   If Katana is not passed an argument indicating one of the
+   hot-patching commands (described later in  \hyperref[sec-3.5]{*How to Use Katana For Hotpatching}), then it is assumed to be operating as a shell. If it
+   is provided an argument, that argument is taken as the name of a
+   file to read shell commands from. Otherwise commands are read from
+   stdin using the readline library. 
+\subsubsection{Syntax and Data Model}
+\label{sec-2.1.1}
+
+    The Katana shell syntax is very simple. There are no control flow
+    structures, only commands and variables. A line is terminated by a
+    semicolon (;) or a newline character. Each line may be either
+    blank, contain exactly one COMMAND, or contain an ASSIGNMENT.
+
+    A COMMAND is of the form COMMAND$_{\mathrm{IDENTIFIER}}$ PARAM PARAM PARAM \ldots{}., where
+    tokens are seperated by spaces and the number of PARAMs depends on
+    the command.
+
+    An ASSIGNMENT is currently of the form VARIABLE=COMMAND although
+    in the future it may be possible to write other sorts of
+    assignments.
+
+    A VARIABLE reference consists of a dollar-sign (\$) followed by a
+    letter or underscore followed by any number of letters,
+    underscores, or digits.
+
+    A COMMAND$_{\mathrm{IDENTIFIER}}$ is one or more words which identify a
+    COMMAND. In many cases a command is identified by only one word,
+    but sometimes similar commands are grouped by sharing the first
+    word in their identifier.
+
+    A PARAM is a VARIABLE reference, STRING, or NUMBER
+
+    A STRING is any literal beginning and ending with the character ``.
+
+    A NUMBER is a decimal, hex, or float literal.
+
+\begin{itemize}
+
+\item Data Types\\
+\label{sec-2.1.1.1}
+
+     The following types of variables exist
+\begin{itemize}
+\item string
+\item elf
+\item elf section
+\item raw data
+\item array
+\end{itemize}
+\end{itemize} % ends low level
+\subsubsection{Available Commands}
+\label{sec-2.1.2}
+
+\begin{itemize}
+
+\item load\\
+\label{sec-2.1.2.1}
+
+     \emph{Usage}: \texttt{load FILENAME}\\
+     \emph{Params}: FILENAME must a string literal or variable that can be interpreted
+               as a string.\\
+     \emph{Function}: Loads the data in the given file as an ELF object if
+               possible. If not, loads it as raw data.
+
+\item save\\
+\label{sec-2.1.2.2}
+
+     \emph{Usage}: \texttt{save VAR FILENAME}\\
+     \emph{Params}: VAR must be a variable that can be interpreted as an ELF
+               object or that can be interpreted as raw data. FILENAME must be a
+               literal or variable that can be interpreted as a string.\\
+     \emph{Function}: Saves VAR to FILENAME.
+
+\item dwarfscript\\
+\label{sec-2.1.2.3}
+
+\begin{itemize}
+
+\item dwarfscript emit\\
+\label{sec-2.1.2.3.1}
+
+      \emph{Usage}: \texttt{dwarfscript emit [SECTION] ELF OUTFILE}\\
+      \emph{Params}: SECTION must be the name (string) of the section to write as
+                Dwarfscript. If not specified it defaults to
+                ``.eh$_{\mathrm{frame}}$''. ELF must be an ELF object. OUTFILE must
+                be a string with the name of a file to write the resulting
+                Dwarfscript to.\\
+      \emph{Function}: Writes the Dwarfscript representation of the given
+                  SECTION from the given ELF to OUTFILE.
+
+\item dwarfscript compile\\
+\label{sec-2.1.2.3.2}
+
+      \emph{Usage}: \texttt{dwarfscript compile INFILE}\\
+      \emph{Params}: INFILE must be a string containing the name of a file.\\
+      \emph{Function}: Interprets the contents of the file named by INFILE
+                  as Dwarfscript and compiles the Dwarfscript into
+                  beinary form. Returns an array with 3 items
+                  0: raw data for .eh$_{\mathrm{frame}}$ 
+                  1: raw data for .eh$_{\mathrm{frame}}$$_{\mathrm{hdr}}$
+                  2: raw data for .gcc$_{\mathrm{except}}$$_{\mathrm{table}}$.
+\end{itemize} % ends low level
+
+\item replace\\
+\label{sec-2.1.2.4}
+
+\begin{itemize}
+
+\item replace section\\
+\label{sec-2.1.2.4.1}
+
+      \emph{Usage}: \texttt{replace section ELF SECTION\_NAME NEW\_SECTION}
+      \emph{Params}: ELF must be an ELF object. SECTION$_{\mathrm{NAME}}$ must be a
+                string. NEW$_{\mathrm{SECTION}}$ must be either an ELF section or raw data.
+      \emph{Function}: Replaces the section with the name SECTION$_{\mathrm{NAME}}$ in
+                  the oject ELF with the data from
+                  NEW$_{\mathrm{SECTION}}$. Section headers are replaced if NEW$_{\mathrm{SECTION}}$ is
+                  able to provide them, but not if it is only raw data.
+
+\item replace raw\\
+\label{sec-2.1.2.4.2}
+
+      \emph{Usage}: \texttt{replace raw ELF OFFSET NEW\_DATA}
+      \emph{Params}: ELF must be an ELF object. ADDRESS must be an
+                integer. NEW$_{\mathrm{DATA}}$ must be raw data.
+      \emph{Function}: Replaces the raw data at OFFSET in the ELF object
+                  with NEW$_{\mathrm{DATA}}$. OFFSET must refer to a location in an
+                  existing section.
+\end{itemize} % ends low level
+
+\item info\\
+\label{sec-2.1.2.5}
+
+\begin{itemize}
+
+\item info eh\\
+\label{sec-2.1.2.5.1}
+
+      \emph{Usage}: \texttt{info eh ELF [OUTFILE]}
+      \emph{Params}: ELF must be an ELF object. OUTFILE, if present, must
+                be the name of a writable file (which may or may not
+                exist yet). 
+      \emph{Function}: Prints out information about the exception-handling
+                  structures in ELF. If OUTFILE is present, this
+                  information is written to it.
+\end{itemize} % ends low level
+
+\item hash\\
+\label{sec-2.1.2.6}
+
+\begin{itemize}
+
+\item hash elf\\
+\label{sec-2.1.2.6.1}
+
+      \emph{Usage}: \texttt{hash elf STR}
+      \emph{Params}: STR must be a string.
+      \emph{Function}: Prints the result of running elf$_{\mathrm{hash}}$ (from libelf)
+                  on the string.
+
+\end{itemize} % ends low level
+
+\item patch\\
+\label{sec-2.1.2.7}
+
+\begin{itemize}
+
+\item gen\\
+\label{sec-2.1.2.7.1}
+
+      \emph{Usage}: \texttt{patch gen OLD\_OBJECTS\_DIR NEW\_OBJECTS\_DIR EXECUTABLE}
+      \emph{Params}: All three params are strings. The first two are the
+                old and new object file directories respectively. The
+                last is the name of the executable that can be found
+                in both directories.
+      \emph{Function}: Generates (and returns) a patch object ELF.
+
+\item apply\\
+\label{sec-2.1.2.7.2}
+
+      \emph{Usage}: \texttt{patch apply PO PID}
+      \emph{Params}: The PO parameter should be an ELF patch object. PID
+                should be the (integer) pid of the process that PO is
+                to be applied to.
+      \emph{Function}: Applies the patch object PO to the running process
+                  described by PID.
+\end{itemize} % ends low level
+
+\item ! (shell command)\\
+\label{sec-2.1.2.8}
+
+     The rest of the line following by ! is executed in a shell.
+\end{itemize} % ends low level
+\subsubsection{History}
+\label{sec-2.1.3}
+
+    Command history is saved using libreadline in \texttt{\$HOME/.katana\_history}.
+\section{Hotpatching}
+\label{sec-3}
+
+\subsection{Other Systems}
+\label{sec-3.1}
 
@@
    general ELF-based patch format.
-\section{What Katana Does}
-\label{sec-3}
+\subsection{What Katana Does}
+\label{sec-3.2}
 
@@
 \end{itemize}
-\section{What Katana Does Not Do (Yet)}
-\label{sec-4}
+\subsection{What Katana Does Not Do (Yet)}
+\label{sec-3.3}
 
@@
 \end{itemize}
-  See \hyperref[sec-10]{Roadmap} for more things which are not complete
+   See \hyperref[sec-3.14]{Roadmap} for more things which are not complete
 
-\section{What Katana May Never Do}
-\label{sec-5}
+\subsection{What Katana May Never Do}
+\label{sec-3.4}
 
@@
 \end{itemize}
-\section{How to Use Katana}
-\label{sec-6}
+\subsection{How to Use Katana For Hotpatching}
+\label{sec-3.5}
 
@@
    makes, etc.
-\subsection{Preparing a Package for Patching Support}
-\label{sec-6.1}
+\subsubsection{Preparing a Package for Patching Support}
+\label{sec-3.5.1}
 
@@
 \subsubsection{Source Code Practices}
-\label{sec-6.1.1}
+\label{sec-3.5.2}
 
@@
 \subsubsection{Compilation/Linking}
-\label{sec-6.1.2}
+\label{sec-3.5.3}
 
@@
 \end{itemize}
-\subsection{To Generate a Patch}
-\label{sec-6.2}
+\subsubsection{To Generate a Patch}
+\label{sec-3.5.4}
 
@@
     If \texttt{-o OUTPUT\_FILE} is not specified, the output file will be \texttt{OLD\_OBJECTS\_DIR/EXECUTABLE\_NAME.po}
-\subsection{To Apply a Patch}
-\label{sec-6.3}
+\subsubsection{To Apply a Patch}
+\label{sec-3.5.5}
 
@@
     debugging Katana.
-\subsection{To View a Patch}
-\label{sec-6.4}
+\subsubsection{To View a Patch}
+\label{sec-3.5.6}
 
@@
     \texttt{katana [OPTIONS] -l PATCH}
-\subsection{Options}
-\label{sec-6.5}
+\subsubsection{Options}
+\label{sec-3.5.7}
 
@@
 \end{itemize}
-\subsection{Configuration Files}
-\label{sec-6.6}
+\subsubsection{Configuration Files}
+\label{sec-3.5.8}
 
@@
 \end{itemize}
-\subsection{See Also}
-\label{sec-6.7}
+\subsubsection{See Also}
+\label{sec-3.5.9}
 
@@
     considerably more extensive than in the manpage)
-\section{Patch Object Format}
-\label{sec-7}
+\subsection{Patch Object Format}
+\label{sec-3.6}
 
    This section of the document is not yet written. It will provide a description and specification of the PO format used by Katana
-\section{Patch Generation Process}
-\label{sec-8}
+\subsection{Patch Generation Process}
+\label{sec-3.7}
 
@@
 \subsection{Configuration}
-\label{sec-8.1}
+\label{sec-3.8}
 
@@
 \subsection{Initializing the patch object}
-\label{sec-8.2}
+\label{sec-3.9}
 
    Katana sets up a patch object ELF file with the necessary sections,
-   see \hyperref[sec-7]{Patch Object Format}
+   see \hyperref[sec-3.6]{Patch Object Format}
 \subsection{Comparing source trees}
-\label{sec-8.3}
+\label{sec-3.10}
 
@@
 \end{itemize}
-\section{Patch Application Process}
-\label{sec-9}
+\subsection{Type Diffing}
+\label{sec-3.11}
+
+   This section of the document still needs to be written. The general
+   idea is that structures are examined for for added members, moved
+   members, and changed members.
+\subsection{Function Diffing}
+\label{sec-3.12}
+
+\subsection{Patch Application Process}
+\label{sec-3.13}
 
@@
    patch. Understanding it is not necessary for using Katana.
-\section{Roadmap}
-\label{sec-10}
+\subsection{Roadmap}
+\label{sec-3.14}
 
@@
 \end{itemize}
+\section{DWARF Manipulation}
+\label{sec-4}
+
 \section{Credits and Licensing}
-\label{sec-11}
+\label{sec-5}
 
@@
   specified in the file COPYING distributed with Katana. This document
-  is Copyright 2010 Dartmouth College and may be distributed under the
-  terms of the GNU Free Documentation License as found in the file FDL
-  which should have been distributed with this documentation. If it
-  was not, it may be found at \href{http://www.gnu.org/licenses/fdl.txt}{http://www.gnu.org/licenses/fdl.txt}.
+  is Copyright 2010-2011 Dartmouth College and may be distributed
+  under the terms of the GNU Free Documentation License as found in
+  the file FDL which should have been distributed with this
+  documentation. If it was not, it may be found at
+  \href{http://www.gnu.org/licenses/fdl.txt}{http://www.gnu.org/licenses/fdl.txt}.
 
-  Katana is being written by James Oakley and was designed
-  by Sergey Bratus, Ashwin Ramaswamy, James Oakley, Michael Locasto,
-  and Sean Smith.
+  Katana is being written by James Oakley and was designed by Sergey
+  Bratus, Ashwin Ramaswamy, James Oakley, Michael Locasto, and Sean
+  Smith.
 

### 2011-04-18 src/patcher/patchapply.c
made a comment clearer

@@ -591,6 +591,7 @@ void fixupPatchRelocations(ElfInfo* patch)
   }
 }
 
+//copy PLT and GOT to new locations, as we may need to expand them
 ///todo: this function may not work for x86-64 Large Code Model
 void katanaPLT()
 {

### 2011-04-18 doc/katana.tex
updated documentation for 0.2 release

@@
-% Created 2011-04-13 Wed 00:01
+% Created 2011-04-18 Mon 01:44
 \documentclass[11pt]{article}
@@
 \author{James Oakley}
-\date{13 April 2011}
+\date{18 April 2011}
 
@@
 
-    A COMMAND is of the form COMMAND$_{\mathrm{IDENTIFIER}}$ PARAM PARAM PARAM \ldots{}., where
+    A COMMAND is of the form COMMAND\_{}IDENTIFIER PARAM PARAM PARAM \ldots{}., where
     tokens are seperated by spaces and the number of PARAMs depends on
@@
 
-    A COMMAND$_{\mathrm{IDENTIFIER}}$ is one or more words which identify a
+    A COMMAND\_{}IDENTIFIER is one or more words which identify a
     COMMAND. In many cases a command is identified by only one word,
@@
 \item string
-\item elf
-\item elf section
+\item ELF
+\item ELF section
 \item raw data
@@
                 Dwarfscript. If not specified it defaults to
-                ``.eh$_{\mathrm{frame}}$''. ELF must be an ELF object. OUTFILE must
+                ``.eh\_{}frame''. ELF must be an ELF object. OUTFILE must
                 be a string with the name of a file to write the resulting
@@
                   beinary form. Returns an array with 3 items
-                  0: raw data for .eh$_{\mathrm{frame}}$ 
-                  1: raw data for .eh$_{\mathrm{frame}}$$_{\mathrm{hdr}}$
-                  2: raw data for .gcc$_{\mathrm{except}}$$_{\mathrm{table}}$.
+                  0: raw data for .eh\_{}frame 
+                  1: raw data for .eh\_{}frame\_{}hdr
+                  2: raw data for .gcc\_{}except\_{}table.
 \end{itemize} % ends low level
@@
       \emph{Usage}: \texttt{replace section ELF SECTION\_NAME NEW\_SECTION}
-      \emph{Params}: ELF must be an ELF object. SECTION$_{\mathrm{NAME}}$ must be a
-                string. NEW$_{\mathrm{SECTION}}$ must be either an ELF section or raw data.
-      \emph{Function}: Replaces the section with the name SECTION$_{\mathrm{NAME}}$ in
+      \emph{Params}: ELF must be an ELF object. SECTION\_{}NAME must be a
+                string. NEW\_{}SECTION must be either an ELF section or raw data.
+      \emph{Function}: Replaces the section with the name SECTION\_{}NAME in
                   the oject ELF with the data from
-                  NEW$_{\mathrm{SECTION}}$. Section headers are replaced if NEW$_{\mathrm{SECTION}}$ is
+                  NEW\_{}SECTION. Section headers are replaced if NEW\_{}SECTION is
                   able to provide them, but not if it is only raw data.
@@
       \emph{Params}: ELF must be an ELF object. ADDRESS must be an
-                integer. NEW$_{\mathrm{DATA}}$ must be raw data.
+                integer. NEW\_{}DATA must be raw data.
       \emph{Function}: Replaces the raw data at OFFSET in the ELF object
-                  with NEW$_{\mathrm{DATA}}$. OFFSET must refer to a location in an
+                  with NEW\_{}DATA. OFFSET must refer to a location in an
                   existing section.
@@
       \emph{Params}: STR must be a string.
-      \emph{Function}: Prints the result of running elf$_{\mathrm{hash}}$ (from libelf)
+      \emph{Function}: Prints the result of running elf\_{}hash (from libelf)
                   on the string.
@@
 
+    Note that this feature is a work in progress. There isn't much you
+   can do with configuration files right now and the information here
+   may be out of date. Please do not rely on it.
+
     Katana loads configuration files as follows. Configuration files
@@
 
-    the katana manpage (although the information in this document is
+\begin{itemize}
+\item The katana manpage (although the information in this document is
       considerably more extensive than in the manpage)
+\item S. Bratus, J. Oakley, A. Ramaswamy, S. Smith,
+      M. Locasto. \emph{Katana: Towards Patching as a Runtime part of the       Compiler-Linker-Loader Toolchain}. International Journal of
+      Secure Software Engineering (IJSEE). 1, 3 (2010).
+\end{itemize}
 \subsection{Patch Object Format}
@@
 
-   This section of the document is not yet written. It will provide a description and specification of the PO format used by Katana
+   We have developed a patch object (PO) format which we hope will
+   eventually pave the way for a standardized vendor-neutral patch
+   format for hotpatching. We are not advancing our format as such,
+   but it embodies some of the principles which we think are
+   important. Why should patching not be a part of the ABI and of the
+   standard toolchain?
+\begin{itemize}
+\item A PO is a valid ELF file.
+\item A PO utilizes DWARF information to describe types, variables, and
+     functions requiring patching.
+\item A PO allows type transformations to be specified using a language
+     based on the DWARF standard.
+\end{itemize}
+   Through the use of existing standards and well-structured ELF files
+   utilizing a simple expression language for data patching, we aim to
+   create patches that are easily examined (or modified) with existing
+   tools. Relocatable objects containing new code and data which may
+   be inserted at runtime are nothing new. This is the entire premise
+   of the dynamic library. User-written functions which may have this
+   code injection (in the case of patching data where the desired
+   actions cannot be determined automatically) already exist as the
+   .init and .fini sections. It is our view, however, that it is
+   important to have a seperate patch format as opposed to patches
+   merely being dynamic libraries which contain both the patch data
+   and the logic to perform the patching (as is done by some other
+   hotpatching systems). We view this as an unnecessary mixing of data
+   and logic. The code to apply patches should live in one place on
+   any given system, as most other executable content does.
+
+   As an ELF object, our PO files contain the following non-standard
+   sections.
+
+\begin{itemize}
+\item .text.new
+     Contains new/modified functions
+\item .rodata.new, .data.new
+     new data
+\item .unsafe\_{}functions
+     Contains a simple listing (of symbol indices) of the functions in
+     the binary to be patched which should not have activation records
+     on the stack when patching is taking place.
+\item .debug\_{}info
+     Contains listings of the variables and functions which need to be
+     patched using the DWARF data format. This section is standard and
+     is used here with validly formatted data, but is used for
+     patching instead of debugging. The use of the the .debug\_{} name is
+     preserved for compatibility with libdwarf and tools such as
+     readelf, objdump, dwarfdump capable of listing DWARF
+     information. It can be, however, confusing and the name will
+     likely change in the future.
+\item .debug\_{}frame
+     Like .debug\_{}info a standard section used in a nonstandard way,
+     see notes above about the naming. Contains an extended version of
+     DWARF Call Frame Information which describes how various data
+     structures are to be patched. The details are not properly
+     documented at the moment, please email the Authors for more
+     details if you would like further information.
+\end{itemize}
 \subsection{Patch Generation Process}
@@
 
-   This section of the document is still under construction. When
-   complete, it will provide a description of the internal process that
-   Katana uses to generate a patch. Understanding it is not necessary
-   for using Katana.
+   This section of the document is still under construction, but we
+   hope that the information that is provided will be of some use.
 
@@
 \label{sec-3.8}
+
+   Note that this feature is a work in progress. There isn't much you
+   can do with configuration files right now.
 
@@
 
+   High level view:
 \begin{itemize}
@@
 \end{itemize}
+   A more detailed (although still very rough) algorithm:
+
+\begin{verbatim}
+Walk the old and new object trees in parallel
+  For each pair of objects (corresponding old and new objects)
+    If the new object does not exist
+      Issue a warning and continue
+    If the old object does not exist
+      Add all functions and vars to patch
+      Continue
+    If the two objects are the same
+      Continue
+    If the two objects differ
+      For every global variable in the old object
+        Compare with matching variable in the new object
+        If the two are a different type or the type struct changes
+          Generate a type transformation for the patch
+        If the variable initializers changed
+          If the variable is const
+            Add new data to the patch
+          Else
+            Generate a warning (can't determine automatically if
+                the change should be applied)
+        If anything related to the variable changed
+          Find all functions using the variable
+          Add them to the unsafe functions list
+      For every global variable only in the new object
+        Add it to the patch
+      For every function in the old object
+        Compare with matching function in the new object
+          If the functions differ
+            Add the new text to the patch
+            Add the function to the unsafe functions list.
+      For ever function only in the new object
+        Add the function to the patch
+Write out the patch ELF!
+\end{verbatim}
+
 \subsection{Type Diffing}
@@
 
-   This section of the document still needs to be written. The general
-   idea is that structures are examined for for added members, moved
-   members, and changed members.
+   The general idea is that structures are examined for for added
+   members, moved members, and changed members. If you need more
+   detail than this, please contact the Authors.
 \subsection{Function Diffing}
@@
 
+   Functions are compared in an unsophisticated manner. The comparison
+   is essentially byte-by-byte (i.e. no parsing of the machine
+   instruction set is done). If bytes differ between the compiled
+   version of the old function and the new function, then the function
+   is assumed to need patching.  The one exception to this is that
+   relocations are accounted for. If bytes differ at an address that
+   is fixed up by relocations, the relocations are examined to make
+   sure that they are for the same symbol. If in fact they are, then
+   the function is deemed not to have changed. If the symbol referred
+   to corresponds to a variable that has changed then it may need to
+   be moved to be patched. In that event the function may in fact have
+   to be modified, but it will be modified only to apply the
+   relocations rather than as a patch to the function per se and thus
+   the function diffing stage does not concern itself with whether
+   referenced symbols have changed.
 \subsection{Patch Application Process}
@@
    patch. Understanding it is not necessary for using Katana.
+
+   The basic process is as follows
+
+\begin{verbatim}
+Read the patch file
+Calculate versioning. This is currently not implemented.
+Find malloc in the target, as we may need it
+Calculate a safe state for the target (based on the unsafe functions list)
+Wait for the target to reach a safe state
+Map in necessary sections from the patch
+Copy PLT and GOT to new locations as we may need to expand them
+For each variable listed in the patch
+  Apply the variable patch
+For each function listed in the patch
+ Apply the function patch
+Apply necessary relocations
+\end{verbatim}
+
 \subsection{Roadmap}
@@
 
+  For information on how katana can be used specifically for DWARF
+  manipulation, please see Dartmouth College Tech Report TR2011-680.
 \section{Credits and Licensing}
@@
   Katana is being written by James Oakley and was designed by Sergey
-  Bratus, Ashwin Ramaswamy, James Oakley, Michael Locasto, and Sean
+  Bratus, James Oakley, Ashwin Ramaswamy, Michael Locasto, and Sean
   Smith.

### 2011-05-09 doc/katana.tex
updated other forms of the documentation besides just the .org file

@@
-% Created 2011-04-18 Mon 01:44
+% Created 2011-05-08 Sun 21:11
 \documentclass[11pt]{article}
@@
 \author{James Oakley}
-\date{18 April 2011}
+\date{08 May 2011}
 
@@
 
-    A COMMAND is of the form COMMAND\_{}IDENTIFIER PARAM PARAM PARAM \ldots{}., where
+    A COMMAND is of the form COMMAND$_{\mathrm{IDENTIFIER}}$ PARAM PARAM PARAM \ldots{}., where
     tokens are seperated by spaces and the number of PARAMs depends on
@@
 
-    A COMMAND\_{}IDENTIFIER is one or more words which identify a
+    A COMMAND$_{\mathrm{IDENTIFIER}}$ is one or more words which identify a
     COMMAND. In many cases a command is identified by only one word,
@@
                 Dwarfscript. If not specified it defaults to
-                ``.eh\_{}frame''. ELF must be an ELF object. OUTFILE must
+                ``.eh$_{\mathrm{frame}}$''. ELF must be an ELF object. OUTFILE must
                 be a string with the name of a file to write the resulting
@@
                   beinary form. Returns an array with 3 items
-                  0: raw data for .eh\_{}frame 
-                  1: raw data for .eh\_{}frame\_{}hdr
-                  2: raw data for .gcc\_{}except\_{}table.
+                  0: raw data for .eh$_{\mathrm{frame}}$ 
+                  1: raw data for .eh$_{\mathrm{frame}}$$_{\mathrm{hdr}}$
+                  2: raw data for .gcc$_{\mathrm{except}}$$_{\mathrm{table}}$.
 \end{itemize} % ends low level
 
-\item replace\\
+\item extract\\
 \label{sec-2.1.2.4}
@@
 
-\item replace section\\
+\item extract section\\
 \label{sec-2.1.2.4.1}
 
+      \emph{Usage}: \texttt{extract section ELF SECTION\_NAME}
+      \emph{Params}: ELF must be an ELF object. SECTION$_{\mathrm{NAME}}$ must be a
+                string. 
+      \emph{Function}: Returns the data and header information for the
+                  specified section
+
+\item extract section$_{\mathrm{data}}$\\
+\label{sec-2.1.2.4.2}
+
+      \emph{Usage}: \texttt{extract section ELF SECTION\_NAME}
+      \emph{Params}: ELF must be an ELF object. SECTION$_{\mathrm{NAME}}$ must be a
+                string. 
+      \emph{Function}: Like extract section except extracts only the raw
+                 data stored in the section and not any header information.
+\end{itemize} % ends low level
+
+\item replace\\
+\label{sec-2.1.2.5}
+
+\begin{itemize}
+
+\item replace section\\
+\label{sec-2.1.2.5.1}
+
       \emph{Usage}: \texttt{replace section ELF SECTION\_NAME NEW\_SECTION}
-      \emph{Params}: ELF must be an ELF object. SECTION\_{}NAME must be a
-                string. NEW\_{}SECTION must be either an ELF section or raw data.
-      \emph{Function}: Replaces the section with the name SECTION\_{}NAME in
-                  the oject ELF with the data from
-                  NEW\_{}SECTION. Section headers are replaced if NEW\_{}SECTION is
-                  able to provide them, but not if it is only raw data.
+      \emph{Params}: ELF must be an ELF object. SECTION$_{\mathrm{NAME}}$ must be a
+      string. NEW$_{\mathrm{SECTION}}$ must be either an ELF section or raw data.
+      \emph{Function}: Replaces the section with the name SECTION$_{\mathrm{NAME}}$ in
+      the oject ELF with the data from NEW$_{\mathrm{SECTION}}$. Section headers
+      are replaced if NEW$_{\mathrm{SECTION}}$ is able to provide them, but not if
+      it is only raw data.
 
 \item replace raw\\
-\label{sec-2.1.2.4.2}
+\label{sec-2.1.2.5.2}
 
@@
       \emph{Params}: ELF must be an ELF object. ADDRESS must be an
-                integer. NEW\_{}DATA must be raw data.
+                integer. NEW$_{\mathrm{DATA}}$ must be raw data.
       \emph{Function}: Replaces the raw data at OFFSET in the ELF object
-                  with NEW\_{}DATA. OFFSET must refer to a location in an
+                  with NEW$_{\mathrm{DATA}}$. OFFSET must refer to a location in an
                   existing section.
@@
 \item info\\
-\label{sec-2.1.2.5}
+\label{sec-2.1.2.6}
 
@@
 \item info eh\\
-\label{sec-2.1.2.5.1}
+\label{sec-2.1.2.6.1}
 
@@
 \item hash\\
-\label{sec-2.1.2.6}
+\label{sec-2.1.2.7}
 
@@
 \item hash elf\\
-\label{sec-2.1.2.6.1}
+\label{sec-2.1.2.7.1}
 
@@
       \emph{Params}: STR must be a string.
-      \emph{Function}: Prints the result of running elf\_{}hash (from libelf)
+      \emph{Function}: Prints the result of running elf$_{\mathrm{hash}}$ (from libelf)
                   on the string.
@@
 \item patch\\
-\label{sec-2.1.2.7}
+\label{sec-2.1.2.8}
 
@@
 \item gen\\
-\label{sec-2.1.2.7.1}
+\label{sec-2.1.2.8.1}
 
@@
 \item apply\\
-\label{sec-2.1.2.7.2}
+\label{sec-2.1.2.8.2}
 
@@
 \item ! (shell command)\\
-\label{sec-2.1.2.8}
+\label{sec-2.1.2.9}
 
@@
      new data
-\item .unsafe\_{}functions
+\item .unsafe$_{\mathrm{functions}}$
      Contains a simple listing (of symbol indices) of the functions in
@@
      on the stack when patching is taking place.
-\item .debug\_{}info
+\item .debug$_{\mathrm{info}}$
      Contains listings of the variables and functions which need to be
@@
      likely change in the future.
-\item .debug\_{}frame
-     Like .debug\_{}info a standard section used in a nonstandard way,
+\item .debug$_{\mathrm{frame}}$
+     Like .debug$_{\mathrm{info}}$ a standard section used in a nonstandard way,
      see notes above about the naming. Contains an extended version of
